---
title: "Culling and hazard ratios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Culling and hazard ratios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

Culling (and mortality) impacts are usually reported as hazard ratios (HRs)
from survival models. A raw HR for one disease is conflated with the HRs of
associated diseases, just like a yield impact, but HRs combine
multiplicatively, and the culling they cause has to be counted without
counting a cow twice. This vignette compares the ways the package can handle
them.

## Options

| Approach | What is adjusted | How diseases combine | How to use it |
|---|---|---|---|
| Multiplicative, exact | log HR, over the distribution of disease combinations | HRs multiply | `scale = "hazard_ratio"`, `method = "global"` |
| Multiplicative, first-order | log HR, pairwise | HRs multiply (approximately) | `scale = "hazard_ratio"`, `method = "simultaneous"` |
| HR - 1 (Rasmussen et al. 2024) | HR - 1, with eq. 16 | excess HRs add | `scale = "hazard_ratio"`, `method = "published"` |
| Excess risk (Rasmussen et al. 2022) | excess culling risk | excess risks add | `hr_conversion()`, `as_impacts()` |

Under the multiplicative model, a cow's hazard is
`h0 * exp(sum_i beta_i * D_i)`. The raw HR of disease `i` is the ratio of
the average hazard among cows with and without `i`. The global method
solves for the `beta`s exactly, so that the adjusted HRs reproduce the raw
HRs over the fitted distribution of disease combinations. The simultaneous
method uses the first-order version, which needs only the pairwise tables.

If a source HR comes from a model that already included the associated
diseases, set `adjusted_for` in `cm_impacts()` so that it is not adjusted
again (with the simultaneous method).

## The global dairy example

`example_global_dairy(culling_scale = "hazard_ratio")` enters the culling HRs
of Rasmussen et al. (2024) as hazard ratios:

```{r}
m <- example_global_dairy(culling_scale = "hazard_ratio")
cmp <- compare_methods(m)
cu <- cmp$impacts[cmp$impacts$outcome == "culling", ]
cu[, c("disease", "raw", "published", "simultaneous", "global")]
```

The published column reproduces the paper's adjustment at the central values
(Table 5 reports Monte Carlo means). The methods agree within about 0.1 for
most diseases, and differ most where comorbidity is heaviest (clinical
ketosis, lameness, displaced abomasum, metritis).

## Culling attributable to disease

`attributable_risk()` turns adjusted HRs into the part of the overall
culling risk that disease causes. It finds the baseline hazard for which the
population culling risk equals the observed rate, and compares it with the
risk of a disease-free cow. A cow's risk cannot exceed 1, so a cow with
several diseases is counted once. The attributable risk is then allocated to
diseases with Shapley values.

With the global average replacement rate of 23.66% (Rasmussen et al. 2024,
Table 1):

```{r}
glob <- cmp$results$global
ar <- attributable_risk(glob, overall_risk = 0.2366)
ar
```

The 2024 paper instead converted each adjusted HR to an excess risk relative
to the overall rate, `HR * r / (HR * r + 1 - r) - r`, and summed the excess
risks weighted by prevalence. That sum counts cows with several diseases
more than once:

```{r}
r <- 0.2366
P <- m$diseases$prob
hr_2024 <- cmp$impacts$published[cmp$impacts$outcome == "culling"]
sum(P * (hr_2024 * r / (hr_2024 * r + 1 - r) - r))   # paper's approach
ar$summary$attributable                               # model-based
```

The paper's approach attributes about 14.8 percentage points of the 23.7%
culling rate to disease; the multiplicative model attributes about 11.7.

## Valuing culling with the other outcomes

In `summary()`, `contribution_table()` and `compare_methods()`, a
hazard-ratio outcome is valued with `attributable_risk()`: its `observed`
value is the overall risk (a proportion) and its `unit_value` the value of a
cow removed.

```{r}
eco <- list(observed = c(yield = 5013, culling = 0.2366),
            unit_value = c(yield = 0.5981, culling = 1299.33 - 785.86))
summary(glob, economics = eco)$totals
```

(Global averages from Rasmussen et al. 2024, Table 1: milk yield, milk price
per kg, and replacement price less culled-cow price.)

## Caveats

* The models assume a constant hazard within the period and that the HRs
  apply for the whole period. Most source HRs come from models in which
  disease status changes during the lactation.
* Hazard ratios are not collapsible: even without confounding, an average
  HR differs from the HR within subgroups. The exact multiplicative model
  accounts for this given the joint distribution; the first-order method
  and the HR - 1 approach do not.
* Culling and death compete. Make sure the source HRs are of the same kind
  (cause-specific or subdistribution).
