---
title: "Culling and hazard ratios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Culling and hazard ratios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

Culling (and mortality) impacts are usually reported as hazard ratios (HRs)
from survival models. A raw HR for one disease is conflated with the HRs of
associated diseases, as a yield impact is, but HRs combine multiplicatively
and are not additive impacts. They are therefore adjusted by a separate
adapter, outside the additive engine of `deconflate()`:

* `cm_hazard_ratios()` describes the raw HRs (one per disease; use 1 for no
  effect) and states their estimand: `"snapshot_crude"` or
  `"snapshot_stratified"`;
* `cm_hr_model()` combines them with a population;
* `deconflate_hr()` adjusts them;
* `attributable_risk()` turns adjusted HRs into the part of the overall
  culling risk that disease causes, and allocates it to diseases.

This is a separate model with its own assumptions (the snapshot
hazard-multiplier model, below), not a general conversion of published
hazard ratios. Its estimands are defined by the model:

* `"snapshot_crude"`: the ratio of the average hazard among animals with
  the disease to that among animals without it, at the start of follow-up,
  in the population described by the probabilities and associations;
* `"snapshot_stratified"`: the same ratio within strata of the diseases in
  `adjusted_for`, combined across strata.

A published Cox HR is neither of these. Entering one as a snapshot ratio is
an approximation that you state, not a conversion the package makes: it is
reasonable when follow-up is short relative to the hazards, and it should be
reported and checked (see "What the snapshot model is not" below and in
`?deconflate_hr`). For this reason the estimand has no default. The names
`"crude"` and `"adjusted"` used before version 0.3.0 are not accepted:

```{r}
tryCatch(cm_hazard_ratios("d1", 1.5, estimand = "crude"),
         deconflate_unsupported = function(e) conditionMessage(e))
```

## Methods

```{r}
pop <- example_supplement()   # its yield impacts are not used here
hr <- cm_hr_model(pop, cm_hazard_ratios(c("d1", "d2", "d3"), c(1.5, 2.0, 1.3), estimand = "snapshot_crude"))
hr
compare_methods(hr)
```

`deconflate_hr()` has two methods; the snapshot model (the default) is the
one to use for hazard ratios:

* `"snapshot"` (the default) is a snapshot hazard-multiplier model. A cow's
  hazard is `h0 * exp(sum_i beta_i * D_i)`, and the raw HR of disease `i` is
  taken to be the ratio of the average hazard multiplier among cows with and
  without `i`, over the fitted distribution of disease combinations
  (`fit_joint()`) at the start of follow-up. The `beta`s are solved so that
  these ratios equal the raw HRs, and `exp(beta)` are the adjusted HRs.
* `"first_order"` is the log-linear approximation `log(HR_raw) = A beta`,
  with the conflation matrix `A` of `deconflate()`. It needs only the
  pairwise tables.

The `"published"` column of `compare_methods()` is the approach of Rasmussen
et al. (2024): HR - 1 adjusted with eq. 16 and 1 added back. It is not a
method of `deconflate_hr()`; it is kept for comparison and reproduction only.

```{r}
deconflate_hr(hr)
```

## Source hazard ratios adjusted for other diseases

If a source HR comes from a model that already included other diseases, the
closest snapshot estimand is `"snapshot_stratified"`, with those diseases in
`adjusted_for` (ids separated by `";"`), or `"all"` for every other
disease. In the snapshot model, the ratio for such an estimate is computed
within strata of its adjustment set and combined across strata with
Mantel-Haenszel-type weights. With `adjusted_for = "all"`, the HR is the
disease's own hazard multiplier and is used as it is. A Cox coefficient
adjusted for the same diseases matches this only under the snapshot model's
assumptions:

```{r}
hr_adj <- cm_hr_model(pop, cm_hazard_ratios(
  c("d1", "d2", "d3"), c(1.5, 2.0, 1.3),
  estimand = c("snapshot_stratified", "snapshot_crude", "snapshot_stratified"),
  adjusted_for = c("d2", NA, "all")
))
deconflate_hr(hr_adj)$adjusted
```

The published approach is defined for `"snapshot_crude"` HRs only:

```{r}
compare_methods(hr_adj)$failed
```

## What the snapshot model is not

A Cox HR estimated over follow-up is not, in general, the snapshot ratio:
cows with high hazards leave first, so the mixture of disease combinations
among survivors changes over time, and the marginal HR changes with it. The
snapshot model is exact for its own estimand (instantaneous marginal ratios
at baseline). For published Cox coefficients it is an approximation, which
is reasonable when follow-up is short relative to the hazards or the
diseases are rare.

## Culling attributable to disease

`attributable_risk()` finds the baseline hazard for which the population
culling risk, averaged over the distribution of disease combinations, equals
the observed rate, and compares it with the risk of a disease-free cow. A
cow's risk cannot exceed 1, so a cow with several diseases is counted once.
The attributable risk is allocated to diseases by Shapley values over all
disease combinations (`shapley_by_cell()`); the allocation adds up to the
attributable risk.

```{r}
attributable_risk(deconflate_hr(hr), overall_risk = 0.25)
```

The risks are proportions of animals culled in the period. Converting them
into numbers of animals or a value (e.g. times the replacement price less
salvage value) is left to the user.

## The global dairy example

`example_global_dairy_hr()` contains the culling HRs of Rasmussen et al.
(2024), on the population of `example_global_dairy()`. They are published
Cox HRs, entered as `"snapshot_crude"`; that is an assumption, and the
results below hold under it. `compare_methods()` with `overall_risk` adds
the attributable risk per method (without the allocation). The global average replacement rate is 23.66% (Rasmussen et al.
2024, Table 1). The joint distribution of the 12 diseases is fitted once and
passed on with `joint`, so that it is not refitted for each method:

```{r}
hr_gd <- example_global_dairy_hr()
j_gd <- fit_joint(hr_gd$population)
compare_methods(hr_gd, overall_risk = 0.2366, joint = j_gd)
```

The published column reproduces the paper's adjustment at the central values
(Table 5 reports Monte Carlo means). The methods agree within about 0.1 for
most diseases, and differ most where comorbidity is heaviest (clinical
ketosis, lameness, displaced abomasum, metritis). For clinical ketosis and
metritis, the snapshot and first-order models give adjusted HRs below 1:
under the multiplicative model, the raw HRs are smaller than the associated
diseases alone would produce. `deconflate_hr()` warns about this:

```{r}
glob <- deconflate_hr(hr_gd, joint = j_gd)
ar <- attributable_risk(glob, overall_risk = 0.2366)
ar
```

The model attributes about 11.7 percentage points of the 23.7% culling rate
to disease. The 2024 paper instead converted each adjusted HR to an excess
risk relative to the overall rate, `HR * r / (HR * r + 1 - r) - r`, and
summed these excess risks weighted by prevalence. With the published
adjusted HRs, that sum is about 14.8 percentage points, because it counts
cows with several diseases more than once.

## Culling alongside the additive analyses

Hazard ratios are kept out of `deconflate()` and the contribution tables:
the additive analyses return impacts in their own units, and
`attributable_risk()` returns a proportion of cows. To set them side by side,
convert both to a common unit, here a value per cow and year with the
global average yield of 5013 kg, a milk price of 0.5981 per kg, and a
replacement price less culled-cow price of 1299.33 - 785.86 (Rasmussen et al.
2024, Table 1). The yield impacts are percent losses, so the gap is computed
as in `vignette("deconflate")`:

```{r}
yres <- deconflate(example_global_dairy()$models$yield)
yield_gap <- 5013 / (1 - yres$totals$adjusted_total / 100) - 5013
c(yield = yield_gap * 0.5981,
  culling = ar$summary$attributable * (1299.33 - 785.86))
```

The yield analysis here uses the simultaneous method; the paper's figures
(published approximation, Monte Carlo means) are recomputed by
`reproduce_rasmussen_2024()` (`vignette("reproducing-published")`).

## Methods that are not available for new analyses

The published analyses converted hazard ratios in ways that the package
does not offer for new analyses:

* Rasmussen et al. (2022) treated each HR as an odds ratio of a 2x2 table of
  disease by culling, to obtain an excess annual culling risk, and recovered
  adjusted HRs by rescaling (eq. 23);
* Rasmussen et al. (2024) adjusted HR - 1 as if it were an additive impact,
  and valued culling with the excess risk relative to the overall rate given
  above.

These conversions are used only inside `reproduce_rasmussen_2022()` and
`reproduce_rasmussen_2024()`, so that the published tables can be recomputed
(see `vignette("reproducing-published")`).

## Caveats

* The models assume a constant hazard within the period and that the HRs
  apply for the whole period. Most source HRs come from models in which
  disease status changes during the lactation.
* Hazard ratios are not collapsible: even without confounding, an average HR
  differs from the HR within subgroups. The snapshot model accounts for this
  given the distribution of disease combinations; the first-order method and
  the published approach do not.
* The snapshot model and `attributable_risk()` use the whole distribution of
  disease combinations, so they depend on the maximum-entropy assumption and
  on any three-way terms (`vignette("interactions")`).
* Culling and death compete. Make sure the source HRs are of the same kind
  (cause-specific or subdistribution).
