---
title: "Getting started with deconflate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with deconflate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

## The problem

Single-disease impact estimates compare animals with and without a disease.
Animals with one disease are more (or less) likely to have associated
diseases, so each estimate partly reflects the impacts of those other
diseases. Adding such estimates double counts. `deconflate` adjusts
("de-conflates") the estimates so that they can be aggregated.

The package assumes no species, outcome or unit. The examples use dairy
cattle and milk yield, but any population of animals (or people) and any
additive outcome works the same way.

## A population and an impact table

The population holds what every outcome shares: the disease probabilities
and the pairwise associations. This is the three-disease example from the
Supplementary File of Rasmussen et al. (2022):

```{r}
pop <- cm_population(
  cm_diseases(c("d1", "d2", "d3"), c(0.10, 0.15, 0.20), type = "prevalence"),
  cm_associations(c("d1", "d1", "d2"), c("d2", "d3", "d3"), c(2, 1, 3), measure = "OR")
)
pop
```

Associations are odds ratios by default (risk ratios, risk differences,
conditional probabilities and phi coefficients are also accepted; see
`?cm_associations`). An odds ratio of 1, as for d1 and d3, states that two
diseases are unrelated. A pair with no association at all is *unknown*:
`deconflate()` then fills it in from the other associations (below).

A model adds one impact table: one raw estimate per disease, all in the
same units. The units can be anything additive (percent of yield, kg of
milk, days, euros, a welfare score); the engine does not convert them, and
the results come back in the units supplied.

```{r}
yield <- cm_impacts(c("d1", "d2", "d3"), c(2.5, 5, 7.5),
                    label = "milk yield loss", units = "% of yield")
m <- cm_model(pop, yield)
m
```

The same impacts in kg give the same adjustment, in kg:

```{r}
kg <- cm_model(pop, cm_impacts(c("d1", "d2", "d3"), c(250, 500, 750),
                               units = "kg/animal/year"))
deconflate(kg)$adjusted
```

Each `deconflate()` call adjusts one impact table. For another outcome, make
another model on the same population and call `deconflate()` again:

```{r}
ci <- cm_model(pop, cm_impacts(c("d1", "d2", "d3"), c(4, 10, 2),
                               label = "calving interval increase", units = "days"))
deconflate(ci)$adjusted
```

The tables can also be read from CSV files (with any names) or data frames
with `cm_read_inputs()`, which checks them and reports every problem at
once (see `vignette("own-data")`):

```{r}
ex <- system.file("extdata", "five_diseases", package = "deconflate")
five <- cm_read_inputs(diseases = file.path(ex, "diseases.csv"),
                       associations = file.path(ex, "associations.csv"),
                       impacts = file.path(ex, "impacts_yield.csv"))
five
```

## Adjusting impacts

```{r}
res <- deconflate(m)
res
```

`deconflate()` has two exact methods, chosen automatically
(`method = "auto"`, the default):

* the **simultaneous** method solves the system of impact equations from
  the pairwise 2x2 tables. It works for any number of diseases;
* the **global** method fits the maximum-entropy distribution of disease
  combinations and solves the same equations from it. It also handles
  unknown pairs, impact interactions and three-way terms
  (`vignette("interactions")`). Up to 20 diseases it enumerates all
  combinations; beyond that it uses the sampled backend of `fit_joint()`
  (`vignette("thresholds-and-scaling")`).

When every pair has an association and there are no interactions or
three-way terms, the two give the same answer and the simultaneous method
is used. Otherwise the global method is used, and a note says why. Leaving
out the d1:d3 odds ratio makes that pair unknown:

```{r}
pop_unknown <- cm_population(pop$diseases,
                             cm_associations(c("d1", "d2"), c("d2", "d3"), c(2, 3)))
res_unknown <- deconflate(cm_model(pop_unknown, yield))
res_unknown$notes
res_unknown$unknown_pairs
rbind(odds_ratio_1 = setNames(res$adjusted$adjusted, res$adjusted$disease),
      unknown = res_unknown$adjusted$adjusted)
```

The global fit gives d1 and d3 a positive association (an odds ratio of
about 1.14), implied by their associations with d2, so the adjusted impacts
differ from those with an odds ratio of 1. Unknown is not the same as independent; see
`vignette("interactions")`.

`compare_methods()` sets the methods side by side, without switching:

```{r}
compare_methods(m)
```

The `"published"` column is the proportional approximation of Rasmussen et
al. (2022, 2024), eq. 16. It is not a method of `deconflate()`: it is kept
for comparison (`compare_methods()`) and for reproducing the published
analyses (`vignette("reproducing-published")`) only. It is defined for crude
estimates only.

The diagnostics report how well the adjusted impacts reconstruct the raw
ones, the condition number of the conflation matrix, and adjusted impacts
whose sign differs from the raw impact. A sign change means that the raw
impact is smaller than the associated diseases alone would produce, so the
inputs are inconsistent under the additive model (or the estimands are not
what they were assumed to be):

```{r}
low_d2 <- cm_model(pop, cm_impacts(c("d1", "d2", "d3"), c(2.5, 0.2, 7.5)))
tryCatch(deconflate(low_d2), deconflate_sign_change = function(w) conditionMessage(w))
```

## Contributions

The expected aggregate impact per animal is `sum_i p_i b_i` (plus any
interaction terms). Each disease's contribution is its own term plus half of
each interaction term it is involved in, which is its Shapley value in
closed form. Contributions add up to the aggregate:

```{r}
res$contributions
res$totals
```

`raw_sum` is the naive aggregate of the unadjusted impacts. The shares are
`NA` when the aggregate is zero, since a share of zero is not defined.
`contribution_table()` and `summary()` combine the impacts and
contributions in one table.

How robust is a ranking or a sign to one uncertain input? `cm_threshold()`
finds the input values at which the conclusion changes, for example the odds
ratio of d2 and d3 at which d1 would contribute more than d2:

```{r}
cm_threshold(m, "assoc:d2:d3", c(0.1, 100), conclusion = "rank")$thresholds[
  , c("item", "status", "threshold", "description")]
```

See `vignette("thresholds-and-scaling")`.

## Estimands

Each raw impact must be one of two estimands, set explicitly in
`cm_impacts()`:

* `"crude"` (default): the difference in the outcome between animals with
  and without the disease, not adjusted for other diseases.
* `"adjusted_linear"`: the coefficient of the disease in an additive
  (linear) regression of the outcome on the disease and the diseases listed
  in `adjusted_for` (ids separated by `";"`, or `"all"` for every other
  disease).

`adjusted_for` is never used to infer the estimand. Giving it for a crude
estimate is an error:

```{r}
tryCatch(cm_impacts("d1", 2, adjusted_for = "d2"),
         deconflate_unsupported = function(e) conditionMessage(e))
```

Other adjusted estimands (matched or propensity-score estimates, for
example) are not supported. For an adjusted estimate, the package uses the
population linear projection of the omitted diseases, which depends only on
the pairwise tables. This requires that the probabilities and associations
describe the source population of the estimates.

A worked example: in the population above, the true additive impacts are
0.02, 0.04 and 0.06. A study of d1 that adjusted for d2, and crude studies of
d2 and d3, would report:

```{r}
truth <- c(d1 = 0.02, d2 = 0.04, d3 = 0.06)
raw <- simulate_raw_impacts(pop, truth,
                            estimand = c("adjusted_linear", "crude", "crude"),
                            adjusted_for = c("d2", NA, NA))
round(raw$value, 4)
```

With the estimands recorded, the adjustment recovers the truth:

```{r}
imp_adj <- cm_impacts(c("d1", "d2", "d3"), raw$value,
                      estimand = c("adjusted_linear", "crude", "crude"),
                      adjusted_for = c("d2", NA, NA))
m_adj <- cm_model(pop, imp_adj)
deconflate(m_adj)$adjusted
```

Treating the d1 estimate as crude adjusts it a second time for d2 and gives
the wrong answer:

```{r}
m_wrong <- cm_model(pop, cm_impacts(c("d1", "d2", "d3"), imp_adj$value))
deconflate(m_wrong)$adjusted$adjusted
```

The published approximation is not defined for adjusted estimands;
`compare_methods()` records the reason:

```{r}
compare_methods(m_adj)$failed
```

## Uncertainty

When inputs have distributions (the `dist` columns of the input tables, or
`distributions` in `cm_model()`), `deconflate()` also draws them
(`n_draws = 1000` by default) and adds 95% intervals. The central estimate
always uses the point values. Without distributions no draws are run, and a
note says so (as in the results above). The five-disease example has
distributions for most inputs; 200 draws keep this example fast:

```{r}
res_five <- deconflate(five, n_draws = 200, seed = 1)
res_five
```

Pass `n_draws = 0` for point estimates only. See `vignette("uncertainty")`
for the draws, their stability checks and the sensitivity tools.

## Event impacts

Impacts on the risk of an event, such as culling or death, are hazard
ratios, rate ratios, risk ratios, odds ratios or risk differences rather than
additive amounts. They are given with a `measure` (one per disease; measures
can be mixed) and a snapshot estimand, and adjusted with
`event_model = TRUE` and the overall risk of the event over the period:

```{r}
cull <- cm_model(pop, cm_impacts(c("d1", "d2", "d3"), c(1.5, 2.0, 1.2),
                                 measure = c("HR", "HR", "RR"),
                                 estimand = "snapshot_crude", label = "culling"))
res_cull <- deconflate(cull, event_model = TRUE, overall_risk = 0.25)
res_cull$adjusted
res_cull$attributable$summary
```

The adjusted values are each disease's own hazard ratio, and
`$attributable` gives the risk attributable to disease and its allocation to
diseases. See `vignette("event-impacts")`.

## From adjusted impacts to gaps and values

`deconflate()` returns impacts and contributions in the units of the
impacts; it does not convert them into other quantities. Turning them into a
gap between the observed and the disease-free level of an outcome, or into a
value, needs to know what the impacts mean, and takes a few lines of base R.

When the impacts are percentage losses relative to the disease-free level
(as in `m`, with an observed mean of 10,000 units), the disease-free level is
`observed / (1 - aggregate / 100)`, the gap is the disease-free level minus
the observed level, and each disease's part of the gap is its share of the
aggregate:

```{r}
observed <- 10000
aggregate <- res$totals$adjusted_total            # in percent
disease_free <- observed / (1 - aggregate / 100)
gap <- disease_free - observed
ct <- res$contributions
by_disease <- data.frame(disease = ct$disease, gap = gap * ct$share)
by_disease
c(disease_free = disease_free, gap = gap)
```

A value per unit of outcome turns the gaps into values; any costs that are
not part of the adjustment are added separately:

```{r}
unit_value <- 0.30
by_disease$value <- by_disease$gap * unit_value
sum(by_disease$value)
```

`ct$share` is `NA` when the aggregate is zero; `disease_free * ct$total / 100`
gives the same gaps and stays defined. For an outcome that disease
increases, the disease-free level is `observed / (1 + aggregate / 100)` and the
gap is `observed - disease_free`. When the impacts are absolute effects (in
the outcome's own units, such as the days of the calving-interval model
`ci` above), no conversion is needed: each disease's contribution
(`ct$total`) is its part of the gap, and the aggregate is the gap. With
draws, the gap increases with the aggregate, so the same formula applied to
the interval bounds of the aggregate (`adjusted_total_lower` and
`adjusted_total_upper` in `$totals`) gives an interval for the total gap.

## Feasibility

Each pairwise table can be valid on its own while no population has all of
them. By default the simultaneous method screens every triple of diseases (a
necessary condition), and `feasibility = "lp"` runs an exact check (it
needs the `lpSolve` package):

```{r}
res$diagnostics$feasibility
```

Three diseases with probability 0.5, two strong positive associations and
one strong negative one cannot coexist:

```{r}
bad <- cm_model(
  cm_diseases(c("a", "b", "c"), c(0.5, 0.5, 0.5)),
  cm_impacts(c("a", "b", "c"), c(1, 2, 3)),
  associations = cm_associations(c("a", "a", "b"), c("b", "c", "c"), c(20, 20, 0.05))
)
check_feasibility(bad, method = "triples")
```

## Conditions

Errors and warnings have classes, so that callers (and the draws and
sensitivity tools of the package) can handle them separately:

| Class | Meaning |
|---|---|
| `deconflate_infeasible` | inputs that admit no valid probabilities (e.g. the example above) |
| `deconflate_singular` | a singular or non-identifiable system |
| `deconflate_nonconvergence` | a numerical procedure did not converge |
| `deconflate_unsupported` | an estimand and method combination that is not supported |
| `deconflate_nonfinite` | non-finite results |
| `deconflate_input_problems` | `cm_read_inputs()` found errors in the input tables |

All errors also have class `deconflate_error`, and warnings
`deconflate_warning` (e.g. `deconflate_sign_change`).

```{r}
e <- tryCatch(deconflate(bad), deconflate_error = function(e) e)
class(e)
conditionMessage(e)
```

`deconflate()` also needs at least one association; without any, there is
nothing to de-conflate (the diseases would be treated as independent), and
it points to the tools that show how much associations could matter:

```{r}
no_assoc <- cm_model(pop$diseases, yield)
tryCatch(deconflate(no_assoc), deconflate_error = function(e) conditionMessage(e))
```

## Checking the method with a simulated herd

`simulate_raw_impacts()` computes the raw impacts that single-disease studies
would report in a herd with known true impacts. Adjusting them should
recover the truth; the published approximation (shown for comparison) does
so only approximately:

```{r}
m_sim <- cm_model(pop, simulate_raw_impacts(pop, c(d1 = 2, d2 = 4, d3 = 6)))
compare_methods(m_sim, methods = c("published", "simultaneous"))
```

## Where next

* `vignette("own-data")`: reading your own inputs from CSV files or data
  frames.
* `vignette("event-impacts")`: culling and other event impacts, and the risk
  attributable to disease.
* `vignette("reproducing-published")`: the 2022 and 2024 analyses.
* `vignette("interactions")`: the joint distribution, unknown associations,
  impact interactions, three-way scenarios and attribution.
* `vignette("uncertainty")`: draws and intervals, stability checks and
  sensitivity screening.
* `vignette("thresholds-and-scaling")`: threshold searches with
  `cm_threshold()` (at what input value does a ranking, a sign or the
  aggregate change?), and the sampled backend of `fit_joint()` for the
  global method with more than about 20 diseases.
