---
title: "Getting started with deconflate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with deconflate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

## The problem

Single-disease impact estimates compare animals with and without a disease.
Animals with one disease are more (or less) likely to have associated
diseases, so each estimate partly reflects the impacts of those other
diseases. Adding such estimates double counts. `deconflate` adjusts
("de-conflates") the estimates so that they can be aggregated.

## A population and an impact vector

The population holds what all analyses share: the disease probabilities and
the pairwise associations. This is the three-disease example from the
Supplementary File of Rasmussen et al. (2022):

```{r}
pop <- cm_population(
  cm_diseases(c("d1", "d2", "d3"), c(0.10, 0.15, 0.20), type = "prevalence"),
  cm_associations(c("d1", "d2"), c("d2", "d3"), c(2, 3), measure = "OR")
)
pop
```

Pairs without an association are independent by default
(`missing_associations = "independent"`). Use `"unknown"` to leave them
unconstrained instead; this is a different assumption (see
`vignette("interactions")`).

An analysis adds one impact vector: one value per disease, all in the same
units. The units can be anything additive (percent of yield, kg of milk,
days, euros, a welfare score); the engine does not convert them, and the
results come back in the units supplied.

```{r}
yield <- cm_impacts(c("d1", "d2", "d3"), c(2.5, 5, 7.5),
                    label = "milk yield loss", units = "% of yield")
m <- cm_model(pop, yield)
m
```

The same impacts in kg give the same adjustment, in kg:

```{r}
kg <- cm_model(pop, cm_impacts(c("d1", "d2", "d3"), c(250, 500, 750),
                               units = "kg/cow/year"))
deconflate(kg)$adjusted
```

Several types of impact are separate analyses on one population:

```{r}
a <- cm_analyses(pop,
  yield = yield,
  calving_interval = cm_impacts(c("d1", "d2", "d3"), c(4, 10, 2), units = "days")
)
a
```

`deconflate(a)` adjusts each analysis and returns a named list of results.

## Adjusting impacts

```{r}
compare_methods(m)
```

* `"simultaneous"` (the default) solves the system of impact equations
  exactly, using the pairwise 2x2 tables.
* `"published"` is the proportional approximation of Rasmussen et al.
  (2022, 2024), eq. 16, kept for reproduction and comparison. It is defined
  for crude estimates only.
* `"global"` fits the maximum-entropy distribution of disease combinations
  and can include impact interactions (`vignette("interactions")`). Without
  interactions and unknown pairs, it equals `"simultaneous"`. The pairwise
  methods work for any number of diseases; the global method enumerates
  combinations up to about 20 diseases, and beyond that uses the sampled
  backend of `fit_joint()` (`vignette("thresholds-and-scaling")`).

```{r}
res <- deconflate(m)
res
```

The diagnostics report how well the adjusted impacts reconstruct the raw
ones, the condition number of the conflation matrix, and adjusted impacts
whose sign differs from the raw impact. A sign change means that the raw
impact is smaller than the associated diseases alone would produce, so the
inputs are inconsistent under the additive model (or the estimands are not
what they were assumed to be):

```{r}
low_d2 <- cm_model(pop, cm_impacts(c("d1", "d2", "d3"), c(2.5, 0.2, 7.5)))
tryCatch(deconflate(low_d2), deconflate_sign_change = function(w) conditionMessage(w))
```

## Contributions

The expected aggregate impact per animal is `sum_i p_i b_i` (plus any
interaction terms). Each disease's contribution is its own term plus half of
each interaction term it is involved in, which is its Shapley value in
closed form. Contributions add up to the aggregate:

```{r}
res$contributions
res$totals
```

`raw_sum` is the naive aggregate of the unadjusted impacts. The shares are
`NA` when the aggregate is zero, since a share of zero is not defined.
`contribution_table()` and `summary()` combine the impacts and
contributions in one table.

How robust is a ranking or a sign to one uncertain input? `cm_threshold()`
finds the input values at which the conclusion changes, for example the odds
ratio of d2 and d3 at which d1 would contribute more than d2:

```{r}
cm_threshold(m, "assoc:d2:d3", c(0.1, 100), conclusion = "rank")$thresholds[
  , c("item", "status", "threshold", "description")]
```

See `vignette("thresholds-and-scaling")`.

## Estimands

Each raw impact must be one of two estimands, set explicitly in
`cm_impacts()`:

* `"crude"` (default): the difference in the outcome between animals with
  and without the disease, not adjusted for other diseases.
* `"adjusted_linear"`: the coefficient of the disease in an additive
  (linear) regression of the outcome on the disease and the diseases listed
  in `adjusted_for` (ids separated by `";"`, or `"all"` for every other
  disease).

`adjusted_for` is never used to infer the estimand. Giving it for a crude
estimate is an error:

```{r}
tryCatch(cm_impacts("d1", 2, adjusted_for = "d2"),
         deconflate_unsupported = function(e) conditionMessage(e))
```

Other adjusted estimands (matched or propensity-score estimates, for
example) are not supported. For an adjusted estimate, the package uses the
population linear projection of the omitted diseases, which depends only on
the pairwise tables. This requires that the probabilities and associations
describe the source population of the estimates.

A worked example: in the population above, the true additive impacts are
0.02, 0.04 and 0.06. A study of d1 that adjusted for d2, and crude studies of
d2 and d3, would report:

```{r}
truth <- c(d1 = 0.02, d2 = 0.04, d3 = 0.06)
raw <- simulate_raw_impacts(pop, truth,
                            estimand = c("adjusted_linear", "crude", "crude"),
                            adjusted_for = c("d2", NA, NA))
raw$value
```

that is, 0.018664808346, 0.054064376914 and 0.066681750858. With the
estimands recorded, the simultaneous method recovers the truth:

```{r}
imp_adj <- cm_impacts(c("d1", "d2", "d3"),
                      c(0.018664808346, 0.054064376914, 0.066681750858),
                      estimand = c("adjusted_linear", "crude", "crude"),
                      adjusted_for = c("d2", NA, NA))
m_adj <- cm_model(pop, imp_adj)
deconflate(m_adj)$adjusted
```

Treating the d1 estimate as crude adjusts it a second time for d2 and gives
the wrong answer:

```{r}
m_wrong <- cm_model(pop, cm_impacts(c("d1", "d2", "d3"), imp_adj$value))
deconflate(m_wrong)$adjusted$adjusted
```

The published approximation is not defined for adjusted estimands;
`compare_methods()` records the reason:

```{r}
compare_methods(m_adj)$failed
```

## Productivity gaps and values (optional)

The adjustment works on impacts alone. Turning the aggregate into a
productivity gap needs to know what the impacts mean, so it is a separate,
optional step. Here the impacts are percent decreases in a yield of 10,000
units:

```{r}
gaps <- productivity_gap(res, observed = 10000, direction = "decrease",
                         effect = "percent")
gaps$summary
gaps$attribution
```

`value_losses()` values the gap. Lump-sum costs that are not part of any
adjustment (e.g. veterinary expenditure) can be added with `additional`;
they are reported separately and included in the total:

```{r}
v <- value_losses(gaps, unit_value = 0.30, additional = c(veterinary = 25))
v$by_disease
v$total
```

The same valuation can be passed to `contribution_table()`, `summary()` and
`compare_methods()`:

```{r}
val <- list(observed = 10000, direction = "decrease", effect = "percent",
            unit_value = 0.30)
summary(res, valuation = val)
```

## Feasibility

Each pairwise table can be valid on its own while no population has all of
them. By default the pairwise methods screen every triple of diseases (a
necessary condition), and `feasibility = "lp"` runs an exact check (it
needs the `lpSolve` package):

```{r}
res$diagnostics$feasibility
```

Three diseases with probability 0.5, two strong positive associations and
one strong negative one cannot coexist:

```{r}
bad <- cm_model(
  cm_diseases(c("a", "b", "c"), c(0.5, 0.5, 0.5)),
  cm_impacts(c("a", "b", "c"), c(1, 2, 3)),
  associations = cm_associations(c("a", "a", "b"), c("b", "c", "c"), c(20, 20, 0.05))
)
check_feasibility(bad, method = "triples")
```

## Conditions

Errors and warnings have classes, so that callers (and the Monte Carlo and
screening functions) can handle them separately:

| Class | Meaning |
|---|---|
| `deconflate_infeasible` | inputs that admit no valid probabilities (e.g. the example above) |
| `deconflate_singular` | a singular or non-identifiable system |
| `deconflate_nonconvergence` | a numerical procedure did not converge |
| `deconflate_unsupported` | an estimand and method combination that is not supported |
| `deconflate_nonfinite` | non-finite results |
| `deconflate_input_problems` | `cm_read_inputs()` found errors in the input tables |

All errors also have class `deconflate_error`, and warnings
`deconflate_warning` (e.g. `deconflate_sign_change`).

```{r}
e <- tryCatch(deconflate(bad), deconflate_error = function(e) e)
class(e)
```

## Checking the method with a simulated herd

`simulate_raw_impacts()` computes the raw impacts that single-disease studies
would report in a herd with known true impacts. Adjusting them should
recover the truth; the published approximation does so only approximately:

```{r}
m_sim <- cm_model(pop, simulate_raw_impacts(pop, c(d1 = 2, d2 = 4, d3 = 6)))
compare_methods(m_sim, methods = c("published", "simultaneous"))
```

## Where next

* `vignette("own-data")`: reading your own inputs from CSV files.
* `vignette("culling-hazard-ratios")`: culling hazard ratios and the culling
  attributable to disease.
* `vignette("reproducing-published")`: the 2022 and 2024 analyses.
* `vignette("interactions")`: the joint distribution, unknown associations,
  impact interactions, three-way scenarios and attribution.
* `vignette("uncertainty")`: Monte Carlo analysis, stability checks,
  scenarios and sensitivity screening.
* `vignette("thresholds-and-scaling")`: threshold searches with
  `cm_threshold()` (at what input value does a ranking, a sign or the
  aggregate change?), and the sampled backend of `fit_joint()` for the
  global method with more than about 20 diseases.
