---
title: "Getting started with deconflate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with deconflate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

## The problem

Single-disease impact estimates compare animals with and without a disease.
Animals with one disease are more (or less) likely to have associated
diseases, so each estimate partly reflects the impacts of those other
diseases. Adding such estimates double counts. `deconflate` adjusts
("de-conflates") the estimates so that they can be aggregated.

## Describing a system

A model combines disease probabilities, pairwise associations and raw
impacts. This is the three-disease example from the Supplementary File of
Rasmussen et al. (2022):

```{r}
m <- cm_model(
  diseases = cm_diseases(c("d1", "d2", "d3"), c(0.10, 0.15, 0.20)),
  associations = cm_associations(c("d1", "d2"), c("d2", "d3"), c(2, 3), measure = "OR"),
  impacts = cm_impacts(c("d1", "d2", "d3"), c(2.5, 5, 7.5),
                       outcome = "yield", scale = "percent")
)
m
```

Pairs without an association are independent by default
(`missing_associations = "independent"`). Use `"unknown"` to leave them
unconstrained instead; this is a different assumption (see
`vignette("interactions")`).

## Adjusting impacts

```{r}
compare_methods(m)
```

* `"simultaneous"` (the default) solves the additive impact equations
  exactly.
* `"published"` is the proportional approximation of Rasmussen et al.
  (2022), kept for reproduction and comparison.
* `"global"` fits the distribution of disease combinations and can include
  impact interactions.

```{r}
res <- deconflate(m)
res
```

The diagnostics report how well the adjusted impacts reconstruct the raw
ones, and flag adjusted impacts whose sign differs from the raw impact. A
sign change means that the raw impact is smaller than the associated
diseases alone would produce, so the inputs are inconsistent under the
additive model.

## Productivity gaps and losses

```{r}
gaps <- productivity_gap(res, c(yield = 10000))
gaps$summary
value_losses(gaps, unit_value = c(yield = 0.30))$by_disease
```

## Checking the method with a simulated herd

`simulate_raw_impacts()` computes the raw impacts that single-disease studies
would report in a herd with known true impacts. Adjusting them should
recover the truth:

```{r}
truth <- c(d1 = 0.02, d2 = 0.04, d3 = 0.06)
m_sim <- m
m_sim$impacts <- simulate_raw_impacts(m, truth, outcome = "yield")
compare_methods(m_sim, methods = c("published", "simultaneous"))
```

## Where next

* `vignette("reproducing-published")`: the 2022 and 2024 analyses.
* `vignette("interactions")`: the joint distribution, unknown associations,
  impact interactions and attribution.
* `vignette("uncertainty")`: Monte Carlo analysis, correlated outcomes,
  scenarios and sensitivity screening.
