---
title: "Event impacts (culling and mortality)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Event impacts (culling and mortality)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

Some impacts are not amounts but comparisons of the risk of an event, such
as culling or death, between animals with and without a disease. They are
reported as hazard ratios (HR) from survival models, rate ratios, risk
ratios (RR), odds ratios (OR) or risk differences (RD). Such an estimate for
one disease is conflated with the effects of associated diseases, as an
additive impact is, but the effects combine multiplicatively, not
additively. These are **event impacts**, and `deconflate()` adjusts them
with a separate model, the snapshot hazard model, when called with
`event_model = TRUE`.

All five measures are mapped onto the same model, so they can be mixed in
one impact table (one row per disease): an HR for one disease, a risk
difference for another. The result is the same for every measure: an
adjusted hazard ratio per disease (its own hazard multiplier), and the risk
of the event attributable to disease, allocated to diseases.

## Describing event impacts

An impact table holds event impacts when it has a `measure` column: `"HR"`,
`"rate_ratio"`, `"RR"`, `"OR"` or `"RD"`. Every disease needs a row (1 for
no effect on a ratio scale, 0 for a risk difference). The estimand must be
stated; there is no default, because it is an assumption about what the
source estimate measures (see below):

```{r}
pop <- example_supplement()   # its yield impacts are not used here
cull <- cm_model(pop, cm_impacts(c("d1", "d2", "d3"), c(1.5, 2.0, 1.3), measure = "HR",
                                 estimand = "snapshot_crude", label = "culling"))
cull
tryCatch(cm_impacts("d1", 1.5, measure = "HR"),
         deconflate_error = function(e) conditionMessage(e))
```

In a CSV file read by `cm_read_inputs(impacts = )`, the same information is
in the columns `disease`, `value`, `measure`, `estimand` and `adjusted_for`
(see the five-disease example at the end).

`deconflate()` needs `event_model = TRUE` and the overall risk of the event:

```{r}
res <- deconflate(cull, event_model = TRUE, overall_risk = 0.25, n_draws = 0)
res
```

The kind of impacts and the call must agree. Event impacts without
`event_model = TRUE`, additive impacts with it, or a missing
`overall_risk` are errors:

```{r}
tryCatch(deconflate(cull), deconflate_error = function(e) conditionMessage(e))
tryCatch(deconflate(pop, event_model = TRUE, overall_risk = 0.25),
         deconflate_error = function(e) conditionMessage(e))
tryCatch(deconflate(cull, event_model = TRUE),
         deconflate_error = function(e) conditionMessage(e))
```

## The snapshot model

Within the period of the overall risk, an animal with disease combination
`d` has a constant hazard `h0 * exp(sum_i beta_i * d_i)`, so its risk of the
event over the period is `R(d) = 1 - exp(-h0 * exp(sum_i beta_i * d_i))`.
The baseline `h0` is set so that the population risk, averaged over the
distribution of disease combinations (`fit_joint()`), equals
`overall_risk`. The `beta`s are solved so that the model reproduces each
raw estimate:

* a hazard ratio or rate ratio by the ratio of the average hazard
  multipliers among animals with and without the disease;
* a risk ratio, odds ratio or risk difference by the ratio, odds ratio or
  difference of the average risks `R` among animals with and without the
  disease.

The adjusted hazard ratios are `exp(beta)`. The comparison is made at the
start of the period (the "snapshot"); its estimands are defined by the
model:

* `"snapshot_crude"`: the comparison of animals with and without the
  disease at the start of the period, in the population described by the
  probabilities and associations;
* `"snapshot_stratified"`: the same comparison within strata of the
  diseases in `adjusted_for`, combined across strata (see below).

Because the model uses the whole distribution of disease combinations, it
depends on the maximum-entropy assumption of `fit_joint()` and on any
three-way terms (`vignette("interactions")`). Pairs without an association
are unknown, and the fit fills them in, as in the global method for
additive impacts; the result lists them with their fitted odds ratios
(`$unknown_pairs`).

### What the snapshot model is not

A Cox HR estimated over follow-up is not, in general, the snapshot ratio:
animals with high hazards leave first, so the mixture of disease
combinations among survivors changes over time, and the marginal HR changes
with it. The snapshot model is exact for its own estimands. Entering a
published HR as `"snapshot_crude"` is an approximation that you state, not a
conversion the package makes. It is reasonable when follow-up is short
relative to the hazards or the diseases are rare, and it should be reported
and checked.

## The overall risk

`overall_risk` is the proportion of animals with the event over the period
(e.g. the annual culling rate). It is required. For hazard and rate ratios
it fixes the baseline hazard, and so the attributable risk, but not the
adjusted hazard ratios:

```{r}
risks <- c(0.10, 0.25, 0.40)
sapply(risks, function(r) {
  deconflate(cull, event_model = TRUE, overall_risk = r, n_draws = 0)$adjusted$adjusted
})
```

Risk ratios, odds ratios and risk differences compare risks over a period,
so they depend on the risk level, and the adjusted hazard ratios change
with `overall_risk`. They must refer to the same period as `overall_risk`
(a risk ratio of culling within the lactation needs the culling risk over
the lactation):

```{r}
cull_rr <- cm_model(pop, cm_impacts(c("d1", "d2", "d3"), c(1.4, 1.8, 1.2), measure = "RR",
                                    estimand = "snapshot_crude"))
sapply(risks, function(r) {
  deconflate(cull_rr, event_model = TRUE, overall_risk = r, n_draws = 0)$adjusted$adjusted
})
```

Not every set of estimates fits every overall risk. Risk ratios of 4 for
every disease cannot hold when half of the animals have the event: animals
with d2 would need a risk above 1.

```{r}
cull_rr4 <- cm_model(pop, cm_impacts(c("d1", "d2", "d3"), c(4, 4, 4), measure = "RR",
                                     estimand = "snapshot_crude"))
tryCatch(deconflate(cull_rr4, event_model = TRUE, overall_risk = 0.5, n_draws = 0),
         deconflate_error = function(e) conditionMessage(e))
```

The overall risk can be uncertain: give it as a distribution, and
`deconflate()` draws it with the other uncertain inputs
(`vignette("uncertainty")`). Here only the overall risk is uncertain; the
adjusted hazard ratios do not depend on it, so their intervals have zero
width, while the attributable risk gets an interval. The central estimate
uses the mean of the distribution:

```{r}
deconflate(cull, event_model = TRUE, overall_risk = dist_beta(250, 750),
           n_draws = 200, seed = 1)
```

## Estimates adjusted for other diseases

If a source estimate comes from a model that already included other
diseases, the closest snapshot estimand is `"snapshot_stratified"`, with
those diseases in `adjusted_for` (ids separated by `";"`), or `"all"` for
every other disease. The model value of such an estimate is computed within
strata of its adjustment set and combined across strata with
Mantel-Haenszel-type weights. With `adjusted_for = "all"`, a hazard ratio is
the disease's own hazard multiplier and is used as it is. A Cox coefficient
adjusted for the same diseases matches this only under the snapshot model's
assumptions:

```{r}
cull_adj <- cm_model(pop, cm_impacts(
  c("d1", "d2", "d3"), c(1.5, 2.0, 1.3), measure = "HR",
  estimand = c("snapshot_stratified", "snapshot_crude", "snapshot_stratified"),
  adjusted_for = c("d2", NA, "all")
))
deconflate(cull_adj, event_model = TRUE, overall_risk = 0.25, n_draws = 0)$adjusted
```

The estimate of d1 is a ratio within strata of d2 that still averages over
d3, which it is not adjusted for. Hazard ratios are not collapsible, so it
differs slightly from d1's own hazard multiplier, the adjusted HR (1.509
against the estimate of 1.5).

## Risk attributable to disease

The risk attributable to disease is the overall risk minus the risk of a
disease-free animal, `1 - exp(-h0)`. An animal's risk cannot exceed 1, so an
animal with several diseases is counted once, and the attributable risk is
smaller than the sum of per-disease excess risks when diseases co-occur. It
is allocated to diseases by Shapley values over all disease combinations
(`shapley_by_cell()`); the allocation adds up to the attributable risk
(`unallocated` is zero up to rounding):

```{r}
res$attributable$summary
res$attributable$by_disease
```

A disease whose adjusted hazard ratio is below 1 gets a negative share. The
risks are proportions of animals with the event in the period. Converting
them into numbers of animals or a value (e.g. times the replacement price
less salvage value) is left to the user.

## Comparing methods

`compare_methods(event_model = TRUE)` sets the snapshot model beside two
approximations, each with the attributable risk at `overall_risk`:

* `"snapshot"`: the model of `deconflate()`;
* `"first_order"`: the log-linear approximation `log(HR_raw) = A beta`, with
  the conflation matrix `A` of the additive model. It needs only the
  pairwise tables, so every pair needs an association;
* `"published"`: the approach of Rasmussen et al. (2024), HR - 1 adjusted
  with the proportional approximation (eq. 16 of Rasmussen et al. 2022) and
  1 added back. It is not a method of `deconflate()`; it is kept for
  comparison and reproduction only.

The first-order and published methods need hazard (or rate) ratios, and the
published approach crude estimates. Methods that cannot be run are listed
with the reason:

```{r}
compare_methods(cull, event_model = TRUE, overall_risk = 0.25)
compare_methods(cull_adj, event_model = TRUE, overall_risk = 0.25)$failed
```

## Sensitivity tools

`screen_associations()`, `screen_three_way()`, `sensitivity_oat()` and
`cm_threshold()` take `event_model = TRUE` and `overall_risk` as well. Their
total is then the attributable risk, and disease rankings use its Shapley
allocation. They report point estimates (a distribution of the overall risk
is used at its mean). `sensitivity_oat()` also varies the overall risk:

```{r}
screen_associations(cull, event_model = TRUE, overall_risk = 0.25)
sensitivity_oat(cull, event_model = TRUE, overall_risk = 0.25)
```

In `cm_threshold()`, the overall risk is the input `"risk"`. At what overall
risk does the attributable risk reach 0.04, and what raw hazard ratio of d1
would change the ranking of diseases?

```{r}
cm_threshold(cull, "risk", c(0.05, 0.6), conclusion = "total", target = 0.04,
             event_model = TRUE, overall_risk = 0.25)$thresholds[, c("item", "status", "threshold")]
cm_threshold(cull, "impact:d1", c(0.5, 3), conclusion = "rank",
             event_model = TRUE, overall_risk = 0.25)$thresholds[, c("item", "status", "threshold")]
```

See `vignette("thresholds-and-scaling")` for how threshold searches work.

## Five diseases, mixed measures

The folder `five_diseases` contains an illustrative population of five
diseases (made-up values) and a culling table, `culling.csv`, that mixes
measures and estimands:

```{r}
dir <- system.file("extdata", "five_diseases", package = "deconflate")
read.csv(file.path(dir, "culling.csv"))[, c("disease", "value", "measure", "estimand", "adjusted_for")]
```

LAM has a crude hazard ratio, MAS a hazard ratio stratified by LAM and SCK,
MET a risk ratio and SCK an odds ratio of culling within the lactation, and
RP a risk difference stratified by all other diseases. The risk-based
estimates need the culling risk over the lactation, here taken to be 0.25.
The population includes a three-way term, which the snapshot model uses:

```{r}
five <- cm_read_inputs(diseases = file.path(dir, "diseases.csv"),
                       associations = file.path(dir, "associations.csv"),
                       three_way = file.path(dir, "three_way.csv"),
                       impacts = file.path(dir, "culling.csv"))
five
deconflate(five, event_model = TRUE, overall_risk = 0.25, n_draws = 0)
```

The tables have distributions (the `dist` columns), so the default call,
without `n_draws = 0`, would also report 95% intervals from 1000 draws.
Only the snapshot model applies to these measures:

```{r}
compare_methods(five, event_model = TRUE, overall_risk = 0.25)$failed
```

## The global dairy example

`example_global_dairy("culling")` contains the culling HRs of Rasmussen et
al. (2024) on the 12-disease population of that paper. They are published
Cox HRs, entered as `"snapshot_crude"`; that is an assumption, and the
results below hold under it. The global average replacement rate is 23.66%
(Rasmussen et al. 2024, Table 1). Pairs not in the paper's Table 3 are
unknown, and the joint distribution fills them in. It is fitted once and
passed on with `joint`, so that it can be reused below:

```{r}
gd_cull <- example_global_dairy("culling")
j_gd <- fit_joint(gd_cull)
gd <- deconflate(gd_cull, event_model = TRUE, overall_risk = 0.2366, n_draws = 0, joint = j_gd)
gd
```

For clinical ketosis, dystocia and metritis, the adjusted HRs are below 1:
under the multiplicative model, their raw HRs are smaller than the
associated diseases alone would produce, and `deconflate()` warns about
this. Their shares of the attributable risk are negative.

The 2024 paper instead converted each adjusted HR to an excess risk
relative to the overall rate, `HR * r / (HR * r + 1 - r) - r`, and summed
these excess risks weighted by prevalence. Applied to the adjusted HRs
above, that sum is larger than the attributable risk, because it counts
animals with several diseases more than once:

```{r}
p <- gd_cull$diseases$prob
h <- gd$adjusted$adjusted
c(excess_risk_sum = sum(p * (h * 0.2366 / (h * 0.2366 + 1 - 0.2366) - 0.2366)),
  attributable = gd$attributable$summary$attributable)
```

The paper treated unlisted pairs as independent and used the published
approximation; `reproduce_rasmussen_2024()` recomputes its tables
(`vignette("reproducing-published")`).

## Event impacts alongside additive impacts

Event impacts and additive impacts are adjusted in separate calls: the
additive analyses return impacts in their own units, and the event model
returns hazard ratios and a proportion of animals. To set them side by side,
convert both to a common unit, here a value per cow and year with the
global average yield of 5013 kg, a milk price of 0.5981 per kg, and a
replacement price less culled-cow price of 1299.33 - 785.86 (Rasmussen et
al. 2024, Table 1). The yield analysis has the same population, so the
joint distribution is reused. The yield impacts are percent losses, so the
gap is computed as in `vignette("deconflate")`:

```{r}
yres <- deconflate(example_global_dairy("yield"), n_draws = 0, joint = j_gd)
yield_gap <- 5013 / (1 - yres$totals$adjusted_total / 100) - 5013
c(yield = yield_gap * 0.5981,
  culling = gd$attributable$summary$attributable * (1299.33 - 785.86))
```

The warning says that, with the unknown pairs filled in by the global fit,
the raw yield impacts of four diseases are smaller than their associated
diseases alone would produce under the additive model.

## Methods that are not available for new analyses

The published analyses converted hazard ratios in ways that the package
does not offer for new analyses:

* Rasmussen et al. (2022) treated each HR as an odds ratio of a 2x2 table of
  disease by culling, to obtain an excess annual culling risk, and recovered
  adjusted HRs by rescaling (eq. 23);
* Rasmussen et al. (2024) adjusted HR - 1 as if it were an additive impact,
  and valued culling with the excess risk relative to the overall rate given
  above.

These conversions are used only inside `reproduce_rasmussen_2022()` and
`reproduce_rasmussen_2024()`, so that the published tables can be recomputed
(see `vignette("reproducing-published")`).

## Caveats

* The model assumes a constant hazard within the period and that the
  estimates apply for the whole period. Most source estimates come from
  models in which disease status changes during the period.
* Hazard ratios and odds ratios are not collapsible: even without
  confounding, an average ratio differs from the ratio within subgroups. The
  snapshot model accounts for this given the distribution of disease
  combinations; the first-order method and the published approach do not.
* The snapshot model and the attributable risk use the whole distribution
  of disease combinations, so they depend on the maximum-entropy assumption,
  on how unknown pairs are filled in, and on any three-way terms.
* Culling and death compete. Make sure the source estimates are of the same
  kind (cause-specific or subdistribution), and that `overall_risk` is the
  risk of the same event.
