---
title: "Disease combinations, interactions and attribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Disease combinations, interactions and attribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

## The distribution of disease combinations

`fit_joint()` fits the distribution of all 2^n combinations of n diseases,
using iterative proportional fitting from independence. The result is the
maximum-entropy distribution consistent with the disease probabilities and
the pairwise associations. That is a log-linear model with all two-way
terms and no explicit three-way or higher terms. Three-disease co-occurrence
still differs from independence, through the pairwise terms.

```{r}
j <- fit_joint(example_supplement())
j
head(combination_probs(j))
```

With three or more diseases the constraints do not determine a unique
distribution, which is why the maximum-entropy criterion is needed. If the
pairwise associations cannot hold together, no distribution exists.
`check_feasibility()` detects this and identifies the conflicting
associations:

```{r}
bad <- cm_model(cm_diseases(c("a", "b", "c"), c(0.5, 0.5, 0.5)),
                cm_associations(c("a", "a", "b"), c("b", "c", "c"), c(20, 20, 0.05)))
check_feasibility(bad, method = "triples")
```

## Unknown associations are not independence

Setting an association to an odds ratio of 1 imposes independence. Leaving
it unknown lets the maximum-entropy fit imply an association through the
other diseases:

```{r}
d <- cm_diseases(c("a", "b", "c"), c(0.2, 0.3, 0.25))
a <- cm_associations(c("a", "b"), c("b", "c"), c(3, 2))
imp <- cm_impacts(c("a", "b", "c"), c(0.03, 0.04, 0.05))
ind <- cm_model(d, a, imp, missing_associations = "independent")
unk <- cm_model(d, a, imp, missing_associations = "unknown")
deconflate(ind)$adjusted$adjusted
deconflate(unk, method = "global")$adjusted$adjusted
```

## Impact interactions

An interaction is the extra impact when two diseases occur together:
positive values are synergistic and negative values antagonistic.
Interactions cannot be inferred from associations; they need evidence or
explicit scenarios. They are adjusted with the global method:

```{r}
m <- example_supplement()
mi <- set_interaction(m, "d1", "d2", 0.01, outcome = "yield")
res <- deconflate(mi, method = "global")
res$adjusted[, c("disease", "raw", "adjusted")]
attribute_burden(res)
```

## Attribution

With additive terms, each disease's Shapley value is its own term plus an
equal share of every interaction term it is involved in, which is what
`attribute_burden()` reports. For general loss functions,
`shapley_by_cell()` computes Shapley values cell by cell over the joint
distribution. For example, with proportional impacts that compound:

```{r}
shapley_by_cell(fit_joint(m), loss_multiplicative(c(d1 = 0.02, d2 = 0.04, d3 = 0.06)))
```

## Which interactions would matter?

```{r}
screen_interactions(m, outcome = "yield", values = c(-0.01, 0.01))
```
