---
title: "Disease combinations, interactions and attribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Disease combinations, interactions and attribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

## The distribution of disease combinations

`fit_joint()` fits the distribution of all 2^n combinations of n diseases,
using iterative proportional fitting from independence. The result is the
maximum-entropy distribution consistent with the disease probabilities and
the pairwise associations.

```{r}
m <- example_supplement()
j <- fit_joint(m)
j
head(combination_probs(j))
```

Enumerating 2^n combinations limits this exact fit to about 20 diseases.
For more, `fit_joint(backend = "sampled")` fits the same model by Monte
Carlo (Gibbs sampling with calibrated parameters, followed by raking to
the pairwise tables); the functions below that use the joint distribution
accept a fit from either backend. See
`vignette("thresholds-and-scaling")` for how it works, its diagnostics and
a 24-disease example.

## What maximum entropy assumes

With three or more diseases, the probabilities and pairwise associations do
not determine a unique distribution: there are 2^n - 1 free probabilities
and n(n + 1) / 2 constraints. Maximum entropy picks one solution. It is an
assumption, not something the pairwise evidence identifies:

* the fitted distribution is a log-linear model with all two-way terms and
  no three-way or higher terms. The odds ratio of a pair is the same whether
  or not a third disease is present (given the other diseases);
* three diseases still occur together more (or less) often than under
  independence, through the pairwise terms;
* a pair whose association is unknown has no term of its own. Its
  association is implied by the associations both diseases have with other
  diseases.

Additive results without interactions do not depend on this assumption:
they need the pairwise tables only. It matters for interactions, for
three-way scenarios, and for the results that use the whole distribution
(the snapshot hazard-ratio model and `attributable_risk()`, see
`vignette("culling-hazard-ratios")`).

If the pairwise associations cannot hold together, no distribution exists.
`check_feasibility()` detects this and identifies the conflicting
associations:

```{r}
bad <- cm_population(cm_diseases(c("a", "b", "c"), c(0.5, 0.5, 0.5)),
                     cm_associations(c("a", "a", "b"), c("b", "c", "c"), c(20, 20, 0.05)))
check_feasibility(bad, method = "triples")
```

## Unknown associations are not independence

Setting an association to an odds ratio of 1 imposes independence. Leaving
it unknown lets the maximum-entropy fit imply an association through the
other diseases. Only the global method accepts unknown pairs:

```{r}
d <- cm_diseases(c("a", "b", "c"), c(0.2, 0.3, 0.25))
a <- cm_associations(c("a", "b"), c("b", "c"), c(3, 2))
imp <- cm_impacts(c("a", "b", "c"), c(3, 4, 5), units = "%")
ind <- cm_model(d, imp, associations = a, missing_associations = "independent")
unk <- cm_model(d, imp, associations = a, missing_associations = "unknown")
deconflate(ind)$adjusted$adjusted
compare_methods(unk)
```

The odds ratio of `a` and `c` implied by the fitted distribution:

```{r}
ju <- fit_joint(unk)
p_ac <- sum(ju$prob[ju$cells[, "a"] == 1 & ju$cells[, "c"] == 1])
joint_to_or(p_ac, 0.2, 0.25)
```

## Impact interactions

An interaction is the extra impact when two diseases occur together, in the
units of the impact vector: positive values are synergistic and negative
values antagonistic. Interactions cannot be inferred from associations; they
need evidence or explicit scenarios.

The raw estimate of disease `i` then includes, besides the conflated main
impacts, the part of the interaction burden that goes with `i`. Its expected
value depends on how often pairs occur together with each disease, that is,
on probabilities of disease triples, so interactions need the joint
distribution and the global method:

```{r}
mi <- set_interaction(m, "d1", "d2", 1)   # 1 percentage point
tryCatch(deconflate(mi), deconflate_unsupported = function(e) conditionMessage(e))
res <- deconflate(mi, method = "global")
res
```

The raw impacts satisfy `raw = A %*% adjusted + offset`, where `offset`
holds the interaction part:

```{r}
res$conflation$offset
as.vector(res$conflation$A %*% res$adjusted$adjusted + res$conflation$offset)
res$adjusted$raw
```

A simulated herd with known impacts (2, 4 and 6 percent) and interactions
(1 point for d1 with d2, 1.5 for d2 with d3) checks the method:

```{r}
int <- cm_interactions(c("d1", "d2"), c("d2", "d3"), c(1, 1.5))
raw <- simulate_raw_impacts(m, c(d1 = 2, d2 = 4, d3 = 6), interactions = int)
raw$value
deconflate(cm_model(m, raw, int), method = "global")$adjusted
```

In files, interactions go in `interactions_<analysis>.csv` (see
`vignette("own-data")`).

## Attribution

The expected aggregate is `sum_i p_i b_i + sum_{j<k} P(j and k) delta_jk`.
Removing any disease involved in a term removes that term, so each disease's
Shapley value is its own term plus an equal share of every interaction term
it is involved in. This is the `contributions` table:

```{r}
attribute_burden(res)
```

For general loss functions, `shapley_by_cell()` computes Shapley values cell
by cell over the joint distribution. For example, with proportional impacts
that compound (here as proportions):

```{r}
shapley_by_cell(j, loss_multiplicative(c(d1 = 0.02, d2 = 0.04, d3 = 0.06)))
```

## Three-way scenarios

Pairwise evidence cannot identify how often three diseases occur together.
`cm_three_way()` (or `set_three_way()`) sets a three-way term, the ratio of
conditional odds ratios `OR(d1, d2 | d3 present) / OR(d1, d2 | d3 absent)`.
A ratio of 1 is the maximum-entropy assumption. All pairwise associations
are still matched:

```{r}
m3 <- set_three_way(m, "d1", "d2", "d3", ratio = 2)
m3$three_way
j3 <- fit_joint(m3)
c(maximum_entropy = sum(j$prob[rowSums(j$cells) == 3]),
  ratio_2 = sum(j3$prob[rowSums(j3$cells) == 3]))
```

Three-way terms are sensitivity scenarios. When all pairs of a triple are
constrained, they keep the pairwise tables fixed and change additive results
only through interactions: without interactions every scenario reproduces
the baseline, and with an interaction they change the offset. When a pair is
unknown, its fitted table moves with the three-way term, so additive results
can change even without interactions.
`screen_three_way()` sets each triple in turn:

```{r}
screen_three_way(m, ratios = c(0.5, 2))
screen_three_way(mi, ratios = c(0.5, 2))
```

They also change the joint-based results, such as snapshot hazard ratios:

```{r}
hr <- cm_hazard_ratios(c("d1", "d2", "d3"), c(1.5, 2.0, 1.3), estimand = "snapshot_crude")
rbind(maximum_entropy = deconflate_hr(cm_hr_model(m, hr))$adjusted$adjusted,
      ratio_2 = deconflate_hr(cm_hr_model(m3, hr))$adjusted$adjusted)
```

In files, three-way terms go in `three_way.csv`.

## Which interactions would matter?

`screen_interactions()` adds an interaction of each size to each pair in
turn and reports the change in the aggregate and in the ranking of
diseases:

```{r}
screen_interactions(m, values = c(-1, 1))
```
