---
title: "Disease combinations, interactions and attribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Disease combinations, interactions and attribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

## The distribution of disease combinations

`fit_joint()` fits the distribution of all 2^n combinations of n diseases,
using iterative proportional fitting from independence. The result is the
maximum-entropy distribution consistent with the disease probabilities and
the pairwise associations.

```{r}
m <- example_supplement()
j <- fit_joint(m)
j
head(combination_probs(j))
```

Enumerating 2^n combinations limits this exact fit to about 20 diseases.
For more, `fit_joint(backend = "sampled")` fits the same model by Monte
Carlo (Gibbs sampling with calibrated parameters, followed by raking to
the pairwise tables); `deconflate()` switches to it by itself above 20
diseases, and the functions below that use the joint distribution accept a
fit from either backend. See `vignette("thresholds-and-scaling")` for how it
works, its diagnostics and a 24-disease example.

## What maximum entropy assumes

With three or more diseases, the probabilities and pairwise associations do
not determine a unique distribution: there are 2^n - 1 free probabilities
and n(n + 1) / 2 constraints. Maximum entropy picks one solution. It is an
assumption, not something the pairwise evidence identifies:

* the fitted distribution is a log-linear model with all two-way terms and
  no three-way or higher terms. The odds ratio of a pair is the same whether
  or not a third disease is present (given the other diseases);
* three diseases still occur together more (or less) often than under
  independence, through the pairwise terms;
* a pair whose association is unknown has no term of its own. Its
  association is implied by the associations both diseases have with other
  diseases.

Additive results without interactions do not depend on this assumption when
every pair has an association: they need the pairwise tables only. It
matters for unknown pairs, for interactions, for three-way scenarios, and
for event impacts (`deconflate(..., event_model = TRUE)`, which uses the
whole distribution; see `vignette("event-impacts")`).

If the pairwise associations cannot hold together, no distribution exists.
`check_feasibility()` detects this and identifies the conflicting
associations:

```{r}
bad <- cm_population(cm_diseases(c("a", "b", "c"), c(0.5, 0.5, 0.5)),
                     cm_associations(c("a", "a", "b"), c("b", "c", "c"), c(20, 20, 0.05)))
check_feasibility(bad, method = "triples")
```

## How deconflate() chooses the method

`deconflate()` has two exact methods. The simultaneous method solves the
impact equations from the pairwise 2x2 tables; the global method solves the
same equations from the fitted distribution of disease combinations. They
give the same answer when every pair has an association and there are no
interactions and no three-way terms. With `method = "auto"` (the default),
`deconflate()` uses the simultaneous method in that case and the global
method otherwise, and says why in `$notes`. The reasons are:

* **unknown pairs**: pairs with no association (below);
* **interactions** (below);
* **three-way terms** (below).

`method = "simultaneous"` switches in the same way (with a note saying so),
and `method = "global"` always uses the global method.

## Unknown associations are not independence

A pair without an association is unknown. Giving it an odds ratio of 1
states that the two diseases are unrelated, which is an assumption of its
own. Leaving it unknown lets the maximum-entropy fit imply an association
through the other diseases. Here `a` and `c` are unrelated in one model and
unknown in the other:

```{r}
d <- cm_diseases(c("a", "b", "c"), c(0.2, 0.3, 0.25))
imp <- cm_impacts(c("a", "b", "c"), c(3, 4, 5), units = "%")
ind <- cm_model(d, imp, associations = cm_associations(c("a", "b", "a"), c("b", "c", "c"), c(3, 2, 1)))
unk <- cm_model(d, imp, associations = cm_associations(c("a", "b"), c("b", "c"), c(3, 2)))
deconflate(ind, n_draws = 0)$adjusted$adjusted
res_unk <- deconflate(unk, n_draws = 0)
res_unk
```

`$unknown_pairs` lists the unknown pairs and the odds ratios the global fit
gave them. Both `a` and `c` are positively associated with `b`, so the fit
implies a positive association between them too:

```{r}
res_unk$unknown_pairs
```

`compare_methods()` runs each method as asked, without switching, and
reports why the pairwise methods cannot run (the `"published"`
approximation is listed for comparison only):

```{r}
compare_methods(unk)
```

## Impact interactions

An interaction is the extra impact when two diseases occur together, in the
units of the impact table: positive values are synergistic and negative
values antagonistic. Interactions cannot be inferred from associations; they
need evidence or explicit scenarios. They apply to additive impacts only.

The raw estimate of disease `i` then includes, besides the conflated main
impacts, the part of the interaction burden that goes with `i`. Its expected
value depends on how often pairs occur together with each disease, that is,
on probabilities of disease triples, so interactions need the joint
distribution. `deconflate()` therefore uses the global method:

```{r}
mi <- set_interaction(m, "d1", "d2", 1)   # 1 percentage point
res <- deconflate(mi, n_draws = 0)
res
```

The raw impacts satisfy `raw = A %*% adjusted + offset`, where `offset`
holds the interaction part:

```{r}
res$conflation$offset
as.vector(res$conflation$A %*% res$adjusted$adjusted + res$conflation$offset)
res$adjusted$raw
```

A simulated herd with known impacts (2, 4 and 6 percent) and interactions
(1 point for d1 with d2, 1.5 for d2 with d3) checks the method:

```{r}
int <- cm_interactions(c("d1", "d2"), c("d2", "d3"), c(1, 1.5))
raw <- simulate_raw_impacts(m, c(d1 = 2, d2 = 4, d3 = 6), interactions = int)
raw$value
deconflate(cm_model(m, raw, int), n_draws = 0)$adjusted
```

In files, interactions are a table of their own (`disease1`, `disease2`,
`value`), passed to `cm_read_inputs(interactions = )` with the impact table
they belong to (see `vignette("own-data")`). The five-disease example has
one for its welfare scores:

```{r}
ex <- system.file("extdata", "five_diseases", package = "deconflate")
welfare <- cm_read_inputs(diseases = file.path(ex, "diseases.csv"),
                          associations = file.path(ex, "associations.csv"),
                          impacts = file.path(ex, "impacts_welfare.csv"),
                          interactions = file.path(ex, "interactions_welfare.csv"))
welfare$interactions[, c("disease1", "disease2", "value")]
res_w <- deconflate(welfare, n_draws = 0)
res_w$notes
res_w$totals
```

## Attribution

The expected aggregate is `sum_i p_i b_i + sum_{j<k} P(j and k) delta_jk`.
Removing any disease involved in a term removes that term, so each disease's
Shapley value is its own term plus an equal share of every interaction term
it is involved in. This is the `contributions` table:

```{r}
attribute_burden(res)
```

For general loss functions, `shapley_by_cell()` computes Shapley values cell
by cell over the joint distribution. For example, with proportional impacts
that compound (here as proportions):

```{r}
shapley_by_cell(j, loss_multiplicative(c(d1 = 0.02, d2 = 0.04, d3 = 0.06)))
```

## Three-way scenarios

Pairwise evidence cannot identify how often three diseases occur together.
`cm_three_way()` (or `set_three_way()`) sets a three-way term, the ratio of
conditional odds ratios `OR(d1, d2 | d3 present) / OR(d1, d2 | d3 absent)`.
A ratio of 1 is the maximum-entropy assumption. All pairwise associations
are still matched:

```{r}
m3 <- set_three_way(m, "d1", "d2", "d3", ratio = 2)
m3$three_way
j3 <- fit_joint(m3)
c(maximum_entropy = sum(j$prob[rowSums(j$cells) == 3]),
  ratio_2 = sum(j3$prob[rowSums(j3$cells) == 3]))
```

Three-way terms are scenarios: they explore an assumption the data cannot
check, rather than describe an estimate. A model with a three-way term
needs the joint distribution, so `deconflate()` uses the global method.
When all pairs of a triple have an association, the term keeps the pairwise
tables fixed and changes additive results only through interactions:
without interactions the scenario reproduces the baseline.

```{r}
res3 <- deconflate(m3, n_draws = 0)
res3$notes
all.equal(res3$adjusted$adjusted, deconflate(m, n_draws = 0)$adjusted$adjusted)
```

With an interaction they change the offset. When a pair is unknown, its
fitted table moves with the three-way term, so additive results can change
even without interactions. `screen_three_way()` sets each triple in turn:

```{r}
screen_three_way(m, ratios = c(0.5, 2))
screen_three_way(mi, ratios = c(0.5, 2))
```

They also change results that use the whole distribution, such as event
impacts (here hazard ratios of culling, with an overall culling risk of
0.25; see `vignette("event-impacts")`):

```{r}
hr <- cm_impacts(c("d1", "d2", "d3"), c(1.5, 2.0, 1.3), measure = "HR",
                 estimand = "snapshot_crude", label = "culling")
e0 <- deconflate(cm_model(m, hr), event_model = TRUE, overall_risk = 0.25, n_draws = 0)
e3 <- deconflate(cm_model(m3, hr), event_model = TRUE, overall_risk = 0.25, n_draws = 0)
rbind(maximum_entropy = setNames(e0$adjusted$adjusted, e0$adjusted$disease),
      ratio_2 = setNames(e3$adjusted$adjusted, e3$adjusted$disease))
```

In files, three-way terms are a table of their own (`disease1`,
`disease2`, `disease3`, `ratio`), passed to `cm_read_inputs(three_way = )`;
see `three_way.csv` in the five-disease example.

## Which interactions would matter?

`screen_interactions()` adds an interaction of each size to each pair in
turn and reports the change in the aggregate and in the ranking of
diseases:

```{r}
screen_interactions(m, values = c(-1, 1))
```
