---
title: "Using your own data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using your own data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

Inputs can be kept in CSV files (one per table) or typed in R as data
frames. `cm_read_inputs()` checks every table, reports all problems at once,
and returns the population, the model (one analysis per impact table), a
hazard-ratio model if there are hazard ratios, and a Monte Carlo sampler if
any value has a distribution.

## Files

| File | Contents | Required columns |
|---|---|---|
| `diseases.csv` | one row per disease | `id`, `value` |
| `associations.csv` | one row per associated pair | `disease1`, `disease2` |
| `three_way.csv` | optional three-way association scenarios | `disease1`, `disease2`, `disease3`, `ratio` |
| `impacts.csv` | the impact table of a single analysis | `disease`, `value` |
| `impacts_<analysis>.csv` | for several analyses, one impact table each, e.g. `impacts_yield.csv` | `disease`, `value` |
| `interactions.csv`, `interactions_<analysis>.csv` | optional impact interactions | `disease1`, `disease2`, `value` |
| `hazard_ratios.csv` | optional culling (or mortality) hazard ratios | `disease`, `value`, `estimand` |

Only `diseases.csv` is required. A single `impacts.csv` is read as one
analysis named `impacts`. All analyses share the diseases, associations and
three-way terms. The optional columns are described in `?cm_read_inputs`:

* `diseases.csv`: `type` (`prevalence`, `probability` or `incidence_rate`),
  `time_horizon`, `reference_population`, `source`;
* `associations.csv`: `value`, `measure` (`OR`, `RR`, `RD`, `cond_prob`,
  `phi`, `table`, `independent` or `unknown`), the counts `n11`, `n10`, `n01`,
  `n00` for `measure = table`, `adjusted`, `adjusted_for`, `source`;
* impact tables: `estimand` (`crude` or `adjusted_linear`), `adjusted_for`,
  `source`, and one `label` and one `units` per table;
* `hazard_ratios.csv`: `adjusted_for`, `source`. The `estimand` column is
  required (`snapshot_crude` or `snapshot_stratified`; see below).

Every disease needs a row in every impact table (use 0 for no impact) and in
the hazard-ratio table (use 1 for no effect). Impact values can be in any
units, the same within a table; results come back in those units.

There is no separate uncertainty file: an uncertain value gets its
distribution in its own row, in the columns `dist` and `p1`-`p4` (see
Uncertainty below), so point values and uncertain values sit side by side.
Every table can also have a free-text `note` column.

## Start from the template

`cm_template()` writes an example set of files to a folder. By default it is
a single analysis, the standard workflow:

```{r}
dir <- file.path(tempdir(), "my-inputs")
cm_template(dir, overwrite = TRUE)
list.files(dir)
read.csv(file.path(dir, "impacts.csv"))
```

`interactions.csv` and `three_way.csv` are empty (header only); fill them in
if needed.

## Read the files

```{r}
inp <- cm_read_inputs(dir = dir)
inp
```

The result holds:

* `population`: the diseases, associations and three-way terms, as a
  `cm_population()` object;
* `model`: the `cm_model()` when there is exactly one analysis (otherwise
  `NULL`);
* `analyses`: a `cm_analyses()` object with one model per impact table
  (here one, named `impacts`);
* `hr_model`: the hazard-ratio model, for `deconflate_hr()` (`NULL` without
  `hazard_ratios.csv`);
* `sampler`: a `cm_sampler()` for one analysis, or a `cm_batch_sampler()`
  for several, for `cm_monte_carlo()`.

```{r}
inp$population
deconflate(inp$model)
```

## From the files to a Monte Carlo run

In the template, four values have a distribution in their row: the lameness
prevalence (`diseases.csv`), the lameness:subclinical ketosis odds ratio
(`associations.csv`) and two yield impacts (`impacts.csv`). The other values
are point values (empty `dist`). With one analysis, the sampler is a
`cm_sampler()`:

```{r}
read.csv(file.path(dir, "diseases.csv"))[, c("id", "value", "type", "dist", "p1", "p2")]
read.csv(file.path(dir, "associations.csv"))[, c("disease1", "disease2", "value", "dist", "p1", "p2", "p3")]
inp$sampler
mc <- cm_monte_carlo(inp$sampler, 200, seed = 1)
mc
summary(mc, diagnose = FALSE)[, c("disease", "mean", "mcse", "q0.025", "q0.975")]
```

A single-analysis sampler supports Latin hypercube sampling
(`sampling = "lhs"`) and importance sampling. For importance sampling, each
proposal must cover the whole support of the input's own distribution. A
defensive mixture, part the input's own distribution and part a wider or
shifted one, does so by construction. Here the lameness yield impact is
sampled more often around 6%:

```{r}
specs <- attr(inp$sampler, "specs")
prop <- list("impact:LAM" = dist_mixture(specs[["impact:LAM"]], dist_normal(6, 1.5),
                                         weights = c(0.5, 0.5)))
mc_is <- cm_monte_carlo(inp$sampler, 200, proposal = prop, seed = 1)
mc_is
summary(mc_is, diagnose = FALSE)[, c("disease", "mean", "mcse", "q0.025", "q0.975")]
```

The draws are weighted by the ratio of the densities, so the estimates still
refer to the input's own distribution; compare the standard errors (`mcse`)
to see whether the proposal helped. See `vignette("uncertainty")`.

## Several analyses

`cm_template(type = "analyses")` writes two analyses on one population,
milk yield (in percent) and calving interval (in days):

```{r}
dir2 <- file.path(tempdir(), "my-analyses")
cm_template(dir2, type = "analyses", overwrite = TRUE)
list.files(dir2)
read.csv(file.path(dir2, "impacts_calving_interval.csv"))
inp2 <- cm_read_inputs(dir = dir2)
inp2
names(inp2$analyses$models)
deconflate(inp2$analyses)
```

With several analyses, `model` is `NULL` and the sampler is a batch sampler:
every draw of the shared disease and association inputs is used by both
analyses. Batch runs use simple random sampling.

```{r}
inp2$sampler
mcb <- cm_monte_carlo(inp2$sampler, 200, seed = 1)
mcb
s <- summary(mcb, diagnose = FALSE)
s[, c("analysis", "disease", "mean", "q0.025", "q0.975")]
```

## Hazard ratios

Hazard ratios go in `hazard_ratios.csv`. They are adjusted by a separate
model (`vignette("culling-hazard-ratios")`), and the `estimand` column
states how each value is to be read: `snapshot_crude`, or
`snapshot_stratified` with `adjusted_for`. Entering a published Cox hazard
ratio as either is an approximation and your assumption, so there is no
default:

```{r}
write.csv(data.frame(disease = c("LAM", "SCK", "MET"), value = c(1.74, 1.5, 1.9),
                     estimand = "snapshot_crude", source = "Illustrative"),
          file.path(dir, "hazard_ratios.csv"), row.names = FALSE)
inp_hr <- cm_read_inputs(dir = dir)
deconflate_hr(inp_hr$hr_model)
```

A table without the `estimand` column, or with the names used before
version 0.3.0 (`crude`, `adjusted`), is reported:

```{r}
d3 <- data.frame(id = c("LAM", "SCK", "MET"), value = c(0.25, 0.38, 0.10))
cm_check_inputs(diseases = d3,
                hazard_ratios = data.frame(disease = c("LAM", "SCK", "MET"),
                                           value = c(1.74, 1.5, 1.9)))
cm_check_inputs(diseases = d3,
                hazard_ratios = data.frame(disease = c("LAM", "SCK", "MET"),
                                           value = c(1.74, 1.5, 1.9), estimand = "crude"))
```

## When something is wrong

All problems are listed with their table, row (counting from the first row
below the header) and column, so that they can be fixed in one pass.
`cm_check_inputs()` returns them without stopping:

```{r}
cm_check_inputs(
  diseases = data.frame(id = c("LAM", "SCK", "SCK"), value = c(0.25, 1.4, 0.3)),
  associations = data.frame(disease1 = "LAM", disease2 = "MET", value = 2.0),
  impacts = data.frame(disease = c("LAM", "SCK"), value = c("4.81", "8,40"))
)
```

`cm_read_inputs()` stops with the same list (an error of class
`deconflate_input_problems`) when there are errors, and prints notes
otherwise. The installed example with errors shows more cases:

```{r}
cm_check_inputs(dir = system.file("extdata", "example_with_errors", package = "deconflate"))
```

## The global dairy example

The folder `global_dairy_2024` holds the inputs of Rasmussen et al. (2024):
`diseases.csv` (12 diseases), `associations.csv` (38 odds ratios),
`impacts_yield.csv` and `impacts_fertility.csv` (in percent) and
`hazard_ratios.csv` (culling, entered as `snapshot_crude`). The input
distributions of the paper's Monte Carlo analysis are in the `dist` and
`p1`-`p4` columns of the association and impact tables.

```{r}
gd <- cm_read_inputs(dir = system.file("extdata", "global_dairy_2024", package = "deconflate"))
gd
cmp <- compare_methods(gd$analyses, methods = c("published", "simultaneous"))
cmp$impacts[cmp$impacts$analysis == "yield", ]
```

The `"published"` column (the approximation of the paper, kept for
comparison only) and the simultaneous solution differ; the simultaneous
solution changes the sign of several adjusted impacts (see
`vignette("reproducing-published")`). The culling hazard ratios are adjusted
separately with the snapshot model (`vignette("culling-hazard-ratios")`), and
`compare_methods()` sets the paper's approach beside it:

```{r}
hr_gd <- deconflate_hr(gd$hr_model)
hr_gd
compare_methods(gd$hr_model, methods = c("published", "snapshot"), joint = hr_gd$joint)$impacts
```

## Tables typed in R

Any table can be a data frame instead of a file. Tables that come one per
analysis (`impacts`, `interactions`) are named lists, named after the
analyses. Distributions are columns of the tables, as in the files:

```{r}
inp3 <- cm_read_inputs(
  diseases = data.frame(id = c("d1", "d2", "d3"), value = c(0.10, 0.15, 0.20)),
  associations = data.frame(disease1 = c("d1", "d2"), disease2 = c("d2", "d3"),
                            value = c(2, 3), measure = "OR",
                            dist = c("lognormal_ci", NA), p1 = c(2, NA), p2 = c(1.4, NA),
                            p3 = c(2.9, NA)),
  impacts = list(
    yield = data.frame(disease = c("d1", "d2", "d3"), value = c(2.5, 5, 7.5),
                       units = "% of yield", dist = c("normal", NA, NA),
                       p1 = c(2.5, NA, NA), p2 = c(0.5, NA, NA)),
    fertility = data.frame(disease = c("d1", "d2", "d3"), value = c(1, 2, 0),
                           units = "% of calving interval")
  )
)
inp3
inp3$sampler
deconflate(inp3$analyses)$yield$adjusted
```

With a single analysis (e.g. `impacts = list(yield = ...)`, or one data
frame, which becomes the analysis `impacts`), `$model` is set and `$sampler`
is a single-analysis sampler.

## Associations from other designs

### Adjusted association measures

The 2x2 algebra needs marginal (crude) association measures. An odds ratio
from a multivariable model is conditional on its covariates, so it is
rejected unless you choose to use it as if it were marginal, which is an
approximation:

```{r}
d12 <- cm_diseases(c("d1", "d2"), c(0.10, 0.15))
a_adj <- cm_associations("d1", "d2", 1.8, adjusted = TRUE, adjusted_for = "parity; herd")
tryCatch(cm_population(d12, a_adj), deconflate_unsupported = function(e) conditionMessage(e))
cm_population(d12, a_adj, adjusted_associations = "use_as_marginal")
```

In files, set the column `adjusted` to `TRUE` and call
`cm_read_inputs(..., adjusted_associations = "use_as_marginal")`; each such
row is then reported as a note.

### Contingency tables and zero cells

A study table can be given as counts (`measure = "table"`). Its odds ratio is
applied to the modelled population's own probabilities. A table with a zero
cell has no finite odds ratio, so by default 0.5 is added to every cell
(Haldane-Anscombe correction), and the row is flagged in column `corrected`:

```{r}
tab <- cm_associations("d1", "d2", measure = "table", n11 = 0, n10 = 30, n01 = 40, n00 = 930)
tab[, c("disease1", "disease2", "measure", "value", "corrected")]
tryCatch(cm_associations("d1", "d2", measure = "table", n11 = 0, n10 = 30, n01 = 40,
                         n00 = 930, zero_cell = "error"),
         deconflate_error = function(e) conditionMessage(e))
```

When reading files, the correction is applied and reported as a note:

```{r}
cm_check_inputs(
  diseases = data.frame(id = c("d1", "d2"), value = c(0.10, 0.15)),
  associations = data.frame(disease1 = "d1", disease2 = "d2", measure = "table",
                            n11 = 0, n10 = 30, n01 = 40, n00 = 930)
)
```

Tables in which every count is zero are rejected.

## Disease ids

Ids are used in keys such as `assoc:d1:d2` and in lists such as `"d1; d2"`,
so they must not contain `|`, `;` or `:`, nor leading or trailing spaces.
`"all"` is reserved (it means every other disease in `adjusted_for`):

```{r}
tryCatch(cm_diseases(c("LAM", "SCK:early"), c(0.25, 0.40)),
         deconflate_error = function(e) conditionMessage(e))
tryCatch(cm_diseases(c("LAM", "all"), c(0.25, 0.40)),
         deconflate_error = function(e) conditionMessage(e))
```

## Uncertainty

The tables `diseases`, `associations`, `three_way`, `impacts` and
`interactions` can have the columns `dist` and `p1`-`p4`. A row with a
`dist` has an uncertain value; a row with an empty `dist` is a point value.
The deterministic methods use the point value (`value`, or `ratio` for a
three-way term); Monte Carlo runs draw from the distribution, on the scale
the value was entered on:

* a disease's `value` (e.g. a prevalence or an incidence rate);
* an association's `value` (e.g. an odds ratio); the measure must be
  numeric (`OR`, `RR`, `RD`, `cond_prob` or `phi`), not `table`,
  `independent` or `unknown`;
* a three-way `ratio`;
* an impact or an interaction, in the units of its table.

Hazard ratios cannot have distributions (the Monte Carlo tools adjust
additive impacts). `dist` and the parameters `p1`-`p4`:

| `dist` | `p1` | `p2` | `p3` | `p4` |
|---|---|---|---|---|
| `fixed` | value | | | |
| `normal` | mean | sd | lower bound (optional) | upper bound (optional) |
| `lognormal` | meanlog | sdlog | | |
| `lognormal_ci` | estimate | lower CI | upper CI | level (default 0.95) |
| `beta` | shape1 | shape2 | min (default 0) | max (default 1) |
| `pert` | min | mode | max | lambda (default 4) |
| `pert_mean` | min | mean | max | lambda (default 4) |
| `uniform` | min | max | | |

The checks report a distribution with missing or invalid parameters,
parameters without a `dist`, and a distribution on a non-numeric
association measure; a point value outside its distribution's range is
noted. In the sampler, inputs are keyed `prob:<disease>`,
`assoc:<d1>:<d2>`, `three:<d1>:<d2>:<d3>`, `impact:<disease>` and
`inter:<d1>:<d2>` (the keys of `mc$params`, of proposals in importance
sampling and of `cm_scenario()`):

```{r}
names(attr(inp2$sampler$samplers$yield, "specs"))
```

Files named `uncertainty.csv` or `uncertainty_<analysis>.csv`, used by
version 0.2, are reported as an error: move each distribution into the row
of its value.

`cm_dist_table()` builds distributions from a table with columns `key`,
`dist` and `p1`-`p4`, if you want to pass them to `cm_sampler()` yourself.
See `vignette("uncertainty")` for the Monte Carlo analysis.

## A fuller example

The folder `five_diseases` uses every feature on five diseases: all
probability types and association measures, a three-way term, crude and
adjusted impacts, interactions, hazard ratios, and every distribution type
mixed with point values. Its `run_all_features.R` script runs every part of
the package on it:

```{r}
five <- system.file("extdata", "five_diseases", package = "deconflate")
list.files(five)
inp5 <- cm_read_inputs(dir = five)
inp5
read.csv(file.path(five, "impacts_yield.csv"))[, c("disease", "value", "dist", "p1", "p2", "p3")]
```
