---
title: "Using your own data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using your own data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

Inputs are tables: the diseases, their associations, one impact table and,
if needed, interactions and three-way terms. Each table can be a CSV file
(with any name) or a data frame typed in R. `cm_read_inputs()` checks every
table, reports all problems at once, and returns a model (`cm_model()`) that
`deconflate()` adjusts.

## The tables

Each table is a named argument of `cm_read_inputs()`:

| Argument | Contents | Required columns |
|---|---|---|
| `diseases` | one row per disease | `id`, `value` |
| `associations` | one row per pair with an association estimate | `disease1`, `disease2`, `value` |
| `impacts` | the raw impact estimates of one outcome, one row per disease | `disease`, `value` |
| `interactions` | optional pairwise impact interactions (additive impacts only) | `disease1`, `disease2`, `value` |
| `three_way` | optional three-way association terms | `disease1`, `disease2`, `disease3`, `ratio` |

`diseases` and `impacts` are required. `deconflate()` also needs at least
one association; without any, the sensitivity tools (`screen_associations()`,
`cm_threshold()`) show how much associations could matter. The optional
columns are described in `?cm_read_inputs`:

* `diseases`: `type` (`prevalence` (default), `probability` or
  `incidence_rate`), `time_horizon`, `reference_population`, `source`;
* `associations`: `measure` (`OR` (default), `RR`, `RD`, `cond_prob` or
  `phi`), `adjusted`, `adjusted_for`, `source`;
* `impacts`: `estimand` (`crude` (default) or `adjusted_linear`),
  `adjusted_for`, `source`, and one `label` and one `units` per table; a
  `measure` column makes them event impacts (see below);
* `interactions` and `three_way`: `source`.

Every table can also have a free-text `note` column, and the columns `dist`
and `p1`-`p4` for uncertain values (see Uncertainty below). Column names are
not case-sensitive.

Every disease needs a row in the impact table (use 0 for no impact).
Additive impacts can be in any units, the same for every row; results come
back in those units.

## Start from the template

`cm_template()` writes an example set of files to a folder:

```{r}
dir <- file.path(tempdir(), "my-inputs")
cm_template(dir, overwrite = TRUE)
list.files(dir)
read.csv(file.path(dir, "yield.csv"))
```

The files are three diseases (`diseases.csv`), their odds ratios
(`associations.csv`), additive impacts on milk yield in percent
(`yield.csv`), event impacts on culling (`culling.csv`, see below), and an
empty interactions table (`yield_interactions.csv`) and three-way table
(`three_way.csv`), with a header only. Rename them, or add files for other
outcomes, as you like: the file names carry no meaning.

## Read the files

Pass each file to its argument:

```{r}
yield <- cm_read_inputs(
  diseases = file.path(dir, "diseases.csv"),
  associations = file.path(dir, "associations.csv"),
  impacts = file.path(dir, "yield.csv"),
  interactions = file.path(dir, "yield_interactions.csv"),
  three_way = file.path(dir, "three_way.csv")
)
yield
deconflate(yield, n_draws = 0)
```

Empty tables (here the interactions and three-way terms) are the same as
leaving the argument out. Some values in the template have a distribution,
so `deconflate()` would by default also run 1000 draws for intervals;
`n_draws = 0` gives point estimates only (see Uncertainty below).

## Associations

### Unknown pairs and odds ratios of 1

A pair of diseases with no row in the association table is **unknown**.
Then `deconflate()` uses the global method, which fills in the association
of unknown pairs from the other associations (the maximum-entropy
distribution of disease combinations), and says so in a note. To state that
two diseases are unrelated, give the pair an odds ratio of 1 (with no
distribution).

Here the MET:SCK row is left out:

```{r}
a <- read.csv(file.path(dir, "associations.csv"))
a[, c("disease1", "disease2", "value", "measure")]
y_unknown <- cm_read_inputs(diseases = file.path(dir, "diseases.csv"),
                            associations = a[-2, ],
                            impacts = file.path(dir, "yield.csv"))
y_unknown
r_unknown <- deconflate(y_unknown, n_draws = 0)
r_unknown
r_unknown$unknown_pairs
```

`$unknown_pairs` lists the pairs without an association and the odds ratio
the global fit gave them. With an odds ratio of 1 for MET:SCK instead, every
pair has an association and the simultaneous method is used:

```{r}
a_indep <- a
a_indep$value[2] <- 1
y_indep <- cm_read_inputs(diseases = file.path(dir, "diseases.csv"),
                          associations = a_indep,
                          impacts = file.path(dir, "yield.csv"))
deconflate(y_indep, n_draws = 0)
```

The two are different statements: an unknown pair is filled in from the
other associations, while an odds ratio of 1 is an estimate of no
association. Some published analyses treated unlisted pairs as independent;
to do the same, add those pairs with an odds ratio of 1.

### Measures

The measures are `OR` (default), `RR`, `RD`, `cond_prob` (P(disease1 |
disease2)) and `phi`. Files written for earlier versions may use the
measures `unknown`, `independent` or `table`, or the count columns `n11`,
`n10`, `n01` and `n00`; these are reported with what to do instead:

```{r}
cm_check_inputs(
  diseases = data.frame(id = c("d1", "d2", "d3"), value = c(0.10, 0.15, 0.20)),
  associations = data.frame(disease1 = c("d1", "d1", "d2"), disease2 = c("d2", "d3", "d3"),
                            value = c(NA, 1, NA), measure = c("unknown", "independent", "table")),
  impacts = data.frame(disease = c("d1", "d2", "d3"), value = c(2.5, 5, 7.5))
)
```

For a study's 2x2 table, compute the odds ratio yourself, `n11 * n00 /
(n10 * n01)` (both diseases, only the first, only the second, neither), and
enter it with measure `OR`. If a cell is zero, decide how to handle it (e.g.
add 0.5 to every cell) before computing the odds ratio.

### Adjusted association measures

The 2x2 algebra needs marginal (crude) association measures. An odds ratio
from a multivariable model is conditional on its covariates, so it is
rejected unless you choose to use it as if it were marginal, which is an
approximation:

```{r}
d12 <- cm_diseases(c("d1", "d2"), c(0.10, 0.15))
a_adj <- cm_associations("d1", "d2", 1.8, adjusted = TRUE, adjusted_for = "parity; herd")
tryCatch(cm_population(d12, a_adj), deconflate_unsupported = function(e) conditionMessage(e))
cm_population(d12, a_adj, adjusted_associations = "use_as_marginal")
```

In files, set the column `adjusted` to `TRUE` and call
`cm_read_inputs(..., adjusted_associations = "use_as_marginal")`; each such
row is then reported as a note.

## Several outcomes

An impact table holds one outcome, and `deconflate()` adjusts one impact
table per call. For several outcomes, keep one impact table per outcome and
repeat the call; the disease and association tables are shared. The
installed folder `five_diseases` has impact tables for milk yield (in
percent) and calving interval (in days), among others:

```{r}
five <- system.file("extdata", "five_diseases", package = "deconflate")
list.files(five)
outcomes <- c(yield = "impacts_yield.csv", calving_interval = "impacts_calving_interval.csv")
results <- lapply(outcomes, function(f) {
  m <- cm_read_inputs(diseases = file.path(five, "diseases.csv"),
                      associations = file.path(five, "associations.csv"),
                      three_way = file.path(five, "three_way.csv"),
                      impacts = file.path(five, f))
  deconflate(m, n_draws = 0)
})
results$calving_interval
```

The results can then be combined in base R, e.g. into one long table:

```{r}
do.call(rbind, lapply(names(results), function(o) {
  data.frame(outcome = o, units = results[[o]]$units,
             results[[o]]$adjusted[, c("disease", "raw", "adjusted")])
}))
```

## Event impacts

Impacts on an event such as culling or death are event impacts: hazard
ratios, rate ratios, risk ratios, odds ratios or risk differences. Their
table has a `measure` column (`HR`, `rate_ratio`, `RR`, `OR` or `RD`, which
can be mixed, one row per disease) and an `estimand` (`snapshot_crude`, or
`snapshot_stratified` with `adjusted_for`). They are adjusted with
`event_model = TRUE`, which needs the overall risk of the event in the
population over the period; risk ratios, odds ratios and risk differences
must refer to the same period. The template's `culling.csv`:

```{r}
read.csv(file.path(dir, "culling.csv"))[, c("disease", "value", "measure", "estimand")]
culling <- cm_read_inputs(diseases = file.path(dir, "diseases.csv"),
                          associations = file.path(dir, "associations.csv"),
                          impacts = file.path(dir, "culling.csv"))
deconflate(culling, event_model = TRUE, overall_risk = 0.25, n_draws = 0)
```

The adjusted hazard ratios are each disease's own effect; the result also
gives the risk attributable to disease and its split by disease. If an
adjusted hazard ratio falls on the other side of 1 from its raw estimate,
`deconflate()` warns: the raw estimate is weaker than the disease's
associations alone would produce, so check the estimate and its source
population.

A table with a `measure` column needs `event_model = TRUE`, and
`event_model = TRUE` needs a `measure` column:

```{r}
tryCatch(deconflate(culling, n_draws = 0), deconflate_unsupported = function(e) conditionMessage(e))
```

See `vignette("event-impacts")` for the model and its results.

## Uncertainty

There is no separate uncertainty file: an uncertain value gets its
distribution in its own row, in the columns `dist` and `p1`-`p4`, so point
values and uncertain values sit side by side. A row with an empty `dist` is
a point value. In the template, the LAM prevalence, the LAM:SCK odds ratio
and two yield impacts have a distribution:

```{r}
read.csv(file.path(dir, "diseases.csv"))[, c("id", "value", "type", "dist", "p1", "p2")]
read.csv(file.path(dir, "associations.csv"))[, c("disease1", "disease2", "value", "dist", "p1", "p2", "p3")]
```

`cm_read_inputs()` keeps them in the model, and `deconflate()` draws them
(1000 draws by default) and adds 95% intervals to the results:

```{r}
deconflate(yield, n_draws = 200, seed = 1)
```

The central estimate always uses the point values (`value`); the draws give
the intervals. Values are drawn on the scale they were entered on:

* a disease's `value` (e.g. a prevalence or an incidence rate);
* an association's `value`, on its own measure (e.g. an odds ratio);
* a three-way `ratio`;
* an impact or an interaction, in the units of its table (for event
  impacts, on the scale of the measure, e.g. a hazard ratio).

`dist` and the parameters `p1`-`p4`:

| `dist` | `p1` | `p2` | `p3` | `p4` |
|---|---|---|---|---|
| `fixed` | value | | | |
| `normal` | mean | sd | lower bound (optional) | upper bound (optional) |
| `lognormal` | meanlog | sdlog | | |
| `lognormal_ci` | estimate | lower CI | upper CI | level (default 0.95) |
| `beta` | shape1 | shape2 | min (default 0) | max (default 1) |
| `pert` | min | mode | max | lambda (default 4) |
| `pert_mean` | min | mean | max | lambda (default 4) |
| `uniform` | min | max | | |

The checks report a distribution with missing or invalid parameters and
parameters without a `dist`; a point value outside its distribution's
range is noted. In the model, the distributions are kept in
`$distributions`, keyed `prob:<disease>`, `assoc:<d1>:<d2>`,
`three:<d1>:<d2>:<d3>`, `impact:<disease>` and `inter:<d1>:<d2>`, the same
keys as `cm_model(distributions = )`:

```{r}
names(yield$distributions)
```

`cm_dist_table()` builds distributions from a table with columns `key`,
`dist` and `p1`-`p4`. See `vignette("uncertainty")` for the draws, their
intervals and rejected draws.

## When something is wrong

All problems are listed with their table, row (counting from the first row
below the header) and column, so that they can be fixed in one pass.
`cm_check_inputs()` returns them without stopping:

```{r}
cm_check_inputs(
  diseases = data.frame(id = c("LAM", "SCK", "SCK"), value = c(0.25, 1.4, 0.3)),
  associations = data.frame(disease1 = "LAM", disease2 = "MET", value = 2.0),
  impacts = data.frame(disease = c("LAM", "SCK"), value = c("4.81", "8,40"))
)
```

`cm_read_inputs()` stops with the same list (an error of class
`deconflate_input_problems`) when there are errors, and prints notes
otherwise.

The installed folder `example_with_errors` has a disease table, an
association table, an additive impact table (`impacts_yield.csv`) and an
event impact table (`culling.csv`), each with mistakes explained in its
`note` column: a missing distribution parameter, a prevalence above 1, a
duplicate disease or pair, a missing association value, an unknown disease,
a covariate-adjusted odds ratio, measures and count columns from earlier
versions, a decimal comma, two units in one table, an unsupported
distribution, and a point value outside its distribution:

```{r}
err <- system.file("extdata", "example_with_errors", package = "deconflate")
list.files(err)
cm_check_inputs(diseases = file.path(err, "diseases.csv"),
                associations = file.path(err, "associations.csv"),
                impacts = file.path(err, "impacts_yield.csv"))
```

The event impact table has a missing estimand, a hazard ratio of 0, and a
stratified estimand without `adjusted_for` (the disease table's problems,
listed above, are left out here):

```{r}
p <- cm_check_inputs(diseases = file.path(err, "diseases.csv"),
                     impacts = file.path(err, "culling.csv"))
p[p$table == "impacts", ]
```

## Tables typed in R

Any table can be a data frame instead of a file, with the same columns,
including the distribution columns:

```{r}
typed <- cm_read_inputs(
  diseases = data.frame(id = c("d1", "d2", "d3"), value = c(0.10, 0.15, 0.20)),
  associations = data.frame(disease1 = c("d1", "d2"), disease2 = c("d2", "d3"),
                            value = c(2, 3), measure = "OR",
                            dist = c("lognormal_ci", NA), p1 = c(2, NA), p2 = c(1.4, NA),
                            p3 = c(2.9, NA)),
  impacts = data.frame(disease = c("d1", "d2", "d3"), value = c(2.5, 5, 7.5),
                       units = "% of yield", dist = c("normal", NA, NA),
                       p1 = c(2.5, NA, NA), p2 = c(0.5, NA, NA))
)
typed
deconflate(typed, n_draws = 0)$adjusted
```

Here the pair d1:d3 has no row, so it is unknown and the global method is
used. The R constructors (`cm_diseases()`, `cm_associations()`,
`cm_population()`, `cm_impacts()`, `cm_model()`) build the same model
directly.

## Disease ids

Ids are used in keys such as `assoc:d1:d2` and in lists such as `"d1; d2"`,
so they must not contain `|`, `;` or `:`, nor leading or trailing spaces.
`"all"` is reserved (it means every other disease in `adjusted_for`):

```{r}
tryCatch(cm_diseases(c("LAM", "SCK:early"), c(0.25, 0.40)),
         deconflate_error = function(e) conditionMessage(e))
tryCatch(cm_diseases(c("LAM", "all"), c(0.25, 0.40)),
         deconflate_error = function(e) conditionMessage(e))
```

## Fuller examples

The folder `five_diseases` uses every feature on five diseases: all
probability types and association measures, a three-way term, crude and
adjusted impacts, interactions (`interactions_welfare.csv`, for the welfare
impacts), event impacts with mixed measures (`culling.csv`), and every
distribution type mixed with point values. Its `run_all_features.R` script
runs every part of the package on it.

The folder `global_dairy_2024` holds the inputs of Rasmussen et al. (2024)
for 12 diseases: `diseases.csv`, `associations.csv` (the 38 odds ratios of
Table 3), `impacts_yield.csv`, `impacts_fertility.csv` and `culling.csv`
(hazard ratios entered as `snapshot_crude`), with the input distributions of
the paper's analysis in the `dist` columns. Pairs not in Table 3 have no row,
so they are unknown:

```{r}
gd <- system.file("extdata", "global_dairy_2024", package = "deconflate")
cm_read_inputs(diseases = file.path(gd, "diseases.csv"),
               associations = file.path(gd, "associations.csv"),
               impacts = file.path(gd, "impacts_yield.csv"))
```

`example_global_dairy()` returns the same inputs as a model, and
`vignette("reproducing-published")` compares it with the published analysis,
which treated these pairs as independent.
