---
title: "Using your own data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using your own data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

Inputs can be kept in CSV files (one per table) or typed in R as data
frames. `cm_read_inputs()` checks every table, reports all problems at once,
and returns the population, one analysis per impact table, a hazard-ratio
model if there are hazard ratios, and a Monte Carlo sampler if there are
input distributions.

## Files

| File | Contents | Required columns |
|---|---|---|
| `diseases.csv` | one row per disease | `id`, `value` |
| `associations.csv` | one row per associated pair | `disease1`, `disease2` |
| `three_way.csv` | optional three-way association scenarios | `disease1`, `disease2`, `disease3`, `ratio` |
| `impacts_<analysis>.csv` | one impact table per analysis, e.g. `impacts_yield.csv` | `disease`, `value` |
| `interactions_<analysis>.csv` | optional impact interactions of that analysis | `disease1`, `disease2`, `value` |
| `uncertainty.csv` | optional input distributions shared by all analyses | `key`, `dist` |
| `uncertainty_<analysis>.csv` | optional input distributions of one analysis | `key`, `dist` |
| `hazard_ratios.csv` | optional culling (or mortality) hazard ratios | `disease`, `value` |

Only `diseases.csv` is required. A single `impacts.csv` is read as one
analysis named `impacts`. All analyses share the diseases, associations and
three-way terms. The optional columns are described in `?cm_read_inputs`:

* `diseases.csv`: `type` (`prevalence`, `probability` or `incidence_rate`),
  `time_horizon`, `reference_population`, `source`;
* `associations.csv`: `value`, `measure` (`OR`, `RR`, `RD`, `cond_prob`,
  `phi`, `table`, `independent` or `unknown`), the counts `n11`, `n10`, `n01`,
  `n00` for `measure = table`, `adjusted`, `adjusted_for`, `source`;
* `impacts_<analysis>.csv`: `estimand` (`crude` or `adjusted_linear`),
  `adjusted_for`, `source`, and one `label` and one `units` per table;
* `hazard_ratios.csv`: `estimand` (`crude` or `adjusted`), `adjusted_for`,
  `source`.

Every disease needs a row in every impact table (use 0 for no impact) and in
the hazard-ratio table (use 1 for no effect). Impact values can be in any
units, the same within a table; results come back in those units.

## Start from the template

`cm_template()` writes an example set of files to a folder:

```{r}
dir <- file.path(tempdir(), "my-inputs")
cm_template(dir, overwrite = TRUE)
list.files(dir)
```

It has two analyses, milk yield (in percent) and calving interval (in days):

```{r}
read.csv(file.path(dir, "impacts_yield.csv"))
read.csv(file.path(dir, "impacts_calving_interval.csv"))
```

## Read the files

```{r}
inp <- cm_read_inputs(dir = dir)
inp
```

The result holds:

* `population`: the diseases, associations and three-way terms, as a
  `cm_population()` object;
* `analyses`: a `cm_analyses()` object with one model per impact table;
* `model`: the `cm_model()` when there is exactly one analysis (otherwise
  `NULL`);
* `hr_model`: the hazard-ratio model, for `deconflate_hr()`;
* `sampler`: a `cm_sampler()` for one analysis, or a `cm_batch_sampler()`
  for several, for `cm_monte_carlo()`.

```{r}
inp$population
names(inp$analyses$models)
deconflate(inp$analyses)
deconflate_hr(inp$hr_model)
```

The template has two analyses, so the sampler is a batch sampler: every
draw of the shared disease and association inputs is used by both analyses.

```{r}
inp$sampler
mc <- cm_monte_carlo(inp$sampler, 200, seed = 1)
mc
s <- summary(mc, diagnose = FALSE)
s[, c("analysis", "disease", "mean", "q0.025", "q0.975")]
```

## When something is wrong

All problems are listed with their table, row (counting from the first row
below the header) and column, so that they can be fixed in one pass.
`cm_check_inputs()` returns them without stopping:

```{r}
cm_check_inputs(
  diseases = data.frame(id = c("LAM", "SCK", "SCK"), value = c(0.25, 1.4, 0.3)),
  associations = data.frame(disease1 = "LAM", disease2 = "MET", value = 2.0),
  impacts = data.frame(disease = c("LAM", "SCK"), value = c("4.81", "8,40"))
)
```

`cm_read_inputs()` stops with the same list (an error of class
`deconflate_input_problems`) when there are errors, and prints notes
otherwise. The installed example with errors shows more cases:

```{r}
cm_check_inputs(dir = system.file("extdata", "example_with_errors", package = "deconflate"))
```

## The global dairy example

The folder `global_dairy_2024` holds the inputs of Rasmussen et al. (2024):
`diseases.csv` (12 diseases), `associations.csv` (38 odds ratios),
`impacts_yield.csv` and `impacts_fertility.csv` (in percent),
`hazard_ratios.csv` (culling) and `uncertainty.csv` (the input
distributions).

```{r}
gd <- cm_read_inputs(dir = system.file("extdata", "global_dairy_2024", package = "deconflate"))
gd
cmp <- compare_methods(gd$analyses, methods = c("published", "simultaneous"))
cmp$impacts[cmp$impacts$analysis == "yield", ]
```

The simultaneous solution changes the sign of several adjusted impacts (see
`vignette("reproducing-published")`). The culling hazard ratios are adjusted
separately (`vignette("culling-hazard-ratios")`):

```{r}
deconflate_hr(gd$hr_model, method = "published")
```

## Tables typed in R

Any table can be a data frame instead of a file. Tables that come one per
analysis (`impacts`, `interactions`, `uncertainty`) are named lists, named
after the analyses; `uncertainty` can also have an element named `shared`
for the population inputs:

```{r}
inp2 <- cm_read_inputs(
  diseases = data.frame(id = c("d1", "d2", "d3"), value = c(0.10, 0.15, 0.20)),
  associations = data.frame(disease1 = c("d1", "d2"), disease2 = c("d2", "d3"),
                            value = c(2, 3), measure = "OR"),
  impacts = list(
    yield = data.frame(disease = c("d1", "d2", "d3"), value = c(2.5, 5, 7.5),
                       units = "% of yield"),
    fertility = data.frame(disease = c("d1", "d2", "d3"), value = c(1, 2, 0),
                           units = "% of calving interval")
  ),
  uncertainty = list(
    shared = data.frame(key = "assoc:d1:d2", dist = "lognormal_ci", p1 = 2, p2 = 1.4, p3 = 2.9),
    yield = data.frame(key = "impact:d1", dist = "normal", p1 = 2.5, p2 = 0.5)
  )
)
inp2
deconflate(inp2$analyses)$yield$adjusted
```

With a single analysis (e.g. `impacts = list(yield = ...)`, or one data
frame, which becomes the analysis `impacts`), `$model` is set and `$sampler`
is a single-analysis sampler.

## Associations from other designs

### Adjusted association measures

The 2x2 algebra needs marginal (crude) association measures. An odds ratio
from a multivariable model is conditional on its covariates, so it is
rejected unless you choose to use it as if it were marginal, which is an
approximation:

```{r}
d12 <- cm_diseases(c("d1", "d2"), c(0.10, 0.15))
a_adj <- cm_associations("d1", "d2", 1.8, adjusted = TRUE, adjusted_for = "parity; herd")
tryCatch(cm_population(d12, a_adj), deconflate_unsupported = function(e) conditionMessage(e))
cm_population(d12, a_adj, adjusted_associations = "use_as_marginal")
```

In files, set the column `adjusted` to `TRUE` and call
`cm_read_inputs(..., adjusted_associations = "use_as_marginal")`; each such
row is then reported as a note.

### Contingency tables and zero cells

A study table can be given as counts (`measure = "table"`). Its odds ratio is
applied to the modelled population's own probabilities. A table with a zero
cell has no finite odds ratio, so by default 0.5 is added to every cell
(Haldane-Anscombe correction), and the row is flagged in column `corrected`:

```{r}
tab <- cm_associations("d1", "d2", measure = "table", n11 = 0, n10 = 30, n01 = 40, n00 = 930)
tab[, c("disease1", "disease2", "measure", "value", "corrected")]
tryCatch(cm_associations("d1", "d2", measure = "table", n11 = 0, n10 = 30, n01 = 40,
                         n00 = 930, zero_cell = "error"),
         deconflate_error = function(e) conditionMessage(e))
```

When reading files, the correction is applied and reported as a note:

```{r}
cm_check_inputs(
  diseases = data.frame(id = c("d1", "d2"), value = c(0.10, 0.15)),
  associations = data.frame(disease1 = "d1", disease2 = "d2", measure = "table",
                            n11 = 0, n10 = 30, n01 = 40, n00 = 930)
)
```

Tables in which every count is zero are rejected.

## Disease ids

Ids are used in keys such as `assoc:d1:d2` and in lists such as `"d1; d2"`,
so they must not contain `|`, `;` or `:`, nor leading or trailing spaces.
`"all"` is reserved (it means every other disease in `adjusted_for`):

```{r}
tryCatch(cm_diseases(c("LAM", "SCK:early"), c(0.25, 0.40)),
         deconflate_error = function(e) conditionMessage(e))
tryCatch(cm_diseases(c("LAM", "all"), c(0.25, 0.40)),
         deconflate_error = function(e) conditionMessage(e))
```

## Uncertainty

Each row of an uncertainty table gives one input a distribution. The key
names the input:

* `prob:<disease>`: the disease's `value` (e.g. a prevalence or incidence
  rate);
* `assoc:<d1>:<d2>`: the association's `value` (e.g. an odds ratio);
* `impact:<analysis>:<disease>`: an impact, in the units of its table;
* `inter:<analysis>:<d1>:<d2>`: an interaction.

In `uncertainty_<analysis>.csv`, or when there is only one analysis, the
analysis can be left out (`impact:<disease>`, `inter:<d1>:<d2>`). Disease
and association keys belong in the shared `uncertainty.csv`.

`dist` and the parameters `p1`-`p4`:

| `dist` | `p1` | `p2` | `p3` | `p4` |
|---|---|---|---|---|
| `fixed` | value | | | |
| `normal` | mean | sd | lower bound (optional) | upper bound (optional) |
| `lognormal` | meanlog | sdlog | | |
| `lognormal_ci` | estimate | lower CI | upper CI | level (default 0.95) |
| `beta` | shape1 | shape2 | min (default 0) | max (default 1) |
| `pert` | min | mode | max | lambda (default 4) |
| `pert_mean` | min | mean | max | lambda (default 4) |
| `uniform` | min | max | | |

```{r}
read.csv(file.path(dir, "uncertainty.csv"))
```

`cm_dist_table()` builds the distributions alone, if you want to pass them
to `cm_sampler()` yourself. See `vignette("uncertainty")` for the Monte
Carlo analysis.
