---
title: "Using your own data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using your own data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

Inputs can be kept in CSV files (one per table) or typed in R as data frames.
`cm_read_inputs()` checks every table, reports all problems at once, and
returns a model (and a Monte Carlo sampler if you give an uncertainty table).

## Start from the template

`cm_template()` writes an example set of files to a folder:

```{r}
dir <- file.path(tempdir(), "my-inputs")
cm_template(dir, overwrite = TRUE)
list.files(dir)
```

The five files are:

| File | Contents | Required columns |
|---|---|---|
| `diseases.csv` | one row per disease | `id`, `value` |
| `associations.csv` | one row per associated pair | `disease1`, `disease2` |
| `impacts.csv` | one row per disease and outcome | `disease`, `outcome`, `value` |
| `interactions.csv` | optional impact interactions | `disease1`, `disease2`, `value`, `outcome` |
| `uncertainty.csv` | optional input distributions | `key`, `dist` |

Only `diseases.csv` is required. The optional columns (e.g. `type`, `measure`,
`scale`, `direction`, `units`, `adjusted_for`, `source`) are described in
`?cm_read_inputs`. Here is the template's impacts table:

```{r}
read.csv(file.path(dir, "impacts.csv"))
```

Yield impacts are percent decreases; culling impacts are hazard ratios
(`scale = hazard_ratio`), adjusted on the log scale (see
`vignette("culling-hazard-ratios")`). Every disease needs a row for every
outcome; use 0 for no impact.

Two more example sets are installed with the package:

* `global_dairy_2024`: the 12 diseases, 38 odds ratios, yield, fertility and
  culling impacts, and input distributions of Rasmussen et al. (2024), as
  used in the published analysis (culling as hazard ratios);
* `example_with_errors`: a small set with deliberate mistakes, to see how
  problems are reported.

```{r}
system.file("extdata", package = "deconflate")
```

## Read the files

```{r}
inp <- cm_read_inputs(dir = dir)
inp
```

`inp$model` is an ordinary model:

```{r}
res <- deconflate(inp$model, method = "global")
res$adjusted[, c("outcome", "disease", "raw", "adjusted")]
```

Compare the methods side by side:

```{r}
compare_methods(inp$model)
```

## When something is wrong

All problems are listed with their table, row (counting from the first row
below the header) and column, so they can be fixed in one pass.
`cm_check_inputs()` returns them without stopping:

```{r}
cm_check_inputs(
  diseases = data.frame(id = c("LAM", "SCK", "SCK"), value = c(0.25, 1.4, 0.3)),
  associations = data.frame(disease1 = "LAM", disease2 = "MET", value = 2.0),
  impacts = data.frame(disease = c("LAM", "SCK"), outcome = "yield",
                       value = c("4.81", "8,40"), scale = "percent")
)
```

`cm_read_inputs()` stops with the same list when there are errors. The
installed example with errors shows more cases:

```{r}
cm_check_inputs(dir = system.file("extdata", "example_with_errors", package = "deconflate"))
```

## The global dairy example

```{r}
gd <- cm_read_inputs(dir = system.file("extdata", "global_dairy_2024", package = "deconflate"))
gd
cmp <- compare_methods(gd$model, methods = c("published", "simultaneous"))
head(cmp$impacts, 12)
```

## Tables typed in R

Any table can be a data frame instead of a file:

```{r}
inp2 <- cm_read_inputs(
  diseases = data.frame(id = c("d1", "d2", "d3"), value = c(0.10, 0.15, 0.20)),
  associations = data.frame(disease1 = c("d1", "d2"), disease2 = c("d2", "d3"),
                            value = c(2, 3), measure = "OR"),
  impacts = data.frame(disease = c("d1", "d2", "d3"), outcome = "yield",
                       value = c(2.5, 5, 7.5), scale = "percent")
)
deconflate(inp2$model)$adjusted
```

## Uncertainty

Each row of `uncertainty.csv` gives one input a distribution. The key names
the input:

* `prob:<disease>`: the disease's `value` (e.g. a prevalence or incidence
  rate);
* `assoc:<d1>:<d2>`: the association's `value` (e.g. an odds ratio);
* `impact:<outcome>:<disease>`: the impact, on the scale it was entered
  (e.g. percent);
* `inter:<outcome>:<d1>:<d2>`: an interaction.

`dist` and the parameters `p1`-`p4`:

| `dist` | `p1` | `p2` | `p3` | `p4` |
|---|---|---|---|---|
| `fixed` | value | | | |
| `normal` | mean | sd | lower bound (optional) | upper bound (optional) |
| `lognormal` | meanlog | sdlog | | |
| `lognormal_ci` | estimate | lower CI | upper CI | level (default 0.95) |
| `beta` | shape1 | shape2 | min (default 0) | max (default 1) |
| `pert` | min | mode | max | lambda (default 4) |
| `pert_mean` | min | mean | max | lambda (default 4) |
| `uniform` | min | max | | |

```{r}
read.csv(file.path(dir, "uncertainty.csv"))
mc <- cm_monte_carlo(inp$sampler, 200, method = c("published", "simultaneous"), seed = 1)
compare_methods(mc)
```

`cm_dist_table()` builds the distributions alone, if you want to pass them
to `cm_sampler()` yourself.
