---
title: "Reproducing the published analyses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing the published analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

This vignette reproduces the calculations in Rasmussen et al. (2022, *Prev.
Vet. Med.* 203:105617) and Rasmussen et al. (2024, *J. Dairy Sci.*
107:6945-6970), and explains where the package's results differ from the
published tables.

## Supplementary File example (2022)

```{r}
res <- deconflate(example_supplement(), method = "published")
res$adjusted[, c("disease", "raw", "adjusted")]
productivity_gap(res, c(yield = 10000))
```

The published gap of 10,225 units (20, 61 and 143 units per disease) comes
from rounding the adjusted impacts to 0.02, 0.04 and 0.07 before computing
the gap. Without rounding, the published method gives 10,215.7 units.

## UK dairy example (2022, Tables 8-10)

`example_uk_dairy_2022()` contains the 13 diseases, the 19 odds ratios and the
yield, fertility and culling inputs of Tables 2-6. Culling hazard ratios are
converted to excess annual culling risk by treating them as odds ratios, as in
the paper (`culling_method = "or_approx"`).

```{r}
m <- example_uk_dairy_2022()
eco <- uk_dairy_2022_economics()
res <- deconflate(m, method = "published")
gaps <- productivity_gap(res, eco$observed)
gaps$summary
value_losses(gaps, eco$unit_value, eco$additional)$total
```

Adjusted culling hazard ratios (Table 8, eq. 23):

```{r}
adjusted_hr(res, attr(m, "culling_conversion"), method = "published")
```

**What reproduces.**

* Fertility matches Tables 9-10: a disease-free calving interval of 375.09
  days, valued at GBP 101.79 per cow per year.
* The adjusted culling hazard ratios match Table 8 to within 0.01.
* The culling gap is within 0.5% of the published value.

**What does not.** Yield cannot be reproduced exactly from the printed
Table 4.

* The tables imply different inputs. For subclinical ketosis, note i of
  Table 4 (340 kg per lactation) gives 340 / 8737 = 3.89%, not the printed
  3.05%. With 3.89% the adjusted values match Table 8 (SCK 2.66 vs 2.65,
  lameness 4.75 vs 4.76).
* Table 10's yield column implies further small differences in the inputs,
  for example for fasciolosis.

```{r}
m389 <- example_uk_dairy_2022(yield_sck = 100 * 340 / 8737)
y <- deconflate(m389, method = "published")$adjusted
y[y$outcome == "yield", c("disease", "raw", "adjusted")]
```

### Corrections the package makes available

* **Exact solution.** `method = "simultaneous"` solves the additive
  equations instead of approximating them.
* **Hazard ratios.** `culling_method = "proportional_hazards"` converts
  hazard ratios to risks under proportional hazards, instead of treating
  them as odds ratios. This gives larger excess risks.
* **Additive culling gap.** `culling_scale = "absolute"` subtracts the
  excess risk from the observed culling rate. The published approach
  instead treats excess risk as a proportional increase in the rate.

```{r}
m2 <- example_uk_dairy_2022(culling_method = "proportional_hazards",
                            culling_scale = "absolute")
eco2 <- uk_dairy_2022_economics("absolute")
res2 <- suppressWarnings(deconflate(m2))
summary(res2, economics = eco2)
```

## Global dairy analysis (2024, Table 5)

Table 5 reports the means of comorbidity-adjusted impacts over Monte Carlo
draws of the inputs in Tables 2-4. `sampler_global_dairy()` encodes those
distributions. In Tables 2-4, the central value of a PERT distribution is
its mode. Run with many more draws for stable estimates; 200 are used here
to keep the vignette fast.

```{r}
mc <- cm_monte_carlo(sampler_global_dairy(), 200, method = "published", seed = 2024)
s <- summary(mc)
y <- s[s$outcome == "yield", c("disease", "mean", "q0.025", "q0.975")]
y$mean_pct <- round(100 * y$mean, 2)
y$table5 <- c(CK = 0.03, CM = 1.36, DA = 1.18, DYS = 3.48, LAM = 2.62, MET = 2.87,
              MF = 0.07, OC = 2.59, PTB = 3.37, RP = 2.30, SCK = 7.11, SCM = 5.58)[y$disease]
y[, c("disease", "mean_pct", "table5")]
```

The yield means agree closely with Table 5. For fertility impacts whose
raw distributions extend below zero (e.g. metritis, subclinical ketosis and
subclinical mastitis), the published approximation is unstable, so their
means depend on how such draws are handled.

The simultaneous solution at the input means shows the inconsistency
directly. Several adjusted impacts change sign, while the total burden
stays similar:

```{r}
res24 <- suppressWarnings(deconflate(example_global_dairy()))
res24$diagnostics[, c("outcome", "n_sign_changes", "sign_changes")]
```
