---
title: "Reproducing the published analyses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing the published analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

This vignette reproduces the calculations in Rasmussen et al. (2022, *Prev.
Vet. Med.* 203:105617) and Rasmussen et al. (2024, *J. Dairy Sci.*
107:6945-6970), and explains where the package's results differ from the
published tables. `reproduce_rasmussen_2022()` and
`reproduce_rasmussen_2024()` use the published approximation (eq. 16 of the
2022 paper) and the hazard-ratio conversions and economic valuations of the
papers. None of these is offered for new analyses: the approximation is not
a method of `deconflate()` and appears elsewhere only in `compare_methods()`
and `cm_monte_carlo()`, for comparison (see `vignette("deconflate")` and
`vignette("culling-hazard-ratios")`).

## Supplementary File example (2022)

The impacts are entered in percent, so the adjusted impacts are in percent.
`compare_methods()` gives the published approximation beside the exact
solution; the gap is computed from the published result in base R, as in
`vignette("deconflate")`, for an observed level of 10,000 units:

```{r}
cmp_sup <- compare_methods(example_supplement(), methods = c("published", "simultaneous"))
cmp_sup$impacts
pub <- cmp_sup$results$published
disease_free <- 10000 / (1 - pub$totals$adjusted_total / 100)
gap <- disease_free - 10000
c(disease_free = disease_free, gap = gap)
setNames(gap * pub$contributions$share, pub$contributions$disease)
```

The published disease-free level of 10,225 units (gaps of 20, 61 and 143
units per disease) comes from rounding the adjusted impacts to 2%, 4% and 7%
before computing the gap. Without rounding, the published approximation
gives 10,215.7 units (21.1, 56.7 and 137.9 units per disease).

## UK dairy example (2022, Tables 8-10)

`example_uk_dairy_2022()` contains the 13 diseases and the 19 odds ratios of
Tables 2-3, and the yield and fertility inputs of Tables 4-5 as two analyses.
The culling hazard ratios of Table 6 are attached as the attribute
`"hazard_ratios"`, for the hazard-ratio adapter. The paper's culling
analysis, which converted the hazard ratios to excess annual culling risks
by treating them as odds ratios and recovered adjusted hazard ratios by
rescaling (eq. 23), exists only inside `reproduce_rasmussen_2022()`.
The observed means and unit values of Table 1, the veterinary expenditure
and the paper's culling valuation are used only inside the reproduction,
which computes the productivity gaps and their values as in the paper.
`reproduce_rasmussen_2022()` runs the whole calculation:

```{r}
r22 <- reproduce_rasmussen_2022()
r22
r22$total
```

### Comparison with Table 9

Using the inputs as printed in Tables 2-6:

| | Published (Table 9) | Package |
|---|---|---|
| Yield potential (kg/cow/year) | 9306.32 | 9299.21 |
| Yield gap value (GBP/cow/year) | 172.05 | 169.90 |
| Calving interval potential (days) | 375.09 | 375.09 |
| Fertility gap value (GBP/cow/year) | 101.79 | 101.78 |
| Culling rate potential (%) | 22.56 | 22.55 |
| Culling gap value (GBP/cow/year) | 59.27 | 59.48 |
| Total including veterinary costs (GBP/cow/year) | 404.21 | 402.25 |

**What reproduces.**

* Fertility matches Table 9 (375.09 days, GBP 101.79 vs 101.78 after
  rounding).
* The adjusted culling hazard ratios match Table 8 to within 0.01, and the
  culling gap is within 0.5% of the published value.

**What does not.**

* **Yield.** The adjusted yield impacts cannot be reproduced exactly from
  the printed Table 4. For subclinical ketosis, note i of Table 4 (340 kg per
  lactation) gives 340 / 8737 = 3.89%, not the printed 3.05%. With 3.89%,
  most adjusted values match Table 8 (SCK 2.66 vs 2.65, lameness 4.75 vs
  4.76), but displaced abomasum does not (2.28 vs 2.49). Table 10's yield
  column implies further small input differences, for example for
  fasciolosis (7.04 vs 7.33).
* **Allocation of the fertility gap.** Table 10 allocates GBP 15.09 to cystic
  ovary and GBP 5.80 to metritis. Allocating the gap in proportion to each
  disease's adjusted impact times its prevalence, as described in the paper
  (eq. 22), gives GBP 14.76 and GBP 6.04. The total is the same.

The original spreadsheets for the 2022 analysis are no longer available, so
the source of these differences cannot be checked. The published corrigendum
does not explain them: it corrects Table 7 only (the dystocia loss taken from
the literature, GBP 20.97 to GBP 31.92, and the directly aggregated total, now
GBP 660.26), which is not an input to the comorbidity adjustment.

Adjusted yield impacts (percent) with the printed value of 3.05% and with
3.89% for subclinical ketosis:

```{r}
r389 <- reproduce_rasmussen_2022(yield_sck = 100 * 340 / 8737)
data.frame(disease = r22$adjusted$disease,
           sck_3.05 = r22$adjusted$yield,
           sck_3.89 = r389$adjusted$yield)
```

### What the package offers instead

**Exact solution.** `method = "simultaneous"` solves the additive equations
instead of approximating them. The example has the yield and fertility
analyses. The gaps and values below use Table 1 in base R: milk yield of 8737
kg/cow/year (impacts are percent decreases) valued at GBP 0.3022/kg, and a
calving interval of 401 days (impacts are percent increases), each day
valued at 13 kg of milk:

```{r}
uk <- example_uk_dairy_2022()
uk
cmp <- compare_methods(uk, methods = c("published", "simultaneous"))
tt <- cmp$totals
yield <- tt$analysis == "yield"
L <- tt$adjusted_total / 100
tt$gap <- ifelse(yield, 8737 / (1 - L) - 8737, 401 - 401 / (1 + L))
tt$value <- tt$gap * ifelse(yield, 0.3022, 13 * 0.3022)
tt
tapply(tt$value, tt$method, sum)
```

The published rows reproduce the yield and fertility values of the
reproduction above (GBP 169.90 and 101.78 per cow per year). The
simultaneous method gives GBP 168.00 for yield and 95.43 for fertility. The
published total from `reproduce_rasmussen_2022()$total` (GBP 402.25) also
includes culling, valued with the paper's conversion, and veterinary costs.
That culling conversion is not available for new analyses, so no
simultaneous total including culling is given here.

**Hazard ratios.** For new analyses, the culling hazard ratios are adjusted
with the hazard-ratio adapter, and the culling attributable to disease is
computed without counting a cow twice. The hazard ratios of Table 6 are
attached to the example:

```{r}
hr_uk <- cm_hr_model(uk$population, attr(uk, "hazard_ratios"))
compare_methods(hr_uk, overall_risk = 0.27, joint = fit_joint(uk$population))
```

The attributable risk times the replacement price (GBP 1335.36) gives its
value per cow and year; as for the other analyses, that conversion is left
to the user.

## Global dairy analysis (2024, Table 5)

Table 5 reports the means of comorbidity-adjusted impacts over 50,000 Monte
Carlo draws per analysis. `sampler_global_dairy()` encodes the input
distributions. By default (`inputs = "analysis"`) it follows the published
analysis code rather than the printed tables:

* **Fixed disease probabilities.** Prevalence and incidence were not drawn
  in the adjustment. Each probability is `1 - exp(-incidence)` at the
  unrounded global mean, except paratuberculosis (a prevalence) and
  subclinical mastitis, which was entered without conversion (0.409 rather
  than 0.336).
* **PERT distributions** use the central value in Tables 2-4 as the mode.
* **Unrounded impact parameters**, e.g. a clinical ketosis yield impact of
  0.4322% rather than 0.43%.
* **Culling** was adjusted on the hazard ratio scale: HR - 1 was adjusted,
  and 1 was added back. Normal standard deviations were scaled by
  (HR - 1) / HR, and metritis used a PERT distribution (mode 1.12) rather
  than the normal distribution in Table 4 (mean 1.05). This analysis
  (`culling_hr_minus_1`) exists only inside `reproduce_rasmussen_2024()`,
  for reproduction; `example_global_dairy()` and `sampler_global_dairy()`
  have the yield and fertility analyses only.

`sampler_global_dairy(inputs = "tables")` uses Tables 2-4 as printed
instead. `reproduce_rasmussen_2024()` runs the published method on the
sampler, with the historical culling analysis added, and sets the means
beside Table 5 (yield and fertility in percent,
culling as hazard ratios). Run it with many more draws for stable estimates;
200 are used here to keep the vignette fast.

```{r}
r24 <- reproduce_rasmussen_2024(n_draws = 200)
cmp24 <- r24$comparison
cmp24[cmp24$analysis == "yield", c("disease", "table5", "mean", "median", "stability")]
cmp24[cmp24$analysis == "culling_hr", c("disease", "table5", "mean", "median", "stability")]
```

With 200,000 draws (`inst/validation/reference_2024_analysis.py`), the means
are:

| Disease | Yield, Table 5 | Yield, package | Culling HR, Table 5 | Culling HR, package |
|---|---|---|---|---|
| CK | 0.03 | 0.04 | 1.18 | 1.18 |
| CM | 1.36 | 1.36 | 1.90 | 1.90 |
| DA | 1.18 | 1.18 | 2.75 | 2.74 |
| DYS | 3.48 | 3.49 | 1.18 | 1.20 |
| LAM | 2.62 | 2.60 | 1.40 | 1.39 |
| MET | 2.87 | 2.86 | 1.03 | 1.03 |
| MF | 0.07 | 0.07 | 2.64 | 2.64 |
| OC | 2.59 | 2.51 | 1.51 | 1.48 |
| PTB | 3.37 | 3.32 | 2.07 | 2.06 |
| RP | 2.30 | 2.30 | 1.29 | 1.29 |
| SCK | 7.11 | 7.11 | 1.67 | 1.67 |
| SCM | 5.58 | 5.58 | 1.25 | 1.25 |

**Negative odds-ratio draws.** About 3% of draws of the lameness:ovarian
cyst odds ratio (normal, mean 2.63, SD 1.44) are negative, as are about 1% of
lameness:paratuberculosis draws. The analysis code used these draws as they
were, taking the absolute value of a negative discriminant in eq. 11. A
negative odds ratio has no valid 2x2 table, so the package truncates normal
odds-ratio distributions at zero. This explains the paratuberculosis yield
difference (3.38 without truncation). It does not explain ovarian cyst
(2.51 with or without truncation): in the analysis output, the spread of
the adjusted ovarian cyst yield impact is several times larger than for the
other diseases, which suggests that a few extreme draws drive its mean.

**Fertility.** Fertility means for clinical mastitis, dystocia, lameness,
milk fever, ovarian cyst and retained placenta agree with Table 5 to within
about 0.05 (for example clinical mastitis 6.10 vs 6.09). Clinical ketosis
(0.30 vs 0.34) and paratuberculosis (4.08 vs 4.23) differ more. Where the raw
impact distribution extends below zero (displaced abomasum, metritis,
subclinical ketosis and subclinical mastitis), the published approximation
`m^2 / (m + c)` is unstable: draws in which `m + c` is close to zero produce
extreme values. For subclinical ketosis the Monte Carlo mean does not settle
even with 200,000 draws, and for subclinical mastitis the archived inputs
give 0.52% rather than the 0.04% in Table 5. Medians, or the exact
(simultaneous) solution, are more robust summaries. The `stability` column
flags estimates whose means are unreliable (see `vignette("uncertainty")`):

```{r}
cmp24[cmp24$analysis == "fertility", c("disease", "table5", "mean", "median", "stability")]
```

The simultaneous solution at the central values shows the inconsistency
directly. Several adjusted impacts change sign, while the aggregates move
less (about 3% for yield and 18% for fertility):

```{r}
res24 <- deconflate(example_global_dairy(), warn = FALSE)
vapply(res24, function(r) r$diagnostics$sign_changes, character(1))
compare_methods(example_global_dairy(), methods = c("published", "simultaneous"))$totals
```

**Culling losses.** The adjusted culling impacts of the 2024 analysis are
adjusted HR - 1; adding 1 gives the adjusted hazard ratios. The central
values are in `r24$central$culling_hr_minus_1`, and they equal the
`"published"` column of `compare_methods()` for the hazard-ratio model:

```{r}
cu <- r24$central$culling_hr_minus_1$adjusted
hr_pub <- compare_methods(example_global_dairy_hr(), methods = "published")
data.frame(disease = cu$disease, hr = 1 + cu$raw, adjusted_hr = 1 + cu$adjusted,
           compare_methods = hr_pub$impacts$published)
```

For the losses, the paper converted each adjusted hazard ratio to an excess
culling risk relative to the overall culling rate `r`,
`HR * r / (HR * r + 1 - r) - r`, which treats the hazard ratio as an odds
ratio and counts cows with several diseases more than once. For new analyses
use `attributable_risk()` (see `vignette("culling-hazard-ratios")`).
