---
title: "Reproducing the published analyses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing the published analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

This vignette reproduces the calculations in Rasmussen et al. (2022, *Prev.
Vet. Med.* 203:105617) and Rasmussen et al. (2024, *J. Dairy Sci.*
107:6945-6970), and explains where the package's results differ from the
published tables. `reproduce_rasmussen_2022()` and
`reproduce_rasmussen_2024()` use the published approximation (eq. 16 of the
2022 paper) and the hazard-ratio conversions and economic valuations of the
papers. None of these is offered for new analyses: the approximation is not
a method of `deconflate()` and appears elsewhere only in `compare_methods()`,
for comparison (see `vignette("deconflate")` and `vignette("event-impacts")`).

The papers also treated every pair of diseases without a published odds
ratio as independent. The reproduction functions do the same. The package
default is different: a pair without an association is unknown, and
`deconflate()` fills it in with the global method (see
`vignette("interactions")`). To compare the published approximation with the
exact solution under the papers' assumption, the sections below give the
unlisted pairs an odds ratio of 1 with this helper:

```{r}
independent_pairs <- function(m) {
  pt <- pair_tables(m)
  u <- pt[pt$status == "unknown", , drop = FALSE]
  a <- rbind(m$associations, cm_associations(u$disease1, u$disease2, 1))
  cm_model(m$diseases, m$impacts, associations = a)
}
```

## Supplementary File example (2022)

The impacts are entered in percent, so the adjusted impacts are in percent.
Every pair has an odds ratio (d1 and d3 have an odds ratio of 1).
`compare_methods()` gives the published approximation beside the exact
solution; the gap is computed from the published result in base R, as in
`vignette("deconflate")`, for an observed level of 10,000 units:

```{r}
cmp_sup <- compare_methods(example_supplement(), methods = c("published", "simultaneous"))
cmp_sup$impacts
pub <- cmp_sup$results$published
disease_free <- 10000 / (1 - pub$totals$adjusted_total / 100)
gap <- disease_free - 10000
c(disease_free = disease_free, gap = gap)
setNames(gap * pub$contributions$share, pub$contributions$disease)
```

The published disease-free level of 10,225 units (gaps of 20, 61 and 143
units per disease) comes from rounding the adjusted impacts to 2%, 4% and 7%
before computing the gap. Without rounding, the published approximation
gives 10,215.7 units (21.1, 56.7 and 137.9 units per disease).

## UK dairy example (2022, Tables 8-10)

`example_uk_dairy_2022(outcome)` contains the 13 diseases and the 19 odds
ratios of Tables 2-3, and the raw impacts of one outcome: milk yield (Table
4), calving interval (`"fertility"`, Table 5) or culling (Table 6, hazard
ratios as event impacts). The paper's culling analysis, which converted the
hazard ratios to excess annual culling risks by treating them as odds
ratios and recovered adjusted hazard ratios by rescaling (eq. 23), exists
only inside `reproduce_rasmussen_2022()`. The observed means and unit values
of Table 1, the veterinary expenditure and the paper's culling valuation are
used only inside the reproduction, which computes the productivity gaps and
their values as in the paper. `reproduce_rasmussen_2022()` runs the whole
calculation:

```{r}
r22 <- reproduce_rasmussen_2022()
r22
r22$total
```

### Comparison with Table 9

Using the inputs as printed in Tables 2-6:

| | Published (Table 9) | Package |
|---|---|---|
| Yield potential (kg/cow/year) | 9306.32 | 9299.21 |
| Yield gap value (GBP/cow/year) | 172.05 | 169.90 |
| Calving interval potential (days) | 375.09 | 375.09 |
| Fertility gap value (GBP/cow/year) | 101.79 | 101.78 |
| Culling rate potential (%) | 22.56 | 22.55 |
| Culling gap value (GBP/cow/year) | 59.27 | 59.48 |
| Total including veterinary costs (GBP/cow/year) | 404.21 | 402.25 |

**What reproduces.**

* Fertility matches Table 9 (375.09 days, GBP 101.79 vs 101.78 after
  rounding).
* The adjusted culling hazard ratios match Table 8 to within 0.01, and the
  culling gap is within 0.5% of the published value.

**What does not.**

* **Yield.** The adjusted yield impacts cannot be reproduced exactly from
  the printed Table 4. For subclinical ketosis, note i of Table 4 (340 kg per
  lactation) gives 340 / 8737 = 3.89%, not the printed 3.05%. With 3.89%,
  most adjusted values match Table 8 (SCK 2.66 vs 2.65, lameness 4.75 vs
  4.76), but displaced abomasum does not (2.28 vs 2.49). Table 10's yield
  column implies further small input differences, for example for
  fasciolosis (7.04 vs 7.33).
* **Allocation of the fertility gap.** Table 10 allocates GBP 15.09 to cystic
  ovary and GBP 5.80 to metritis. Allocating the gap in proportion to each
  disease's adjusted impact times its prevalence, as described in the paper
  (eq. 22), gives GBP 14.76 and GBP 6.04. The total is the same.

The original spreadsheets for the 2022 analysis are no longer available, so
the source of these differences cannot be checked. The published corrigendum
does not explain them: it corrects Table 7 only (the dystocia loss taken from
the literature, GBP 20.97 to GBP 31.92, and the directly aggregated total, now
GBP 660.26), which is not an input to the comorbidity adjustment.

Adjusted yield impacts (percent) with the printed value of 3.05% and with
3.89% for subclinical ketosis:

```{r}
r389 <- reproduce_rasmussen_2022(yield_sck = 100 * 340 / 8737)
data.frame(disease = r22$adjusted$disease,
           sck_3.05 = r22$adjusted$yield,
           sck_3.89 = r389$adjusted$yield)
```

### What the package offers instead

**Exact solution.** The simultaneous method solves the additive equations
instead of approximating them. With the unlisted pairs independent, as in
the paper, every pair has an odds ratio and both methods can run. The gaps
and values below use Table 1 in base R: milk yield of 8737 kg/cow/year
(impacts are percent decreases) valued at GBP 0.3022/kg, and a calving
interval of 401 days (impacts are percent increases), each day valued at 13
kg of milk:

```{r}
uk_yield <- independent_pairs(example_uk_dairy_2022("yield"))
uk_fert <- independent_pairs(example_uk_dairy_2022("fertility"))
uk_yield
tt_y <- compare_methods(uk_yield, methods = c("published", "simultaneous"))$totals
tt_f <- compare_methods(uk_fert, methods = c("published", "simultaneous"))$totals
tt_y$gap <- 8737 / (1 - tt_y$adjusted_total / 100) - 8737
tt_y$value <- tt_y$gap * 0.3022
tt_f$gap <- 401 - 401 / (1 + tt_f$adjusted_total / 100)
tt_f$value <- tt_f$gap * 13 * 0.3022
rbind(cbind(outcome = "yield", tt_y), cbind(outcome = "fertility", tt_f))
```

The published rows reproduce the yield and fertility values of the
reproduction above (GBP 169.90 and 101.78 per cow per year). The
simultaneous method gives GBP 168.00 for yield and 95.43 for fertility. The
published total from `reproduce_rasmussen_2022()$total` (GBP 402.25) also
includes culling, valued with the paper's conversion, and veterinary costs.
That culling conversion is not available for new analyses, so no
simultaneous total including culling is given here.

**Culling as event impacts.** For new analyses, the culling hazard ratios
are event impacts, adjusted with the snapshot hazard model
(`deconflate(..., event_model = TRUE)`), and the culling attributable to
disease is computed without counting a cow twice. `compare_methods()` sets
the 2024 approach (published) beside the snapshot model, again with the
unlisted pairs independent and the paper's annual culling rate of 27%:

```{r}
uk_cull <- independent_pairs(example_uk_dairy_2022("culling"))
compare_methods(uk_cull, methods = c("published", "snapshot"),
                event_model = TRUE, overall_risk = 0.27)
```

The attributable risk times the replacement price (GBP 1335.36) gives its
value per cow and year; as for the other outcomes, that conversion is left
to the user.

## Global dairy analysis (2024, Table 5)

Table 5 reports the means of comorbidity-adjusted impacts over 50,000 Monte
Carlo draws per analysis. `example_global_dairy(outcome)` holds the inputs
and their distributions, one outcome at a time (`"yield"`, `"fertility"` or
`"culling"`). By default (`inputs = "analysis"`) it follows the published
analysis code rather than the printed tables:

* **Fixed disease probabilities.** Prevalence and incidence were not drawn
  in the adjustment. Each probability is `1 - exp(-incidence)` at the
  unrounded global mean, except paratuberculosis (a prevalence) and
  subclinical mastitis, which was entered without conversion (0.409 rather
  than 0.336).
* **PERT distributions** use the central value in Tables 2-4 as the mode.
* **Unrounded impact parameters**, e.g. a clinical ketosis yield impact of
  0.4322% rather than 0.43%.
* **Culling** was adjusted on the hazard ratio scale: HR - 1 was adjusted,
  and 1 was added back. Normal standard deviations were scaled by
  (HR - 1) / HR, and metritis used a PERT distribution (mode 1.12) rather
  than the normal distribution in Table 4 (mean 1.05). This analysis
  (`culling_hr_minus_1`) exists only inside `reproduce_rasmussen_2024()`,
  for reproduction; `example_global_dairy("culling")` gives the hazard
  ratios as event impacts instead.

`example_global_dairy(inputs = "tables")` uses Tables 2-4 as printed
instead. `reproduce_rasmussen_2024()` draws these inputs (with the unlisted
pairs independent, and the same draws of the disease and association inputs
for every outcome), adds the historical culling analysis, runs the published
method on every draw and sets the means beside Table 5 (yield and fertility
in percent, culling as hazard ratios). Run it with many more draws for
stable estimates; 200 are used here to keep the vignette fast.

```{r}
r24 <- reproduce_rasmussen_2024(n_draws = 200)
cmp24 <- r24$comparison
cmp24[cmp24$analysis == "yield", c("disease", "table5", "mean", "median", "stability")]
cmp24[cmp24$analysis == "culling_hr", c("disease", "table5", "mean", "median", "stability")]
```

With 200,000 draws (`inst/validation/reference_2024_analysis.py`), the means
are:

| Disease | Yield, Table 5 | Yield, package | Culling HR, Table 5 | Culling HR, package |
|---|---|---|---|---|
| CK | 0.03 | 0.04 | 1.18 | 1.18 |
| CM | 1.36 | 1.36 | 1.90 | 1.90 |
| DA | 1.18 | 1.18 | 2.75 | 2.74 |
| DYS | 3.48 | 3.49 | 1.18 | 1.20 |
| LAM | 2.62 | 2.60 | 1.40 | 1.39 |
| MET | 2.87 | 2.86 | 1.03 | 1.03 |
| MF | 0.07 | 0.07 | 2.64 | 2.64 |
| OC | 2.59 | 2.51 | 1.51 | 1.48 |
| PTB | 3.37 | 3.32 | 2.07 | 2.06 |
| RP | 2.30 | 2.30 | 1.29 | 1.29 |
| SCK | 7.11 | 7.11 | 1.67 | 1.67 |
| SCM | 5.58 | 5.58 | 1.25 | 1.25 |

**Negative odds-ratio draws.** About 3% of draws of the lameness:ovarian
cyst odds ratio (normal, mean 2.63, SD 1.44) are negative, as are about 1% of
lameness:paratuberculosis draws. The analysis code used these draws as they
were, taking the absolute value of a negative discriminant in eq. 11. A
negative odds ratio has no valid 2x2 table, so the package truncates normal
odds-ratio distributions at zero. This explains the paratuberculosis yield
difference (3.38 without truncation). It does not explain ovarian cyst
(2.51 with or without truncation): in the analysis output, the spread of
the adjusted ovarian cyst yield impact is several times larger than for the
other diseases, which suggests that a few extreme draws drive its mean.

**Fertility.** Fertility means for clinical mastitis, dystocia, lameness,
milk fever, ovarian cyst and retained placenta agree with Table 5 to within
about 0.05 (for example clinical mastitis 6.10 vs 6.09). Clinical ketosis
(0.30 vs 0.34) and paratuberculosis (4.08 vs 4.23) differ more. Where the raw
impact distribution extends below zero (displaced abomasum, metritis,
subclinical ketosis and subclinical mastitis), the published approximation
`m^2 / (m + c)` is unstable: draws in which `m + c` is close to zero produce
extreme values. For subclinical ketosis the Monte Carlo mean does not settle
even with 200,000 draws, and for subclinical mastitis the archived inputs
give 0.52% rather than the 0.04% in Table 5. Medians, or the exact
(simultaneous) solution, are more robust summaries. The `stability` column
flags estimates whose means are unreliable (see `vignette("uncertainty")`):

```{r}
cmp24[cmp24$analysis == "fertility", c("disease", "table5", "mean", "median", "stability")]
```

The simultaneous solution at the central values, with the unlisted pairs
independent as in the paper, shows the inconsistency directly. Several
adjusted impacts change sign, while the aggregates move less (about 3% for
yield and 18% for fertility):

```{r}
gd_yield <- independent_pairs(example_global_dairy("yield"))
gd_fert <- independent_pairs(example_global_dairy("fertility"))
cmp_y <- compare_methods(gd_yield, methods = c("published", "simultaneous"))
cmp_f <- compare_methods(gd_fert, methods = c("published", "simultaneous"))
cmp_y$diagnostics[, c("method", "sign_changes")]
cmp_f$diagnostics[, c("method", "sign_changes")]
data.frame(method = cmp_y$totals$method, yield = cmp_y$totals$adjusted_total,
           fertility = cmp_f$totals$adjusted_total)
```

**The package default: unknown pairs.** `example_global_dairy()` itself
leaves the 28 pairs that are not in Table 3 unknown, as for any input in this
package, so `deconflate()` uses the global method: the maximum-entropy fit
gives each unknown pair the association implied by the associations both
diseases have with other diseases, instead of an odds ratio of 1. For the
yield impacts at the central values (no draws):

```{r}
gd_default <- deconflate(example_global_dairy("yield"), n_draws = 0, warn = FALSE)
gd_default$notes
range(gd_default$unknown_pairs$fitted_or)
c(independent = cmp_y$totals$adjusted_total[cmp_y$totals$method == "simultaneous"],
  unknown = gd_default$totals$adjusted_total)
gd_default$diagnostics$sign_changes
```

The fit gives the 28 unknown pairs odds ratios between 1.08 and 2.37, all
above 1, so animals with one disease are expected to have more of the
others than under independence. More of each raw impact is then attributed
to associated diseases, and the aggregate yield loss falls from 7.49% (unknown
pairs independent, simultaneous method) to 6.92% (unknown pairs filled in,
global method). The same four diseases
change sign (CK, CM, DA, MF). Neither assumption is identified by the
published odds ratios; `screen_associations()` shows which unknown pairs
matter most (`vignette("uncertainty")`).

**Culling losses.** The adjusted culling impacts of the 2024 analysis are
adjusted HR - 1; adding 1 gives the adjusted hazard ratios. The central
values are in `r24$central$culling_hr_minus_1`, and they equal the
`"published"` column of `compare_methods()` for the culling hazard ratios as
event impacts (with the unlisted pairs independent):

```{r}
cu <- r24$central$culling_hr_minus_1$adjusted
gd_cull <- independent_pairs(example_global_dairy("culling"))
hr_pub <- compare_methods(gd_cull, methods = "published", event_model = TRUE,
                          overall_risk = 0.2366)
data.frame(disease = cu$disease, hr = 1 + cu$raw, adjusted_hr = 1 + cu$adjusted,
           compare_methods = hr_pub$impacts$published)
```

For the losses, the paper converted each adjusted hazard ratio to an excess
culling risk relative to the overall culling rate `r`,
`HR * r / (HR * r + 1 - r) - r`, which treats the hazard ratio as an odds
ratio and counts cows with several diseases more than once. For new analyses
use `deconflate(..., event_model = TRUE, overall_risk = )`, whose
`$attributable` gives the culling risk attributable to disease and its
allocation to diseases (see `vignette("event-impacts")`).
