---
title: "Reproducing the published analyses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing the published analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

This vignette reproduces the calculations in Rasmussen et al. (2022, *Prev.
Vet. Med.* 203:105617) and Rasmussen et al. (2024, *J. Dairy Sci.*
107:6945-6970), and explains where the package's results differ from the
published tables.

## Supplementary File example (2022)

```{r}
res <- deconflate(example_supplement(), method = "published")
res$adjusted[, c("disease", "raw", "adjusted")]
productivity_gap(res, c(yield = 10000))
```

The published gap of 10,225 units (20, 61 and 143 units per disease) comes
from rounding the adjusted impacts to 0.02, 0.04 and 0.07 before computing
the gap. Without rounding, the published method gives 10,215.7 units.

## UK dairy example (2022, Tables 8-10)

`example_uk_dairy_2022()` contains the 13 diseases, the 19 odds ratios and the
yield, fertility and culling inputs of Tables 2-6. Culling hazard ratios are
converted to excess annual culling risk by treating them as odds ratios, as in
the paper (`culling_method = "or_approx"`).

```{r}
m <- example_uk_dairy_2022()
eco <- uk_dairy_2022_economics()
res <- deconflate(m, method = "published")
gaps <- productivity_gap(res, eco$observed)
gaps$summary
value_losses(gaps, eco$unit_value, eco$additional)$total
```

Adjusted culling hazard ratios (Table 8, eq. 23):

```{r}
adjusted_hr(res, attr(m, "culling_conversion"), method = "published")
```

### Comparison with Table 9

Using the inputs as printed in Tables 2-6:

| | Published (Table 9) | Package |
|---|---|---|
| Yield potential (kg/cow/year) | 9306.32 | 9299.21 |
| Yield gap value (GBP/cow/year) | 172.05 | 169.90 |
| Calving interval potential (days) | 375.09 | 375.09 |
| Fertility gap value (GBP/cow/year) | 101.79 | 101.78 |
| Culling rate potential (%) | 22.56 | 22.55 |
| Culling gap value (GBP/cow/year) | 59.27 | 59.48 |
| Total including veterinary costs (GBP/cow/year) | 404.21 | 402.25 |

**What reproduces.**

* Fertility matches Table 9 (375.09 days, GBP 101.79 vs 101.78 after
  rounding).
* The adjusted culling hazard ratios match Table 8 to within 0.01, and the
  culling gap is within 0.5% of the published value.

**What does not.**

* **Yield.** The adjusted yield impacts cannot be reproduced exactly from
  the printed Table 4. For subclinical ketosis, note i of Table 4 (340 kg per
  lactation) gives 340 / 8737 = 3.89%, not the printed 3.05%. With 3.89%,
  most adjusted values match Table 8 (SCK 2.66 vs 2.65, lameness 4.75 vs
  4.76), but displaced abomasum does not (2.28 vs 2.49). Table 10's yield
  column implies further small input differences, for example for
  fasciolosis (7.04 vs 7.33).
* **Allocation of the fertility gap.** Table 10 allocates GBP 15.09 to cystic
  ovary and GBP 5.80 to metritis. Allocating the gap in proportion to each
  disease's adjusted impact times its prevalence, as described in the paper
  (eq. 22), gives GBP 14.76 and GBP 6.04. The total is the same.

The original spreadsheets for the 2022 analysis are no longer available, so
the source of these differences cannot be checked. The published corrigendum
does not explain them: it corrects Table 7 only (the dystocia loss taken from
the literature, GBP 20.97 to GBP 31.92, and the directly aggregated total, now
GBP 660.26), which is not an input to the comorbidity adjustment.

```{r}
m389 <- example_uk_dairy_2022(yield_sck = 100 * 340 / 8737)
y <- deconflate(m389, method = "published")$adjusted
y[y$outcome == "yield", c("disease", "raw", "adjusted")]
```

### Corrections the package makes available

* **Exact solution.** `method = "simultaneous"` solves the additive
  equations instead of approximating them.
* **Hazard ratios.** `culling_method = "proportional_hazards"` converts
  hazard ratios to risks under proportional hazards, instead of treating
  them as odds ratios. This gives larger excess risks.
* **Additive culling gap.** `culling_scale = "absolute"` subtracts the
  excess risk from the observed culling rate. The published approach
  instead treats excess risk as a proportional increase in the rate.

```{r}
m2 <- example_uk_dairy_2022(culling_method = "proportional_hazards",
                            culling_scale = "absolute")
eco2 <- uk_dairy_2022_economics("absolute")
res2 <- suppressWarnings(deconflate(m2))
summary(res2, economics = eco2)
```

## Global dairy analysis (2024, Table 5)

Table 5 reports the means of comorbidity-adjusted impacts over 50,000 Monte
Carlo draws per outcome. `sampler_global_dairy()` encodes the input
distributions. By default (`inputs = "analysis"`) it follows the published
analysis code rather than the printed tables:

* **Fixed disease probabilities.** Prevalence and incidence were not drawn
  in the adjustment. Each probability is `1 - exp(-incidence)` at the
  unrounded global mean, except paratuberculosis (a prevalence) and
  subclinical mastitis, which was entered without conversion (0.409 rather
  than 0.336).
* **PERT distributions** use the central value in Tables 2-4 as the mode.
* **Unrounded impact parameters**, e.g. a clinical ketosis yield impact of
  0.4322% rather than 0.43%.
* **Culling** was adjusted on the hazard ratio scale: HR - 1 was adjusted,
  and 1 was added back. Normal standard deviations were scaled by
  (HR - 1) / HR, and metritis used a PERT distribution (mode 1.12) rather
  than the normal distribution in Table 4 (mean 1.05).

`sampler_global_dairy(inputs = "tables")` uses Tables 2-4 as printed
instead. Run with many more draws for stable estimates; 200 are used here to
keep the vignette fast.

```{r}
mc <- cm_monte_carlo(sampler_global_dairy(), 200, method = "published", seed = 2024)
s <- summary(mc)
s$package <- round(ifelse(s$outcome == "culling", 1 + s$mean, 100 * s$mean), 2)
table5 <- list(
  yield = c(0.03, 1.36, 1.18, 3.48, 2.62, 2.87, 0.07, 2.59, 3.37, 2.30, 7.11, 5.58),
  culling = c(1.18, 1.90, 2.75, 1.18, 1.40, 1.03, 2.64, 1.51, 2.07, 1.29, 1.67, 1.25)
)
ids <- c("CK", "CM", "DA", "DYS", "LAM", "MET", "MF", "OC", "PTB", "RP", "SCK", "SCM")
pkg <- function(o) s$package[s$outcome == o][match(ids, s$disease[s$outcome == o])]
data.frame(
  disease = ids,
  yield_package = pkg("yield"), yield_table5 = table5$yield,
  culling_package = pkg("culling"), culling_table5 = table5$culling
)
```

With 200,000 draws (`inst/validation/reference_2024_analysis.py`), the means
are:

| Disease | Yield, Table 5 | Yield, package | Culling HR, Table 5 | Culling HR, package |
|---|---|---|---|---|
| CK | 0.03 | 0.04 | 1.18 | 1.18 |
| CM | 1.36 | 1.36 | 1.90 | 1.90 |
| DA | 1.18 | 1.18 | 2.75 | 2.74 |
| DYS | 3.48 | 3.49 | 1.18 | 1.20 |
| LAM | 2.62 | 2.60 | 1.40 | 1.39 |
| MET | 2.87 | 2.86 | 1.03 | 1.03 |
| MF | 0.07 | 0.07 | 2.64 | 2.64 |
| OC | 2.59 | 2.51 | 1.51 | 1.48 |
| PTB | 3.37 | 3.32 | 2.07 | 2.06 |
| RP | 2.30 | 2.30 | 1.29 | 1.29 |
| SCK | 7.11 | 7.11 | 1.67 | 1.67 |
| SCM | 5.58 | 5.58 | 1.25 | 1.25 |

**Negative odds-ratio draws.** About 3% of draws of the lameness:ovarian
cyst odds ratio (normal, mean 2.63, SD 1.44) are negative, as are about 1% of
lameness:paratuberculosis draws. The analysis code used these draws as they
were, taking the absolute value of a negative discriminant in eq. 11. A
negative odds ratio has no valid 2x2 table, so the package truncates normal
odds-ratio distributions at zero. This explains the paratuberculosis yield
difference (3.38 without truncation). It does not explain ovarian cyst
(2.51 with or without truncation): in the analysis output, the spread of
the adjusted ovarian cyst yield impact is several times larger than for the
other diseases, which suggests that a few extreme draws drive its mean.

**Fertility.** Fertility means for clinical mastitis, dystocia, lameness,
milk fever, ovarian cyst and retained placenta agree with Table 5 to within
about 0.05 (for example clinical mastitis 6.10 vs 6.09). Clinical ketosis
(0.30 vs 0.34) and paratuberculosis (4.08 vs 4.23) differ more. Where the raw
impact distribution extends below zero (displaced abomasum, metritis,
subclinical ketosis and subclinical mastitis), the published approximation
`m^2 / (m + c)` is unstable: draws in which `m + c` is close to zero produce
extreme values. For subclinical ketosis the Monte Carlo mean does not settle
even with 200,000 draws, and for subclinical mastitis the archived inputs
give 0.52% rather than the 0.04% in Table 5. Medians, or the exact
(simultaneous) solution, are more robust summaries.

The simultaneous solution at the central values shows the inconsistency
directly. Several adjusted impacts change sign, while the total burden stays
similar:

```{r}
res24 <- suppressWarnings(deconflate(example_global_dairy()))
res24$diagnostics[, c("outcome", "n_sign_changes", "sign_changes")]
```

Adjusted culling impacts convert back to hazard ratios with
`adjusted_hr(..., method = "excess_hr")`. For the losses, the paper converted
each adjusted hazard ratio to an excess culling risk relative to the overall
culling rate `r`, `HR * r / (HR * r + 1 - r) - r`; this is
`hr_to_risk(hr, NA, r, method = "overall_odds")`.

```{r}
pub24 <- deconflate(example_global_dairy(), method = "published")
adjusted_hr(pub24, method = "excess_hr")
```
