---
title: "Threshold searches and many diseases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Threshold searches and many diseases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

This vignette covers two extensions: threshold searches and screens, which
show how much the conclusions depend on single inputs (including inputs
that have not been estimated), and the sampled backend of the global model,
for more diseases than can be enumerated.

## Threshold searches

`cm_threshold()` varies one input over a range, with every other input fixed
at the model's values, and finds the values at which a conclusion changes.
There are four kinds of conclusion:

* `"rank"`: two diseases swap places, ranked by their contribution to the
  aggregate (`by = "adjusted"` ranks by adjusted impact instead);
* `"sign"`: a disease's adjusted impact crosses zero (for event impacts,
  its adjusted hazard ratio crosses 1). This is a sign change implied by the
  model and its inputs; it is not by itself evidence of a protective effect;
* `"total"`: the aggregate crosses `target`, in the units of the impacts;
* `"change"`: the aggregate departs from its value at the model's own input
  by the relative amount `target` (e.g. `-0.1` for a 10% reduction). This
  answers questions such as "what association strength would reduce the
  total by 10%?", which is useful for pairs without an association.

For event impacts (`event_model = TRUE`), the aggregate is the risk
attributable to disease and the contributions are its Shapley allocation
(`vignette("event-impacts")`).

The input is keyed as in the `distributions` of `cm_model()`:
`"assoc:<d1>:<d2>"`, `"inter:<d1>:<d2>"`, `"prob:<disease>"`,
`"impact:<disease>"`, `"three:<d1>:<d2>:<d3>"` for a three-way ratio, or
`"risk"` for the overall risk of event impacts. A pair without an
association (unknown) is varied as an odds ratio.

Each point is adjusted as `deconflate()` would adjust it, with point
estimates. With `method = "auto"` (the default), the method is chosen for
the point's own inputs: the simultaneous method when every pair has an
association and there are no interactions or three-way terms, the global
method otherwise. When the input is an impact, an interaction or the
overall risk (which do not change the joint distribution) and the
adjustment uses the joint distribution, it is fitted once and reused, or a
`joint` from `fit_joint()` can be passed; other inputs change the joint
distribution, so it is refitted at every point.

### How the search works

The response need not be monotonic, so the range is first scanned on a grid
of `n_grid` points (log-spaced for odds ratios, risk ratios and three-way
ratios). Each grid
point is adjusted and classified as usable, or by the reason it is not
(`"infeasible"`, `"singular"`, `"undefined"` for non-finite results,
`"unresolved"`, `"unsupported"` or `"error"`). Each pair of neighbouring
usable points on which the conclusion differs brackets a crossing, which is
refined by bisection. As a result:

* every crossing in the range is reported, not only the first;
* stretches of unusable grid points are listed in `regions`, so that "no
  crossing found" can be told apart from "part of the range could not be
  evaluated";
* a refined crossing counts as a threshold only if the compared quantity is
  close to zero on both sides of the final bracket. A jump across a pole,
  such as a nearly singular system, is reported as a `"discontinuity"`,
  never as a threshold;
* a crossing whose bisection lands on an unusable point is `"unresolved"`.

Thresholds are deterministic: they refer to the model's input values, one
input at a time. Statements such as "the probability that d1 ranks first"
need the draws of `deconflate()` (`n_draws`; `vignette("uncertainty")`);
probabilistic thresholds are not implemented.

### A relative change in the aggregate

In the Supplementary File example, d1 and d3 are independent (odds ratio 1),
and the adjusted aggregate is about 2.109 percent:

```{r}
m <- example_supplement()
deconflate(m, n_draws = 0)$totals
th <- cm_threshold(m, "assoc:d1:d3", c(1, 50), conclusion = "change", target = -0.1)
th
```

An odds ratio of about 5.26 between d1 and d3 would reduce the aggregate by
10%, to about 1.898. The adjusted results at the two ends of the final
bracket are returned:

```{r}
th$details[[1]]$below
th$details[[1]]$above
```

The search works in either direction. The odds ratio of d2 and d3 is 3 in
the example; a weaker association, about 1.43, would increase the aggregate
by 10%, because less of the raw impacts is then attributed to the associated
disease:

```{r}
cm_threshold(m, "assoc:d2:d3", c(0.2, 20), conclusion = "change", target = 0.1)$thresholds
```

For a pair without an association, the baseline is the global method, which
fills the pair in from the maximum-entropy fit; the change is relative to
that fit. Here the pair d1:d3 is left out. The fitted distribution gives it
an odds ratio of about 1.14, and an odds ratio of about 6.4 would reduce the
aggregate by 10%:

```{r}
m_unk <- cm_model(
  cm_diseases(c("d1", "d2", "d3"), c(0.10, 0.15, 0.20)),
  cm_impacts(c("d1", "d2", "d3"), c(2.5, 5, 7.5), units = "%"),
  associations = cm_associations(c("d1", "d2"), c("d2", "d3"), c(2, 3))
)
deconflate(m_unk, n_draws = 0)$unknown_pairs
th_unk <- cm_threshold(m_unk, "assoc:d1:d3", c(1, 50), conclusion = "change", target = -0.1)
th_unk$thresholds[, c("item", "status", "threshold", "description")]
```

The baseline value of an unknown input is `NA` (`th_unk$baseline`). At the
grid points the pair has a value, so every pair has an association and the
simultaneous method is used there; it gives the same answer as the global
method would.

### Rankings

As the association of d2 and d3 strengthens, more of d2's raw impact is
attributed to d3, and d2's contribution falls. Above an odds ratio of about
10.20, d1 contributes more than d2. Disease d3 contributes most over the
whole range:

```{r}
th_rank <- cm_threshold(m, "assoc:d2:d3", c(0.1, 100), conclusion = "rank", n_grid = 201)
th_rank
th_rank$summary
```

`plot()` shows the compared quantities over the scan (here the differences
in contribution of each pair of diseases), the baseline value of the input
(dotted) and the thresholds (dashed):

```{r, fig.width = 6, fig.height = 4}
plot(th_rank)
```

### Signs

The adjusted impact of d1 is negative when its raw impact is below about
0.374 percent:

```{r}
th_sign <- cm_threshold(m, "impact:d1", c(-5, 5), conclusion = "sign", diseases = "d1")
th_sign$thresholds[, c("item", "status", "threshold", "description")]
```

Below that value, the raw impact of d1 is smaller than its associated
diseases alone would produce under the additive model. This says that the
inputs are inconsistent with the model there (or the estimands are not what
they were assumed to be), not that d1 is protective.

`cm_threshold()` searches with the methods of `deconflate()`. The published
approximation, `m^2 / (m + c)`, kept for comparison only, behaves
differently: it has a pole where its denominator is zero, and its adjusted
impact changes sign there by jumping through infinity. For d1, `c` is the
part of the raw impact that the conflation matrix `A` attributes to the
other diseases, so the pole is at a raw impact of d1 of about -0.527:

```{r}
A <- deconflate(m, n_draws = 0)$conflation$A
-sum(A["d1", c("d2", "d3")] * c(5, 7.5))
```

`compare_methods()` shows this on either side of the pole:

```{r}
sapply(c(-0.60, -0.45), function(v) {
  mv <- cm_model(m, cm_impacts(c("d1", "d2", "d3"), c(v, 5, 7.5)))
  unlist(compare_methods(mv, methods = c("published", "simultaneous"))$impacts[1, -1])
})
```

A search across such a pole reports a `"discontinuity"`, not a threshold.

### A target in the units of the impacts

What raw impact of d2 would make the aggregate 3 percent (about 13.04)?

```{r}
th_total <- cm_threshold(m, "impact:d2", c(0, 20), conclusion = "total", target = 3)
th_total$thresholds[, c("item", "status", "threshold", "description")]
```

### Ranges that cannot be fully evaluated

A probability of 1 or more is impossible, so the grid points from just above
1 are listed as an infeasible region. Below 1, d1 moves above d2 at a
prevalence of about 0.224 and above d3 at about 0.622; d2 and d3 do not swap
where the range could be evaluated:

```{r}
th_prob <- cm_threshold(m, "prob:d1", c(0.05, 1.5), conclusion = "rank")
th_prob
th_prob$summary
```

### Event impacts

With `event_model = TRUE` and `overall_risk`, the search uses the snapshot
model of event impacts. The joint distribution is refitted at every point
when the input is an association, so these searches take longer. With
culling hazard ratios on the same population, the adjusted hazard ratio of
d3 falls below 1 when the odds ratio of d2 and d3 exceeds about 8.09:

```{r}
cull <- cm_model(m, cm_impacts(c("d1", "d2", "d3"), c(1.5, 2.0, 1.3), measure = "HR",
                               estimand = "snapshot_crude", label = "culling"))
th_ev <- cm_threshold(cull, "assoc:d2:d3", c(1, 100), conclusion = "sign",
                      event_model = TRUE, overall_risk = 0.25)
th_ev$thresholds[, c("item", "status", "threshold", "description")]
```

The overall risk itself can be searched with the input `"risk"`.

## Screens

The screens re-run the adjustment over a set of scenarios, one input at a
time, and report how much the aggregate and the ranking of diseases
change. Like `cm_threshold()`, they report point estimates, adjust each
scenario as `deconflate()` would (`method = "auto"`), and take
`event_model = TRUE` with `overall_risk`:

* `screen_associations()` multiplies each given association by each of
  `multipliers`, and sets each pair without an association to each of
  `or_values`;
* `screen_interactions()` adds an interaction of each size in `values` to
  each pair (additive impacts only; always the global method, with the joint
  distribution fitted once);
* `screen_three_way()` sets a three-way term for each triple (always the
  global method);
* `sensitivity_oat()` varies every input by `variation` (20% by default).

```{r}
screen_associations(m, or_values = c(0.5, 3))
sensitivity_oat(m)
```

A three-way term keeps the pairwise tables of its triple fixed, so for
additive impacts without interactions it changes nothing when every pair of
the triple has an association; with an unknown pair it can:

```{r}
screen_three_way(m, ratios = c(0.5, 2))[, c("pair", "scenario", "total", "rel_change")]
screen_three_way(m_unk, ratios = c(0.5, 2))[, c("pair", "scenario", "total", "rel_change")]
```

### Without association estimates

`deconflate()` needs at least one association; without any, there is
nothing to de-conflate:

```{r}
m0 <- m
m0$associations <- NULL
tryCatch(deconflate(m0), deconflate_error = function(e) conditionMessage(e))
```

The screens and `cm_threshold()` work without associations. The baseline
then has every pair unknown, and the maximum-entropy fit with no pairwise
constraints makes the diseases independent, so the baseline aggregate is
the raw one. `screen_associations()` shows which pairs would matter if they
were associated, and `cm_threshold()` how strong an association would have
to be to change the aggregate by a given amount:

```{r}
screen_associations(m0, or_values = c(0.5, 2, 4))
th0 <- cm_threshold(m0, "assoc:d2:d3", c(1, 50), conclusion = "change", target = -0.1)
th0$thresholds[, c("item", "status", "threshold", "description")]
```

An odds ratio of about 2.24 between d2 and d3 (with the other pairs
unknown) would reduce the aggregate by 10%. Such results show which
associations are worth estimating; they are not estimates themselves. Tables
read with `cm_read_inputs()` can leave out `associations` for this purpose.

## Many diseases

### What limits the number of diseases

The simultaneous method (and the published approximation in
`compare_methods()`) uses the pairwise 2x2 tables only. It never enumerates
combinations of diseases, so it works for any number of diseases (the
default triple screen checks n(n-1)(n-2)/6 triples; the exact LP
feasibility check is limited to 14 diseases).

The global method needs the joint distribution of disease combinations,
for interactions, unknown pairs and three-way terms, and so does the
snapshot model of event impacts. The exact backend of `fit_joint()`
enumerates all 2^n combinations, which limits it to about 20 diseases
(`max_diseases`). For more, the sampled backend fits the same
maximum-entropy model without enumeration. `deconflate()` uses it
automatically when a model with more than 20 diseases needs the joint
distribution, and says so in a note; `fit_joint(backend = "sampled")` runs
it directly.

### The sampled backend

The sampled backend works in three steps:

1. **Calibration.** The model has a main effect for each disease and an
   interaction parameter for each constrained pair (three-way terms are
   fixed at the log of their ratios). Many parallel Gibbs chains
   (`n_chains`) estimate the disease probabilities and pairwise tables
   implied by the current parameters, and the parameters are updated by the
   differences between the target and estimated logits and log odds ratios,
   until the estimates match the targets within Monte Carlo error. The
   interaction parameters are conditional log odds ratios, given the other
   diseases. They are calibrated, not set to the marginal log odds ratios of
   the inputs, which are only their starting values: with several
   associated diseases, the two differ.
2. **Sampling.** `n_samples` combinations are drawn from the calibrated
   model.
3. **Raking** (`calibrate = TRUE`, the default). The weights of the sampled
   combinations are adjusted by iterative proportional fitting so that the
   disease probabilities and pairwise tables match the targets exactly on the
   sample. The remaining Monte Carlo error affects only the higher-order
   structure (triples and beyond), and the odds ratios of unknown pairs.

The result has the same form as an exact fit (the distinct sampled
combinations and their weights), so the global method, interaction offsets,
Shapley allocation and the snapshot model of event impacts use it
unchanged. Its `diagnostics` report:

* the constraint residual of the raw sample and after raking;
* R-hat and effective sample sizes of the fitted probabilities and pairwise
  tables, from the spread between chains;
* Monte Carlo standard errors of these moments.

A sampled fit is an approximation with Monte Carlo error. Check it against
the exact backend on smaller systems where that is possible, and compare
fits with different seeds.

### Three diseases: sampled and exact

```{r}
je <- fit_joint(m)
js <- fit_joint(m, backend = "sampled", n_samples = 20000, n_chains = 500, seed = 11)
js
js$diagnostics$summary
js$diagnostics$moments
```

The probabilities of the eight combinations, exact and sampled:

```{r}
cmb <- merge(combination_probs(je), combination_probs(js),
             by = c("d1", "d2", "d3", "n_diseases"), suffixes = c("_exact", "_sampled"))
cmb[order(-cmb$prob_exact), ]
```

A second seed gives a different sample; the differences between seeds show
the size of the Monte Carlo error. The distribution of the number of
diseases per animal:

```{r}
js2 <- fit_joint(m, backend = "sampled", n_samples = 20000, n_chains = 500, seed = 12)
by_n <- function(j) tapply(j$prob, rowSums(j$cells), sum)
rbind(exact = by_n(je), seed_11 = by_n(js), seed_12 = by_n(js2))
```

With an interaction, the global method depends on probabilities of disease
triples, which the sampled fit approximates:

```{r}
mi <- set_interaction(m, "d1", "d2", 1)
rbind(exact = deconflate(mi, n_draws = 0, joint = je)$adjusted$adjusted,
      sampled = deconflate(mi, n_draws = 0, joint = js)$adjusted$adjusted)
```

A fitted joint can be passed to `deconflate()`, `compare_methods()` and
`cm_threshold()` with `joint`. Alternatively, the fitting arguments can be
passed through `...`, e.g. `deconflate(m, method = "global", backend =
"sampled", n_samples = 20000)`.

### Unresolved is not infeasible

If the calibration does not converge, or the raking does not match the
targets, the fit is reported as unresolved: `converged` is `FALSE`, a
warning of class `deconflate_nonconvergence` is given, and `deconflate()`
refuses to use the fit. This does not show that the inputs are infeasible:
more samples, chains or iterations may resolve it. `check_feasibility()`
tests feasibility directly. Here the inputs are infeasible, which the triple
screen shows:

```{r}
bad <- cm_population(cm_diseases(c("a", "b", "c"), c(0.5, 0.5, 0.5)),
                     cm_associations(c("a", "a", "b"), c("b", "c", "c"), c(20, 20, 0.05)))
jb <- fit_joint(bad, backend = "sampled", n_samples = 2000, n_chains = 100,
                fit_iter = 20, seed = 1)
jb$converged
check_feasibility(bad, method = "triples")
```

### Twenty-four diseases

A synthetic population of 24 diseases with prevalences from 0.05 to 0.3, an
odds ratio of 2 between consecutive diseases (x01 with x02, x02 with x03,
and so on; the other 253 pairs are unknown) and raw impacts from 1 to 5
percent:

```{r}
ids <- sprintf("x%02d", 1:24)
pop24 <- cm_population(
  cm_diseases(ids, seq(0.05, 0.3, length.out = 24)),
  cm_associations(ids[-24], ids[-1], 2)
)
m24 <- cm_model(pop24, cm_impacts(ids, seq(1, 5, length.out = 24), units = "%"))
```

The exact backend would need 2^24 (about 16.8 million) combinations and
refuses:

```{r}
tryCatch(fit_joint(pop24), deconflate_error = function(e) conditionMessage(e))
```

The unknown pairs need the global method, and with 24 diseases
`deconflate()` fits the joint distribution with the sampled backend (with
the default `n_samples` and `n_chains` of `fit_joint()`). The `seed`
argument of `deconflate()` is for its draws; to make the sampled fit
reproducible, set the random seed before the call, or fit it with
`fit_joint(..., seed = )` and pass it with `joint`:

```{r}
set.seed(3)
r24 <- deconflate(m24, n_draws = 0)
r24$notes
r24$joint
r24$totals
```

The odds ratios the fit gives the unknown pairs come from the sample, so
they carry Monte Carlo error (see the raking step above), and so do the
adjusted impacts that depend on them:

```{r}
summary(r24$unknown_pairs$fitted_or)
```

An interaction (1 percentage point when x12 and x13 occur together) does
not change the joint distribution, so the fit is reused:

```{r}
m24i <- set_interaction(m24, "x12", "x13", 1)
r_int <- deconflate(m24i, n_draws = 0, joint = r24$joint)
near <- ids[10:15]
data.frame(disease = near,
           raw = r24$adjusted$raw[match(near, ids)],
           no_interaction = r24$adjusted$adjusted[match(near, ids)],
           interaction = r_int$adjusted$adjusted[match(near, ids)])
r_int$totals
```

The adjusted impacts of x12 and x13 fall, because part of their raw impacts
is now attributed to the interaction; the other diseases change little.
These values carry Monte Carlo error from the sampled triples. In an
analysis, refit the joint distribution with another seed (or more samples)
and check that the conclusions do not change.

The run time of a sampled fit grows with `n_samples` and `n_chains`, and
that of the raking step with the number of constrained pairs (the pairs
with an association) times the number of distinct sampled combinations.
Draws (`n_draws`) refit the joint distribution for every draw, which is
slow with the sampled backend; check the run time on a few draws first.
