---
title: "Threshold searches and many diseases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Threshold searches and many diseases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

This vignette covers two extensions: threshold searches, which find the
input values at which a conclusion changes, and the sampled backend of the
global model, for more diseases than can be enumerated.

## Threshold searches

`cm_threshold()` varies one input over a range, with every other input fixed
at the model's values, and finds the values at which a conclusion changes.
There are four kinds of conclusion:

* `"rank"`: two diseases swap places, ranked by their contribution to the
  aggregate (`by = "adjusted"` ranks by adjusted impact instead);
* `"sign"`: a disease's adjusted impact crosses zero. This is a sign change
  implied by the model and its inputs; it is not by itself evidence of a
  protective effect;
* `"total"`: the aggregate crosses `target`, in the units of the impacts;
* `"change"`: the aggregate departs from its value at the model's own input
  by the relative amount `target` (e.g. `-0.1` for a 10% reduction). This
  answers questions such as "what association strength would reduce the
  total by 10%?", which is useful for pairs that are unknown or not listed.

The input is keyed as in `cm_sampler()`: `"assoc:<d1>:<d2>"`,
`"inter:<d1>:<d2>"`, `"prob:<disease>"`, `"impact:<disease>"`, or
`"three:<d1>:<d2>:<d3>"` for a three-way ratio. A pair without a numeric
measure (independent, unknown or not listed) is varied as an odds ratio.
By default, models with interactions or three-way terms, and interaction or
three-way inputs, use the global method; otherwise the method is
`"simultaneous"` (set `method` to override). Interactions and interaction or
three-way inputs need the global method. With the global method, when the
input is an impact or an interaction (which do not change the joint
distribution), the joint distribution is fitted once and reused, or a
`joint` from `fit_joint()` can be passed; other inputs change the joint
distribution, so it is refitted at every point.

### How the search works

The response need not be monotonic, so the range is first scanned on a grid
of `n_grid` points (log-spaced for odds ratios, risk ratios and three-way
ratios). Each grid
point is adjusted and classified as usable, or by the reason it is not
(`"infeasible"`, `"singular"`, `"undefined"` for non-finite results,
`"unresolved"`, `"unsupported"` or `"error"`). Each pair of neighbouring
usable points on which the conclusion differs brackets a crossing, which is
refined by bisection. As a result:

* every crossing in the range is reported, not only the first;
* stretches of unusable grid points are listed in `regions`, so that "no
  crossing found" can be told apart from "part of the range could not be
  evaluated";
* a refined crossing counts as a threshold only if the compared quantity is
  close to zero on both sides of the final bracket. A jump across a pole,
  such as a nearly singular system, is reported as a `"discontinuity"`,
  never as a threshold;
* a crossing whose bisection lands on an unusable point is `"unresolved"`.

Thresholds are deterministic: they refer to the model's input values, one
input at a time. Statements such as "the probability that d1 ranks first"
need the Monte Carlo tools (`vignette("uncertainty")`); probabilistic
thresholds are not implemented.

### A relative change in the aggregate

In the Supplementary File example, d1 and d3 are independent (odds ratio 1),
and the adjusted aggregate is about 2.109 percent:

```{r}
m <- example_supplement()
deconflate(m)$totals
th <- cm_threshold(m, "assoc:d1:d3", c(1, 50), conclusion = "change", target = -0.1)
th
```

An odds ratio of about 5.26 between d1 and d3 would reduce the aggregate by
10%, to about 1.898. The adjusted results at the two ends of the final
bracket are returned:

```{r}
th$details[[1]]$below
th$details[[1]]$above
```

The search works in either direction. The odds ratio of d2 and d3 is 3 in
the example; a weaker association, about 1.43, would increase the aggregate
by 10%, because less of the raw impacts is then attributed to the associated
disease:

```{r}
cm_threshold(m, "assoc:d2:d3", c(0.2, 20), conclusion = "change", target = 0.1)$thresholds
```

For an unknown pair, the simultaneous method cannot compute the baseline, but
the global method can: it leaves the pair unconstrained in the
maximum-entropy fit. The change is then relative to that fit. Here the
fitted distribution implies an odds ratio of about 1.14 for d1 and d3, and
an odds ratio of about 6.4 would reduce the aggregate by 10%:

```{r}
m_unk <- cm_model(
  cm_diseases(c("d1", "d2", "d3"), c(0.10, 0.15, 0.20)),
  cm_impacts(c("d1", "d2", "d3"), c(2.5, 5, 7.5), units = "%"),
  associations = cm_associations(c("d1", "d2"), c("d2", "d3"), c(2, 3)),
  missing_associations = "unknown"
)
th_unk <- cm_threshold(m_unk, "assoc:d1:d3", c(1, 50), conclusion = "change",
                       target = -0.1, method = "global")
th_unk$thresholds[, c("item", "status", "threshold", "description")]
```

The baseline value of an unknown input is `NA` (`th_unk$baseline`).

### Rankings

As the association of d2 and d3 strengthens, more of d2's raw impact is
attributed to d3, and d2's contribution falls. Above an odds ratio of about
10.20, d1 contributes more than d2. Disease d3 contributes most over the
whole range:

```{r}
th_rank <- cm_threshold(m, "assoc:d2:d3", c(0.1, 100), conclusion = "rank", n_grid = 201)
th_rank
th_rank$summary
```

`plot()` shows the compared quantities over the scan (here the differences
in contribution of each pair of diseases), the baseline value of the input
(dotted) and the thresholds (dashed):

```{r, fig.width = 6, fig.height = 4}
plot(th_rank)
```

### Signs

The adjusted impact of d1 is negative when its raw impact is below about
0.374 percent:

```{r}
th_sign <- cm_threshold(m, "impact:d1", c(-5, 5), conclusion = "sign", diseases = "d1")
th_sign$thresholds[, c("item", "status", "threshold", "description")]
```

Below that value, the raw impact of d1 is smaller than its associated
diseases alone would produce under the additive model. This says that the
inputs are inconsistent with the model there (or the estimands are not what
they were assumed to be), not that d1 is protective.

`cm_threshold()` searches with the methods of `deconflate()`
(`"simultaneous"` or `"global"`). The published approximation,
`m^2 / (m + c)`, kept for comparison only, behaves differently: it has a
pole where its denominator is zero, here at a raw impact of d1 of about
-0.527, and its adjusted impact changes sign there by jumping through
infinity. `compare_methods()` shows this on either side of the pole:

```{r}
pop_s <- cm_population(
  cm_diseases(c("d1", "d2", "d3"), c(0.10, 0.15, 0.20)),
  cm_associations(c("d1", "d2"), c("d2", "d3"), c(2, 3), measure = "OR")
)
sapply(c(-0.60, -0.45), function(v) {
  mv <- cm_model(pop_s, cm_impacts(c("d1", "d2", "d3"), c(v, 5, 7.5)))
  unlist(compare_methods(mv, methods = c("published", "simultaneous"))$impacts[1, -1])
})
```

A search across such a pole reports a `"discontinuity"`, not a threshold.

### A target in the units of the impacts

What raw impact of d2 would make the aggregate 3 percent (about 13.04)?

```{r}
th_total <- cm_threshold(m, "impact:d2", c(0, 20), conclusion = "total", target = 3)
th_total$thresholds[, c("item", "status", "threshold", "description")]
```

### Ranges that cannot be fully evaluated

A probability of 1 or more is impossible, so the grid points from just above
1 are listed as an infeasible region. Below 1, d1 moves above d2 at a
prevalence of about 0.224 and above d3 at about 0.622; d2 and d3 do not swap
where the range could be evaluated:

```{r}
th_prob <- cm_threshold(m, "prob:d1", c(0.05, 1.5), conclusion = "rank")
th_prob
th_prob$summary
```

## Many diseases

### What limits the number of diseases

The `"simultaneous"` method (and the published approximation in
`compare_methods()`) uses the pairwise 2x2 tables only. It never enumerates
combinations of diseases, so it works for any
number of diseases (the default triple screen checks n(n-1)(n-2)/6 triples;
the exact LP feasibility check is limited to 14 diseases).

The `"global"` method needs the joint distribution of disease combinations,
for interactions, unknown pairs and three-way terms, and so do the snapshot
hazard-ratio model and `attributable_risk()`. The exact backend of
`fit_joint()` enumerates all 2^n combinations, which limits it to about 20
diseases (`max_diseases`). For more, `fit_joint(backend = "sampled")` fits
the same maximum-entropy model without enumeration.

### The sampled backend

The sampled backend works in three steps:

1. **Calibration.** The model has a main effect for each disease and an
   interaction parameter for each constrained pair (three-way terms are
   fixed at the log of their ratios). Many parallel Gibbs chains
   (`n_chains`) estimate the disease probabilities and pairwise tables
   implied by the current parameters, and the parameters are updated by the
   differences between the target and estimated logits and log odds ratios,
   until the estimates match the targets within Monte Carlo error. The
   interaction parameters are conditional log odds ratios, given the other
   diseases. They are calibrated, not set to the marginal log odds ratios of
   the inputs, which are only their starting values: with several
   associated diseases, the two differ.
2. **Sampling.** `n_samples` combinations are drawn from the calibrated
   model.
3. **Raking** (`calibrate = TRUE`, the default). The weights of the sampled
   combinations are adjusted by iterative proportional fitting so that the
   disease probabilities and pairwise tables match the targets exactly on the
   sample. The remaining Monte Carlo error affects only the higher-order
   structure (triples and beyond).

The result has the same form as an exact fit (the distinct sampled
combinations and their weights), so the global method, interaction offsets,
Shapley allocation, the snapshot hazard model and `attributable_risk()` use
it unchanged. Its `diagnostics` report:

* the constraint residual of the raw sample and after raking;
* R-hat and effective sample sizes of the fitted probabilities and pairwise
  tables, from the spread between chains;
* Monte Carlo standard errors of these moments.

A sampled fit is an approximation with Monte Carlo error. Check it against
the exact backend on smaller systems where that is possible, and compare
fits with different seeds.

### Three diseases: sampled and exact

```{r}
je <- fit_joint(m)
js <- fit_joint(m, backend = "sampled", n_samples = 20000, n_chains = 500, seed = 11)
js
js$diagnostics$summary
js$diagnostics$moments
```

The probabilities of the eight combinations, exact and sampled:

```{r}
cmb <- merge(combination_probs(je), combination_probs(js),
             by = c("d1", "d2", "d3", "n_diseases"), suffixes = c("_exact", "_sampled"))
cmb[order(-cmb$prob_exact), ]
```

A second seed gives a different sample; the differences between seeds show
the size of the Monte Carlo error. The distribution of the number of
diseases per animal:

```{r}
js2 <- fit_joint(m, backend = "sampled", n_samples = 20000, n_chains = 500, seed = 12)
by_n <- function(j) tapply(j$prob, rowSums(j$cells), sum)
rbind(exact = by_n(je), seed_11 = by_n(js), seed_12 = by_n(js2))
```

With an interaction, the global method depends on probabilities of disease
triples, which the sampled fit approximates:

```{r}
mi <- set_interaction(m, "d1", "d2", 1)
rbind(exact = deconflate(mi, method = "global", joint = je)$adjusted$adjusted,
      sampled = deconflate(mi, method = "global", joint = js)$adjusted$adjusted)
```

A fitted joint can be passed to `deconflate()`, `deconflate_hr()`,
`compare_methods()` and `attributable_risk()` with `joint`. Alternatively,
the fitting arguments can be passed through `...`, e.g.
`deconflate(m, method = "global", backend = "sampled", n_samples = 20000,
seed = 1)`.

### Unresolved is not infeasible

If the calibration does not converge, or the raking does not match the
targets, the fit is reported as unresolved: `converged` is `FALSE`, a
warning of class `deconflate_nonconvergence` is given, and `deconflate()`
refuses to use the fit. This does not show that the inputs are infeasible:
more samples, chains or iterations may resolve it. `check_feasibility()`
tests feasibility directly. Here the inputs are infeasible, which the triple
screen shows:

```{r}
bad <- cm_population(cm_diseases(c("a", "b", "c"), c(0.5, 0.5, 0.5)),
                     cm_associations(c("a", "a", "b"), c("b", "c", "c"), c(20, 20, 0.05)))
jb <- fit_joint(bad, backend = "sampled", n_samples = 2000, n_chains = 100,
                fit_iter = 20, seed = 1)
jb$converged
check_feasibility(bad, method = "triples")
```

### Twenty-four diseases

A synthetic population of 24 diseases with prevalences from 0.05 to 0.3, an
odds ratio of 2 between consecutive diseases (x01 with x02, x02 with x03,
and so on; all other pairs independent) and raw impacts from 1 to 5 percent:

```{r}
ids <- sprintf("x%02d", 1:24)
pop24 <- cm_population(
  cm_diseases(ids, seq(0.05, 0.3, length.out = 24)),
  cm_associations(ids[-24], ids[-1], 2)
)
m24 <- cm_model(pop24, cm_impacts(ids, seq(1, 5, length.out = 24), units = "%"))
```

The exact backend would need 2^24 (about 16.8 million) combinations and
refuses:

```{r}
tryCatch(fit_joint(pop24), deconflate_error = function(e) conditionMessage(e))
```

The sampled fit:

```{r}
j24 <- fit_joint(pop24, backend = "sampled", n_samples = 20000, n_chains = 500, seed = 3)
j24
j24$diagnostics$summary
by_n(j24)
```

Without interactions, the global method uses the pairwise tables only.
Raking matches every pairwise table exactly on the sample (including the
independent pairs, which are constrained to an odds ratio of 1), so the
global result equals the simultaneous one up to the raking tolerance:

```{r}
r_sim <- deconflate(m24)
r_glob <- deconflate(m24, method = "global", joint = j24)
max(abs(r_glob$adjusted$adjusted - r_sim$adjusted$adjusted))
c(simultaneous = r_sim$totals$adjusted_total, global = r_glob$totals$adjusted_total)
```

An interaction (1 percentage point when x12 and x13 occur together) needs
the probabilities of disease triples, so it uses the sampled distribution
beyond its pairwise tables:

```{r}
m24i <- set_interaction(m24, "x12", "x13", 1)
r_int <- deconflate(m24i, method = "global", joint = j24)
near <- ids[10:15]
data.frame(disease = near,
           raw = r_glob$adjusted$raw[match(near, ids)],
           no_interaction = r_glob$adjusted$adjusted[match(near, ids)],
           interaction = r_int$adjusted$adjusted[match(near, ids)])
r_int$totals
```

The adjusted impacts of x12 and x13 fall, because part of their raw impacts
is now attributed to the interaction; the other diseases change little.
These values carry Monte Carlo error from the sampled triples. In an
analysis, refit the joint distribution with another seed (or more samples)
and check that the conclusions do not change.

The run time of a sampled fit grows with `n_samples` and `n_chains`, and
that of the raking step with the number of constrained pairs (every pair
that is specified or independent by default) times the number of distinct
sampled combinations.
