---
title: "Uncertainty, scenarios and sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uncertainty, scenarios and sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

## Input distributions

`dist_*()` functions describe uncertain inputs. Draws are on the scale the
inputs were entered on (e.g. percent impacts, odds ratios, incidence rates).

```{r}
dist_pert(1.19, 3.30, 10.71)        # min, mode, max
dist_lognormal_ci(2.7, 1.5, 4.9)    # odds ratio and 95% CI
dist_normal(2.63, 1.44, lower = 0)  # truncated at zero
```

## Monte Carlo analysis

`cm_sampler()` attaches distributions to a model. Each draw is a complete
model, so disease probabilities and associations are shared by all
outcomes within a draw. `outcome_correlation` additionally correlates a
disease's impacts across outcomes (Gaussian copula):

```{r}
m <- example_uk_dairy_2022()
R <- matrix(c(1, 0.5, 0.5, 1), 2,
            dimnames = list(c("yield", "fertility"), c("yield", "fertility")))
s <- cm_sampler(
  m,
  diseases = list(LAM = dist_beta(30, 70), MAS = dist_beta(30, 70)),
  associations = list("LAM:PTB" = dist_mixture(dist_lognormal_ci(2.7, 1.5, 4.9),
                                               dist_lognormal(log(2.7), 0.8),
                                               weights = c(0.8, 0.2)),
                      "LAM:SCK" = dist_lognormal_ci(2.01, 1.62, 2.49)),
  impacts = list("yield:LAM" = dist_normal(5.54, 0.8),
                 "fertility:LAM" = dist_normal(12.47, 2)),
  outcome_correlation = R
)
s
mc <- cm_monte_carlo(s, 200, method = "simultaneous",
                     economics = uk_dairy_2022_economics(), seed = 1)
mc
summary(mc, what = "total")
```

Draws with impossible inputs (e.g. a probability outside (0, 1), or an
association incompatible with the sampled probabilities) are rejected and
counted. Report the rejection rate: conditioning on feasibility changes the
effective input distribution.

## Comparing methods on the same draws

Give several methods to `cm_monte_carlo()` and every accepted draw is
adjusted with each of them. `compare_methods()` then tabulates the results:

```{r}
mc2 <- cm_monte_carlo(s, 200, method = c("published", "simultaneous"), seed = 1)
compare_methods(mc2)
```

## Unstable estimates

Monte Carlo means can be unstable. `summary()` checks every estimate and,
when one looks unstable, prints why and what to do. Here the raw yield
impact of `d2` is uncertain enough to be near zero or negative:

```{r}
s_d2 <- cm_sampler(example_supplement(),
                   impacts = list("yield:d2" = dist_normal(0.5, 1.5)))
mc_d2 <- cm_monte_carlo(s_d2, 400, method = c("published", "simultaneous"), seed = 2)
sm <- summary(mc_d2)
sm[, c("disease", "method", "mean", "q0.5", "trimmed_mean", "stability")]
```

There are three kinds of instability (see `?cm_diagnose`):

* **`no_mean`**: the published approximation, `m^2 / (m + c)`, divides by
  a quantity that changes sign across draws. Its distribution then has no
  mean, so no number of draws, and no sampling scheme, makes the mean
  converge. Report the median or trimmed mean, or use the exact method.
* **`heavy_tail`**: a few draws dominate the variance. Importance sampling
  can help (below).
* **`imprecise`**: the Monte Carlo error is large relative to the mean. Use
  more draws, or Latin hypercube sampling.

`cm_diagnose()` lists the flagged estimates with the suggestions:

```{r}
cm_diagnose(mc_d2)
```

### Latin hypercube sampling

`sampling = "lhs"` stratifies each input's distribution, which usually
reduces the Monte Carlo error of means:

```{r}
mc_lhs <- cm_monte_carlo(s_d2, 400, method = "simultaneous", sampling = "lhs", seed = 2)
summary(mc_lhs, diagnose = FALSE)[, c("disease", "mean", "mcse")]
```

### Importance sampling

When extreme values come from one region of one input,
`cm_suggest_proposal()` builds a defensive mixture that samples that region
more often. The draws are then weighted by the ratio of the densities, so
the estimates still refer to the original input distributions:

```{r}
prop <- cm_suggest_proposal(mc_d2, "yield", "d2", method = "simultaneous")
mc_is <- cm_monte_carlo(s_d2, 400, method = "simultaneous", proposal = prop, seed = 3)
mc_is
summary(mc_is, diagnose = FALSE)[, c("disease", "mean", "mcse")]
```

Importance sampling reduces the error of a mean that exists. It cannot fix
a `no_mean` estimate.

## Scenarios by reweighting

`cm_scenario()` replaces input distributions and reweights the existing
draws (importance sampling), so no re-run is needed. Here the sampler used a
defensive mixture for the lameness-paratuberculosis odds ratio, so a
scenario with a stronger association still has support:

```{r}
sc <- cm_scenario(mc, list("assoc:LAM:PTB" = dist_lognormal_ci(4, 2.5, 6.4)))
sc$ess
summary(sc, what = "total")
```

Check the effective sample size: a small value means the scenario is poorly
covered by the original draws. Inputs drawn with an outcome correlation
cannot be reweighted individually.

## Sensitivity

One-at-a-time sensitivity (as in Rasmussen et al. 2024, Fig. 7):

```{r}
oat <- sensitivity_oat(example_supplement(), variation = 0.2)
oat
```

Screening associations, including pairs with no estimate, to find those
worth estimating:

```{r}
screen_associations(example_supplement(), or_values = c(0.5, 2))
```

Comparing complete scenarios:

```{r}
base <- example_supplement()
compare_scenarios(base = base, d1_d3_linked = set_association(base, "d1", "d3", 2))$totals
```
