---
title: "Uncertainty and sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uncertainty and sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

Any input can be uncertain: a disease probability, an association, a
three-way term, an impact, an interaction, and, for event impacts, the
overall risk. `deconflate()` draws the uncertain inputs from their
distributions, adjusts each draw in the same way as the central estimate,
and reports intervals. Monte Carlo runs are part of `deconflate()`
(`n_draws`).

## Input distributions

`dist_*()` functions describe uncertain inputs. Draws are on the scale the
inputs were entered on (e.g. percent impacts, odds ratios, incidence rates).

```{r}
dist_pert(1.19, 3.30, 10.71)        # min, mode, max
dist_lognormal_ci(2.7, 1.5, 4.9)    # odds ratio and 95% CI
dist_normal(2.63, 1.44, lower = 0)  # truncated at zero
```

The others are `dist_fixed()`, `dist_lognormal()`, `dist_beta()`,
`dist_pert_mean()`, `dist_uniform()` and `dist_mixture()` (e.g. two pooled
sources of evidence); see `?distributions`.

## Where distributions are given

In input tables, an uncertain value gets its distribution in its own row,
in the columns `dist` and `p1`-`p4`, and `cm_read_inputs()` keeps it in the
model (see `vignette("own-data")`). In R, give them to `cm_model()` as
`distributions`, a named list whose names are keys: `prob:<disease>`,
`assoc:<d1>:<d2>`, `three:<d1>:<d2>:<d3>`, `impact:<disease>` and
`inter:<d1>:<d2>`. Inputs without a distribution keep their point value in
every draw.

```{r}
m <- example_supplement()
mu <- cm_model(m, m$impacts, distributions = list(
  "prob:d1" = dist_beta(20, 180),
  "assoc:d1:d2" = dist_lognormal_ci(2, 1.4, 2.9),
  "impact:d1" = dist_normal(2.5, 0.5),
  "impact:d2" = dist_normal(5, 1),
  "impact:d3" = dist_pert(5, 7.5, 9)
))
mu
```

## Draws and intervals

When any input has a distribution, `deconflate()` runs `n_draws` draws
(1000 by default; here 200 to keep this vignette fast). Each draw is a
complete set of inputs, so the uncertainty of probabilities, associations
and impacts is propagated jointly:

```{r}
r <- deconflate(mu, n_draws = 200, seed = 1)
r
```

The central estimate (`adjusted`) always uses the point values; the draws
give the 95% intervals (`lower`, `upper`: the 2.5% and 97.5% quantiles over
the accepted draws). The intervals are added to the adjusted impacts, the
contributions and the adjusted total:

```{r}
r$contributions
r$totals
```

`$draws` holds the details: `summary` (mean, standard deviation, Monte
Carlo standard error `mcse`, interval, median and a stability check for
every quantity), the accepted draws (`values`, and the drawn inputs in
`params`), the rejections and the seed:

```{r}
r$draws$summary
```

The stability check is `ok`, `imprecise` (the Monte Carlo standard error is
more than 5% of the mean: use more draws, or Latin hypercube sampling),
`heavy_tail` (the most extreme 1% of draws contribute more than 60% of the
variance) or `insufficient_info` (the precision cannot be assessed, e.g.
because the mean is zero; `mcse` still gives the absolute error). A note
lists the quantities that are not `ok`.

Without any distribution, no draws are run, whatever `n_draws`, and a note
says so; `n_draws = 0` skips the draws (and the note) for a model that has
distributions:

```{r}
deconflate(example_supplement())$notes
```

### Seed

Draws are reproducible with `seed`. Without it, a seed is chosen and stored
in the result, so a run can always be repeated:

```{r}
r_again <- deconflate(mu, n_draws = 200, seed = 1)
identical(r$adjusted, r_again$adjusted)
r_unseeded <- deconflate(mu, n_draws = 200)
r_unseeded$draws$seed
```

### Rejected draws

Some draws are impossible: a probability outside (0, 1), a non-positive
odds ratio, or associations that no population can have together. Such
draws are rejected and not replaced, so the intervals describe the
accepted draws only. The number rejected is printed with the result, each
rejection is listed by type with its reason in `$draws$rejections`, and a
note appears when more than 10% are rejected. Here a prevalence is given a
normal distribution that reaches below zero:

```{r}
m_rej <- cm_model(m, m$impacts, distributions = list("prob:d1" = dist_normal(0.05, 0.05)))
r_rej <- deconflate(m_rej, n_draws = 200, seed = 1)
r_rej
head(r_rej$draws$rejections, 3)
```

Draws can also be rejected because the values cannot hold together, even
when each is valid alone. Three diseases with strong odds ratios are
feasible at their point values, but not when one of them is much weaker
(`warn = FALSE` hides a warning that these made-up impacts change sign):

```{r}
strong <- cm_model(
  cm_diseases(c("a", "b", "c"), c(0.5, 0.5, 0.5)),
  cm_impacts(c("a", "b", "c"), c(4, 5, 6)),
  associations = cm_associations(c("a", "a", "b"), c("b", "c", "c"), c(20, 20, 20)),
  distributions = list("assoc:a:b" = dist_lognormal_ci(20, 1, 400))
)
r_strong <- deconflate(strong, n_draws = 200, seed = 1, warn = FALSE)
r_strong$draws$n_rejected
table(r_strong$draws$rejections$type)
r_strong$draws$rejections$reason[1]
```

Report the rejection share: conditioning on acceptance changes the
effective input distribution. A high share means the input distributions
often combine values that contradict each other; narrower or correlated
evidence may be needed.

### Latin hypercube sampling

`sampling = "lhs"` stratifies each input's distribution, which usually
reduces the Monte Carlo error of means. Latin hypercube draws are not
independent, so they are split into `lhs_replicates` independent blocks (10
by default), and the Monte Carlo standard error is estimated from the
spread of the block means. Compare `mcse` with the random draws above:

```{r}
r_lhs <- deconflate(mu, n_draws = 200, seed = 1, sampling = "lhs")
r_lhs
data.frame(quantity = r$draws$summary$quantity,
           mcse_random = r$draws$summary$mcse,
           mcse_lhs = r_lhs$draws$summary$mcse)
```

## Event impacts: the overall risk

For event impacts (`vignette("event-impacts")`), the impacts can have
distributions like any other input, and `overall_risk` can be a
distribution too. The central estimate uses the mean of that distribution:

```{r}
cull <- cm_model(m, cm_impacts(c("d1", "d2", "d3"), c(1.5, 2.0, 1.2),
                               measure = c("HR", "HR", "RR"),
                               estimand = "snapshot_crude", label = "culling"))
e <- deconflate(cull, event_model = TRUE, overall_risk = dist_beta(250, 750),
                n_draws = 200, seed = 1)
e
e$attributable$summary
```

Here only the overall risk is uncertain. The attributable risk and its
allocation to diseases vary with it, while the adjusted hazard ratios
hardly change: raw hazard ratios do not depend on the overall risk, and
only the risk ratio of d3 does.

## Comparing methods on the same draws

`compare_methods()` runs several methods on the same inputs, including the
approximation of Rasmussen et al. (2022, eq. 16) (`"published"`), kept for
comparison with earlier analyses only. With `n_draws` (default 0), every
method is applied to the same draws; a draw is rejected if any method fails
on it. There is no automatic switch of method here: methods that cannot be
run are listed in `$failed`.

```{r}
mu2 <- cm_model(m, m$impacts, distributions = list(
  "assoc:d1:d2" = dist_lognormal_ci(2, 1.4, 2.9),
  "impact:d2" = dist_normal(5, 1)
))
cmp <- compare_methods(mu2, methods = c("published", "simultaneous"), n_draws = 200, seed = 1)
cmp
s <- cmp$draws$summary
s[s$quantity == "adjusted:d2", c("method", "mean", "lower", "upper")]
```

## Gaps and values over the draws

The package reports impacts in their own units. A gap or a value per draw
is computed in base R from the accepted draws in `$draws$values` (one row
per draw, one column per quantity). Here the impacts are percent losses of
an observed level of 10,000 units, and the value of a unit is drawn per
draw:

```{r}
total <- r$draws$values[, "total"]
gap <- 10000 / (1 - total / 100) - 10000
set.seed(5)
value <- gap * runif(length(gap), 0.25, 0.35)
sapply(list(gap = gap, value = value),
       function(x) c(mean = mean(x), quantile(x, c(0.025, 0.975))))
```

The central gap uses the central total:

```{r}
10000 / (1 - r$totals$adjusted_total / 100) - 10000
```

## Sensitivity

The sensitivity tools give point estimates (they do not run draws), and
choose the method as `deconflate()` does. They also accept
`event_model = TRUE` with `overall_risk`. One-at-a-time sensitivity (as in
Rasmussen et al. 2024, Fig. 7):

```{r}
sensitivity_oat(example_supplement(), variation = 0.2)
```

Screening associations, including pairs with no estimate, finds those
worth estimating. Specified associations are multiplied by each of
`multipliers`; pairs without an estimate (here d1 and d3) are set to each of
`or_values`:

```{r}
m_pairs <- cm_model(
  cm_diseases(c("d1", "d2", "d3"), c(0.10, 0.15, 0.20)),
  cm_impacts(c("d1", "d2", "d3"), c(2.5, 5, 7.5)),
  associations = cm_associations(c("d1", "d2"), c("d2", "d3"), c(2, 3))
)
screen_associations(m_pairs, or_values = c(0.5, 2), multipliers = c(0.5, 2))
```

The screens report scenarios that cannot be run, with the reason in column
`failed`, rather than skipping them. Here, weakening any of the three strong
associations above makes the pairs jointly infeasible:

```{r}
screen_associations(strong, multipliers = c(0.05, 2))
```

The screens also work without any association estimates: the baseline then
has every disease independent, and the screen shows which pairs would
matter if they were associated. `screen_interactions()` and
`screen_three_way()` (see `vignette("interactions")`) work in the same way,
and `cm_threshold()` finds the association at which a result changes (see
`vignette("thresholds-and-scaling")`).
