---
title: "Uncertainty, scenarios and sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uncertainty, scenarios and sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

## Input distributions

`dist_*()` functions describe uncertain inputs. Draws are on the scale the
inputs were entered on (e.g. percent impacts, odds ratios, incidence rates).

```{r}
dist_pert(1.19, 3.30, 10.71)        # min, mode, max
dist_lognormal_ci(2.7, 1.5, 4.9)    # odds ratio and 95% CI
dist_normal(2.63, 1.44, lower = 0)  # truncated at zero
```

## Monte Carlo analysis

`cm_sampler()` attaches distributions to a model. Keys name the inputs:
`prob:<disease>`, `assoc:<d1>:<d2>`, `three:<d1>:<d2>:<d3>`,
`impact:<disease>` and `inter:<d1>:<d2>`. (When inputs are read from files
with `cm_read_inputs()`, distributions are given in the rows of the tables,
in the columns `dist` and `p1`-`p4`, and the sampler is built for you; see
`vignette("own-data")`.) Each draw is a complete model, so the uncertainty of
probabilities, associations and impacts is propagated jointly. Here the
association of d2 and d3 is given a defensive mixture (a wider component
with weight 0.2), which is used for a scenario below:

```{r}
m <- example_supplement()
s <- cm_sampler(
  m,
  diseases = list(d1 = dist_beta(20, 180)),
  associations = list(
    "d1:d2" = dist_lognormal_ci(2, 1.4, 2.9),
    "d2:d3" = dist_mixture(dist_lognormal_ci(3, 2, 4.5), dist_lognormal(log(3), 0.8),
                           weights = c(0.8, 0.2))
  ),
  impacts = list(d1 = dist_normal(2.5, 0.5), d2 = dist_normal(5, 1),
                 d3 = dist_pert(5, 7.5, 9))
)
s
mc <- cm_monte_carlo(s, 400, seed = 1)
mc
summary(mc)
summary(mc, what = "total")
```

Draws with impossible inputs (e.g. a probability outside (0, 1), or
associations that no population can have together) are rejected and counted
by type with `summary(mc, what = "rejections")`. A draw whose results are not
finite is rejected as a whole. Report the rejection rate: conditioning on
acceptance changes the effective input distribution.

## Several analyses on shared draws

`cm_batch_sampler()` builds one sampler per analysis of a `cm_analyses()`
object. Each draw of the disease and association inputs is shared by all
analyses, with the same draw identifiers; each analysis draws its own
impacts:

```{r}
pop <- cm_population(m$diseases, m$associations)
a <- cm_analyses(pop,
  yield = cm_impacts(c("d1", "d2", "d3"), c(2.5, 5, 7.5), units = "%"),
  calving_interval = cm_impacts(c("d1", "d2", "d3"), c(4, 10, 2), units = "days")
)
bs <- cm_batch_sampler(a,
  diseases = list(d1 = dist_beta(20, 180)),
  associations = list("d1:d2" = dist_lognormal_ci(2, 1.4, 2.9)),
  impacts = list(yield = list(d1 = dist_normal(2.5, 0.5)),
                 calving_interval = list(d2 = dist_normal(10, 3)))
)
bs
mcb <- cm_monte_carlo(bs, 200, seed = 1)
mcb
summary(mcb, diagnose = FALSE)[, c("analysis", "disease", "mean", "q0.025", "q0.975")]
```

Each analysis's run is in `mcb$analyses`, and can be used with the functions
below.

## Comparing methods on the same draws

Give several methods to `cm_monte_carlo()` and every accepted draw is
adjusted with each of them (a draw is rejected if any method fails on it).
`compare_methods()` then tabulates the results:

```{r}
mc2 <- cm_monte_carlo(s, 400, method = c("published", "simultaneous"), seed = 1)
compare_methods(mc2)
```

## Unstable estimates

Monte Carlo means can be unstable. `summary()` checks every estimate and,
when one looks unstable, prints why and what to do. Here the raw impact of
`d2` is uncertain enough to be near zero or negative:

```{r}
s_d2 <- cm_sampler(example_supplement(), impacts = list(d2 = dist_normal(0.5, 1.5)))
mc_d2 <- cm_monte_carlo(s_d2, 400, method = c("published", "simultaneous"), seed = 2)
sm <- summary(mc_d2)
sm[, c("disease", "method", "mean", "q0.5", "trimmed_mean", "mcse", "stability")]
```

Each estimate gets one of five statuses (see `?cm_diagnose`):

* **`ok`**: no problem detected.
* **`imprecise`**: the Monte Carlo standard error is more than 5% of the
  mean. Use more draws, or Latin hypercube sampling.
* **`insufficient_info`**: the precision cannot be assessed, because fewer
  than two Latin hypercube blocks have draws with positive weight, or
  because the mean is (nearly) zero, so that a relative error is undefined.
  In the second case `mcse` still gives the absolute error, which can be
  judged against the size of effect that matters. A missing precision is
  never reported as `ok`.
* **`heavy_tail`**: the most extreme 1% of draws contribute more than 60% of
  the variance. Importance sampling can help if they come from one region
  of one input (below).
* **`possible_pole`**: the published approximation, `m^2 / (m + c)`,
  divides by `m + c`, which takes both signs within the sampled inputs and
  in some draws has the opposite sign to `m`. The estimate then has a pole
  inside the input distribution, and its mean may not exist; no number of
  draws, and no sampling scheme, makes such a mean converge. Report
  quantiles (e.g. the median), or use the exact method. A trimmed mean is a
  different estimand.
  Removable cases are not flagged: when `c = 0` (e.g. a disease with no
  associated impacts) the formula reduces to `m`, and a sign change of `m`
  alone does not create a pole.

Draws that give non-finite results are rejected, and are noted as
`non_finite`. `cm_diagnose()` lists the flagged estimates with the
suggestions:

```{r}
cm_diagnose(mc_d2)
```

### Latin hypercube sampling

`sampling = "lhs"` stratifies each input's distribution, which usually
reduces the Monte Carlo error of means. LHS draws are not independent, so the
draws are split into `lhs_replicates` independent blocks (10 by default),
and the standard error is estimated from the spread of the block means. The
pooled mean is a ratio estimator over the blocks (each block's weighted sum
over the total weight), and its standard error uses all blocks as sampled
(`mc$n_blocks`), including any whose draws were all rejected; with fewer
than two blocks with positive weight the status is `insufficient_info`:

```{r}
mc_lhs <- cm_monte_carlo(s_d2, 400, method = "simultaneous", sampling = "lhs", seed = 2)
mc_lhs
summary(mc_lhs, diagnose = FALSE)[, c("disease", "mean", "mcse")]
sm[sm$method == "simultaneous", c("disease", "mean", "mcse")]
```

### Importance sampling

When extreme values come from one region of one input,
`cm_suggest_proposal()` builds a defensive mixture that samples that region
more often. The draws are then weighted by the ratio of the densities, so
the estimates still refer to the original input distributions:

```{r}
prop <- cm_suggest_proposal(mc_d2, "d2", method = "simultaneous")
mc_is <- cm_monte_carlo(s_d2, 400, method = "simultaneous", proposal = prop, seed = 3)
mc_is
summary(mc_is, diagnose = FALSE)[, c("disease", "mean", "mcse")]
```

Each proposal must cover the whole support of the input's own
distribution, and point masses (fixed values) cannot be importance-sampled.
Both are checked before any draw is made. A proposal that is too narrow is
rejected:

```{r}
tryCatch(cm_monte_carlo(s_d2, 10, proposal = list("impact:d2" = dist_uniform(-5, 5))),
         deconflate_unsupported = function(e) conditionMessage(e))
```

So is a mixture whose range spans the support but leaves a gap: draws
would never fall in the gap, and the weighted estimates would converge to
the mean over the covered part only. Here the raw impact of d3 has a PERT
distribution on [5, 9], and the proposal has no mass between 6.5 and 7.5:

```{r}
gap <- dist_mixture(dist_uniform(4, 6.5), dist_uniform(7.5, 10))
tryCatch(cm_monte_carlo(s, 10, proposal = list("impact:d3" = gap)),
         deconflate_unsupported = function(e) conditionMessage(e))
```

Including the input's own distribution as a mixture component, as
`cm_suggest_proposal()` does, always covers its support.

A defensive mixture guarantees support, but not a finite variance or better
precision: compare the standard errors. Importance sampling reduces the
error of a mean that exists. It cannot fix a `possible_pole` estimate.

## Scenarios by reweighting

`cm_scenario()` replaces input distributions and reweights the existing
draws (importance sampling), so no re-run is needed. The sampler above used
a defensive mixture for the d2:d3 odds ratio, so a scenario with a stronger
association still has support:

```{r}
sc <- cm_scenario(mc, list("assoc:d2:d3" = dist_lognormal_ci(4, 2.5, 6.4)))
sc$ess
summary(sc, what = "total", diagnose = FALSE)
```

Check the effective sample size: a small value means the scenario is poorly
covered by the original draws. A scenario distribution must lie within the
range the input was sampled from, which is checked. Reweighting cannot
recover rejected draws.

`cm_reweight()` takes any log density ratio as a function of the sampled
inputs (`mc$params`). For example, conditioning on a raw impact of d1 above
2%:

```{r}
cond <- cm_reweight(mc, function(p) ifelse(p[["impact:d1"]] > 2, 0, -Inf))
cond$ess
summary(cond, diagnose = FALSE)[, c("disease", "mean", "q0.5")]
```

## Productivity gaps over the draws

`cm_mc_gap()` applies `productivity_gap()` to every accepted draw. The
observed mean and the unit value can be fixed or drawn:

```{r}
g <- cm_mc_gap(mc, observed = 10000, direction = "decrease", effect = "percent",
               unit_value = dist_uniform(0.25, 0.35), seed = 5)
g$summary
```

## Sensitivity

One-at-a-time sensitivity (as in Rasmussen et al. 2024, Fig. 7):

```{r}
oat <- sensitivity_oat(example_supplement(), variation = 0.2)
oat
```

Screening associations, including pairs with no estimate, to find those
worth estimating. Specified associations are multiplied by each of
`multipliers`; pairs without an estimate (here d1 and d3) are set to each of
`or_values`:

```{r}
m_pairs <- cm_model(
  cm_diseases(c("d1", "d2", "d3"), c(0.10, 0.15, 0.20)),
  cm_impacts(c("d1", "d2", "d3"), c(2.5, 5, 7.5)),
  associations = cm_associations(c("d1", "d2"), c("d2", "d3"), c(2, 3))
)
screen_associations(m_pairs, or_values = c(0.5, 2), multipliers = c(0.5, 2))
```

The screens report scenarios that cannot be run, with the reason in column
`failed`, rather than skipping them. Here, weakening any of three strong
associations makes the pairs jointly infeasible:

```{r}
strong <- cm_model(
  cm_diseases(c("a", "b", "c"), c(0.5, 0.5, 0.5)),
  cm_impacts(c("a", "b", "c"), c(1, 2, 3)),
  associations = cm_associations(c("a", "a", "b"), c("b", "c", "c"), c(20, 20, 20))
)
screen_associations(strong, multipliers = c(0.05, 2))
```

`screen_interactions()` and `screen_three_way()` (see
`vignette("interactions")`) work in the same way. Comparing complete
scenarios:

```{r}
base <- example_supplement()
compare_scenarios(base = base, d1_d3_linked = set_association(base, "d1", "d3", 2))$totals
```
