---
title: "Uncertainty, scenarios and sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uncertainty, scenarios and sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(deconflate)
```

## Input distributions

`dist_*()` functions describe uncertain inputs. Draws are on the scale the
inputs were entered on (e.g. percent impacts, odds ratios, incidence rates).

```{r}
dist_pert(1.19, 3.30, 10.71)        # min, mode, max
dist_lognormal_ci(2.7, 1.5, 4.9)    # odds ratio and 95% CI
dist_normal(2.63, 1.44, lower = 0)  # truncated at zero
```

## Monte Carlo analysis

`cm_sampler()` attaches distributions to a model. Each draw is a complete
model, so disease probabilities and associations are shared by all
outcomes within a draw. `outcome_correlation` additionally correlates a
disease's impacts across outcomes (Gaussian copula):

```{r}
m <- example_uk_dairy_2022()
R <- matrix(c(1, 0.5, 0.5, 1), 2,
            dimnames = list(c("yield", "fertility"), c("yield", "fertility")))
s <- cm_sampler(
  m,
  diseases = list(LAM = dist_beta(30, 70), MAS = dist_beta(30, 70)),
  associations = list("LAM:PTB" = dist_mixture(dist_lognormal_ci(2.7, 1.5, 4.9),
                                               dist_lognormal(log(2.7), 0.8),
                                               weights = c(0.8, 0.2)),
                      "LAM:SCK" = dist_lognormal_ci(2.01, 1.62, 2.49)),
  impacts = list("yield:LAM" = dist_normal(5.54, 0.8),
                 "fertility:LAM" = dist_normal(12.47, 2)),
  outcome_correlation = R
)
s
mc <- cm_monte_carlo(s, 200, method = "simultaneous",
                     economics = uk_dairy_2022_economics(), seed = 1)
mc
summary(mc, what = "total")
```

Draws with impossible inputs (e.g. a probability outside (0, 1), or an
association incompatible with the sampled probabilities) are rejected and
counted. Report the rejection rate: conditioning on feasibility changes the
effective input distribution.

## Scenarios by reweighting

`cm_scenario()` replaces input distributions and reweights the existing
draws (importance sampling), so no re-run is needed. Here the sampler used a
defensive mixture for the lameness-paratuberculosis odds ratio, so a
scenario with a stronger association still has support:

```{r}
sc <- cm_scenario(mc, list("assoc:LAM:PTB" = dist_lognormal_ci(4, 2.5, 6.4)))
sc$ess
summary(sc, what = "total")
```

Check the effective sample size: a small value means the scenario is poorly
covered by the original draws. Inputs drawn with an outcome correlation
cannot be reweighted individually.

## Sensitivity

One-at-a-time sensitivity (as in Rasmussen et al. 2024, Fig. 7):

```{r}
oat <- sensitivity_oat(example_supplement(), variation = 0.2)
oat
```

Screening associations, including pairs with no estimate, to find those
worth estimating:

```{r}
screen_associations(example_supplement(), or_values = c(0.5, 2))
```

Comparing complete scenarios:

```{r}
base <- example_supplement()
compare_scenarios(base = base, d1_d3_linked = set_association(base, "d1", "d3", 2))$totals
```
